### 2018-03-17 EDM_SNI_analysis.Rmd
data prep done and partial on final analysis step-by-step markdown

---
title: "Dynamic Species Interactions in the San Nicolas Island Kelp Forest"
subtitle: "Empirical dynamic modeling analysis"
author: "Owen Liu"
date: "March 13, 2018"
output: html_document
bibliography: citation_bibtex.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE,message=F}
# Load required packages
require(tidyverse)
require(rEDM)
require(quantreg)
require(knitr)
require(extrafont)
```

## Introduction

This document accompanies the manuscript "..." The following analyses lay out the building blocks of the empirical dynamic modeling approach [@Deyle2016; @Chang2017] that produced the final version of the analyses in the paper.

In the analysis that follows, in each step that involves a new type of EDM method, its logic will be briefly described.  However, keep in mind that the basic theory behind each individual method is the same: We use some dynamic information (from single or multiple time series) to reconstruct a shadow attractor, and then use the properties of that manifold to make predictions.

## 1. Variables in the Data

### The San Nicolas Island Dataset

San Nicolas Island is a small, remote island situated about 100 kilometers offshore from southern California.  The island itself is small, about 14 kilometer long and 5 km wide. The data in the analysis is from a sampling station on the western end of San Nicolas Island. The benthic monitoring data herein have been collected more or less every six months for more than 35 years by the USGS and its Western Ecological Research Center (USGS-WERC), and in 2013 the datasets were made available publicly through Ecological Archives:

Michael C. Kenner, James A. Estes, M. Tim Tinker, James L. Bodkin, Robert K. Cowen, Christopher Harrold, Brian B. Hatfield, Mark Novak, Andrew Rassweiler, and Daniel C. Reed. 2013. A multi-decade time series of kelp forest community structure at San Nicolas Island, California (USA). Ecology 94:2654. http://dx.doi.org/10.1890/13-0561.1

And in 2016, the data were again compiled and an updated version was published online:

Miller R., A. Rassweiler, D. Reed, K. Lafferty, L. Kui, M. O'Brien. 2016. Santa Barbara Channel Marine BON: Integrated quad and swath cover. Environmental Data Initiative.

### Physical Oceanographic Data

A body of other research has established that a combination of physical forcing (waves, storms), temperature, and lower frequency climate modes (e.g., El Ninos) have an important influence on the dynamics of kelp forests [@Reed2011; @Cavanaugh2011; @Bell2015,@Young2015]. With these data, we can draw connections between the physical variables and the species interactions in our constrained trophic web.

We have four datasets, already processed into the same time frame (periods) as the SNI benthic monitoring data:

* **[The Multivariate ENSO index (MEI)](http://www.esrl.noaa.gov/psd/enso/mei)**
    + The first principal component of a composite set of physical parameters
    + Positive values of the MEI index are generally associated with El Nino conditions, decreases in wind-driven upwelling, warmer surface waters and nutrient-poor conditions
    + Variable here is the average index value for the four months preceding each Spring or Fall monitoring period (i.e., December to March or June to September, respectively)

* **[The Pacific Decadal Oscillation index (PDO)](http://research.jisao.washington.edu/pdo/)**
    + Leading empirical orthogonal function (EOF) of monthly sea surface temperature anomalies (SST-A) over the North Pacific (poleward of 20° N) after the global average sea surface temperature has been removed
    + Positive PDO values indicate warmer SST, and nutrient-poor conditions along the western coast of the contiguous United States
    + Aggregated and averaged the same way as MEI

* **[The North Pacific Gyre Oscillation (NPGO)](http://www.o3d.org/npgo/)**
    + From [@DiLorenzo2008]
    + Climate pattern that emerges as the 2nd dominant mode of sea surface height variability (2nd EOF SSH) in the Northeast Pacific
    + Better correlated with salinity, nutrients, and chlorophyll than PDO, showing forcing for the planktonic community
    + Strong predictor of upwelling cells south of 38 deg N
    + Aggregated and averaged the same way as MEI and PDO

* **Sea surface temperature (SST)**
    + Two sources (to fill in data gaps):
      + Sea surface temperature data directly from Begg Rock and San Nicolas Island buoys, from the [Coastal Data Information Program (CDIP)](http://cdip.ucsd.edu)
      + NOAA's [Optimally Interpolated Sea Surface Temperature](https://www.ncdc.noaa.gov/oisst)
    + Similar to the above, value is an average SST for the four months preceding each period
* **Maximum significant wave height (Hs)**
    + Combined data from the Begg and SNI buoys with modeled data from the [Geophysical Fluid Dynamics Laboratory](http://cmgwindwave.usgsportals.net/)
    + Signficant wave height is defined as the average height, in meters, of the one third highest waves in the record
    + Instead of an average, value here is the maximum significant wave height of the four months preceding each period. This is meant to capture any large storm events, as well as general level of physical disturbance

Unlike the biological data, where there are unique spatial replicates, the physical data have only one value for each of the 69 monitoring periods, and hence their values are replicated (copied) for each site to match the total length of the biological data.

#### Note about data processing

Usual best practice for EDM is to create normalized time series, so that state-space reconstructions are not distorted by differences in orders of magnitude between variables. In addition, in this study we concatenate each species/station/swath time series into one long time series for each species, while preserving a time indicator (variable 'period') and a site identifier, to ensure that we do not "cross" the boundaries of replicate time series in analyses. Empirical dynamic modeling can use multiple spatial replicates in lieu of increased length of individual time series, to maximize the dynamic information that can be drawn from the system [@Hsieh2008;@Clark2015;@Chang2017].

All raw data is available online or in the included files. We join the monitoring and physical datasets and normalized all time series (see scripts in the "data" folder for details:

```{r load data,message=FALSE,echo=F}
source("data/data preparation scripts/produce_final_data.R")
```

```{r list of variables,echo=F,fig.cap="List of variables"}
tibble(Variable=colnames(westend),Description=c("Monitoring Transect","Monitoring Period","Laminaria Normalized Density",
                                            "Macrocystis pyrifera >1m Normalized Density","Pterygophora Normalized Density",
                                            "Strongylocentrotus purpuratus Normalized Density","Mesocentrotus franciscanus Normalized Density",
                                            "Macrocystis pyrifera <1m Normalized Density","Normalized Multivariate ENSO Index",
                                            "Normalized Pacific Decadal Oscillation","Normalized North Pacific Gyre Oscillation",
                                            "Normalized Maximum Significant Wave Height","Normalized Sea Surface Temperature")) %>%
  knitr::kable()
```

```{r mean density over time, echo=F,message=F}
# Species naming key for use in plotting
namekey <- tibble(dataset=c(rep("Benthic density",6),rep("Physical",5)),short=names(westend[3:13]),
                  long=c("Laminaria","Macrocystis >1m","Pterygophora","Purple urchin","Red urchin","Macrocystis <1m","Multivariate ENSO Index","Pacific Decadal Oscillation","North Pacific Gyre Oscillation","Significant Wave Height","Sea Surface Temperature"),plotting=c("L. far","M. pyr","P. cal","S. pur","M. fra","M. pyr (j)","(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST"))

namekey <- bind_rows(namekey, data_frame(long="NA",dataset="NA",short="const",plotting="Constant")) %>%

  # define column "plotting" as a factor so we can later control plotting order on figures
  mutate(plotting=factor(plotting,levels=c("(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST","L. far","M. pyr","M. pyr (j)","P. cal","S. pur","M. fra","Constant")))

# key relating monitoring period to its year/month
period.key <- data_frame(year=rep(1980:2014,each=12),month=rep(1:12,35),period=c(rep(NA,8),rep(1,4),rep(2:69,each=6)))
periods.first <- period.key %>% 
  distinct(period,.keep_all=T) %>%
  mutate(day=1) %>% #assume all monitoring done on first day of month
  unite(date_str,year,month,day,sep="-") %>%
  mutate(date_str=ymd(date_str))

westend.long <- westend %>%
  gather(key=spp,value=dens,-period,-site,na.rm=T) %>%
  left_join(namekey, by=c("spp"="short"))
westend.meandens <- westend.long %>%
  group_by(spp,period,long) %>% 
  summarise(meandens=mean(dens,na.rm=T),sddens=sd(dens,na.rm=T))%>%
  mutate(lower=meandens-sddens,upper=meandens+sddens)%>%
  left_join(periods.first,by="period") %>%
  ungroup()

# position adjustment for visual clarity
pd <- position_dodge(width=100)
westend.mean_dens_plot <- westend.meandens %>%
  filter(spp %in% c("mac","red","pter","ymac","purp","lam")) %>%
  mutate(type=ifelse(spp %in% c("mac","ymac","pter","lam"),"algae","urchin")) %>%
  ggplot(aes(x=date_str,y=meandens,col=long,linetype=type))+
      geom_hline(yintercept=0)+
      geom_line(lwd=2,alpha=0.8,position=pd)+
      geom_pointrange(aes(ymin=lower,ymax=upper),size=0.3,position=pd)+
      xlab("Year")+ylab("Normalized Density")+
      scale_color_manual(values=c("navyblue","gray50","darkgreen","darkcyan","mediumpurple4","darkred"),
                         guide=guide_legend(title="Species",override.aes = list(size=1)))+
      guides(linetype="none")+
      theme_minimal()+
      theme(text = element_text(family="Rockwell",size=16),
            legend.position = c(0.15,0.8),
            legend.text = element_text(size=12))
westend.mean_dens_plot
rm(pd,period.key,periods.first,westend.long,westend.meandens)
```

***

## 2. Establishing Univariate Predictability and Nonlinearity

For each of these time series, we have a few steps to see if they seem appropriate to analyze together with EDM techniques [@Chang2017]. We want to be careful that our state space reconstructions are reliable and represent valid manifolds. In other words, we don't want to rely on simple cross-correlation or the prior knowledge that all these data were collected around the same locations at around the same times. We want evidence that:

1. Variables can be probably embedded (i.e., they show evidence of limited system dimensionality)
2. Variables display state-dependent (nonlinear) dynamics, and therefore that nonlinear (EDM) methods are appropriate for analysis of these data

First, we use simplex projection to get a sense of system dimensionality for each variable, which will also give us an idea of the appropriate embedding dimension to use in later analyses. Then, we explicitly look for evidence of state-dependence and nonlinear dynamics with a prediction horizon test and S-maps.

***
```{r time series segments, include=F}
# segments of the time series (which rows of the data are individual transects?)
westend_segs_sites <- westend %>%
  ungroup() %>%
  mutate(ind = row_number()) %>% 
  group_by(site) %>%
  summarise(first=first(ind),last=last(ind))

# Time series segments without site identifier. We'll need this later
westend.segs <- select(westend_segs_sites,-site)

# species in the study
study_spp <- c("red","purp","lam","pter","mac","ymac")
# physical variables in the study
phys.vars <- c("mei","pdo","npgo","waves","sst")
```

***

### Simplex Projection and Embedding Dimensions

For each species/variable separately, we will search for signals of deterministic behavior using simplex projection. In simplex projection, we first reconstruct a shadow attractor in *E* dimensions, where *E* is the number of variables, or number of progressive lags of a single variable used in the reconstruction. *E* is called the __*embedding dimension*__. The *E*-length vectors, for example $\it{\bf{x_{t}}} = <x_t,x_{t-1},x_{t-2}>$ are points on the attractor, and the set of *E*-length vectors used for the reconstruction is called the __*library*__. To predict $\it{\bf{x_{t+1}}}$, the simplex algorithm finds the *E* +1 nearest neighbors of $\it{\bf{x_{t}}}$ in the state space, and the prediction $\it{\bf{\hat{x}_{t+1}}}$ is the average of the nearest neighbors' values at $\it{t+1}$, weighted by their Euclidean distance from $\it{\bf{x_{t}}}$ at $\it{t}$. This is the essence of simplex projection: a forecast for a given point in state space is surmised from the forward trajectories of observed nearby points.

In descriptive terms, this is akin to asking, "When the system has been in a state like this before, what happened next?" For example, if we are interested in predicting *Macrocystis* density next year, we might take the current three-year density trend (this year, last year, and the year before, *E*=3) and compare it to a subset of times in the past when successive three-year dynamics looked similar to that set of points. Our prediction, logically, would be the average of that subset, projected foreward one year and weighted by their similarity to the current trend.

By varying the value of *E*, we can determine what the best embedding dimension is for each variable in our analysis, essentially a proxy for the number of variables that best "unfolds" or best represents the shadow attractor. We can measure the skill of an embedding by comparing the estimated forecasts $\it{\bf{\hat{x}_{t+1}}}$ with the observed values $\it{\bf{x_{t+1}}}$, and we report it with $\it{\bf{\rho}}$, the Pearson correlation coefficient between predictions and observations. To avoid in-sample fitting, we use a leave-one-out cross-validation scheme, removing one vector at a time from the library, and predicting its dynamics from the other library vectors.

```{r univariate simplex westend,fig.height=10,fig.width=7,echo=F}
# List to store output of simplex projection
westend.simp.list <- list()

# Run simplex projection for each species at each site, and plot output
par(mfrow=c(3,2))
for(i in 1:length(study_spp)) {
  spp <- study_spp[i]
  dat <- westend %>% select(matches(spp)) %>% as.data.frame()
  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=2:15,silent=T)
  westend.simp.list[[spp]] <- out

  # Plot embedding dimension vs. predictive ability (rho)
  plot(out$E,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],
       ylim=c(0,1),
       xlab="Embedding Dimension (E)",ylab=expression(paste("Skill, ",rho)))
}

# Save best embedding dimensions
westend.bestE <- sapply(westend.simp.list,function(x) {
  temp <- x %>% filter(!is.na(rho))
  temp$E[temp$rho==max(temp$rho)]
})

# on looking at the graph, manually change embedding dimension for macrocystis and pterygophora to one that has a very similar
# predictability, but will let us make use of more data

westend.bestE["mac"] <- 6
westend.bestE["ymac"] <- 8
westend.bestE["pter"] <- 6

rm(spp,dat,out)

# Run simplex for each physical variable
par(mfrow=c(3,2))
phys.simp.list <- list()
for(i in 1:length(phys.vars)) {
  ind <- phys.vars[i]
  dat <- westend %>% select(matches(ind)) %>% as.data.frame()
  out <- simplex(as.numeric(dat[,1]),lib=c(1,69),E=2:10,silent=T)
  phys.simp.list[[ind]] <- out
  westend.simp.list[[ind]] <- out

  # Plot embedding dimension vs. predictive ability (rho)
  plot(out$E,out$rho,type="l",main=namekey$long[match(ind,namekey$short)],xlab="Embedding Dimension (E)",ylab=expression(paste("Skill, ",rho)))
}
rm(ind,dat,out)
phys.bestE <- sapply(phys.simp.list,function(x) {
  temp <- x %>% filter(!is.na(rho))
  temp$E[temp$rho==max(temp$rho)]
})

westend.bestE <- c(westend.bestE,phys.bestE)
```

***

### Prediction Horizon test

We also want to look at prediction decay for each variable, which is one piece of evidence that a dynamic system is nonlinear. Using the best *E* identified in the previous step, we attempt to make predictions increasing far into the future, instead of just one period ahead. A nonlinear system should show decreasing predictive power with increasing prediction horizon[@Sugihara1994]. This phenomenon is a property of deterministic chaos and is analagous to the "butterfly effect", where in a nonlinear system, trajectories in state-space are expected to diverge over time. To examine this effect, we hold *E* constant, and proceed with simplex projection as before, but varying the prediction horizon, $t_p$ (i.e., how many steps ahead we try to predict).

```{r prediction horizon test,message=F,echo=F,fig.height=10,fig.width=7,echo=F}
# List to hold prediction horizon results
westend.tp.list <- list()

par(mfrow=c(3,2))
for(i in 1:length(study_spp)) {
  spp <- study_spp[i]
  dat <- westend %>% select(matches(spp)) %>% as.data.frame()
  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),silent=T,E=westend.bestE[spp],tp=1:10)
  westend.tp.list[[spp]] <- out

  # Plot time horizon vs. predictive ability (rho)
  plot(out$tp,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],xlab="Time to Prediction",ylab=expression(paste("Skill, ",rho)))
}
rm(spp,dat,out)
```

The prediction horizon effect check is promising overall, as most variables decline in predictive ability with increasing time horizon. For *Pterygophora californica*, there is some evidence of cyclic behavior, in that dynamics are more predictable 6 periods in the future than 3 or 4 periods. This is something to be mindful of, but we leave *Pterygophora* in our analysis for now.

***

### S-maps for Each Species

All of the variables show a decent ability to self-predict, as shown by simplex projection, although some variables require rather high embedding dimensions, indicative of higher-dimensionality dynamics. However, the butterfly effect check is evidence of potential nonlinearity, as all variables decline in predictive ability with time horizon.

We can look for further evidence of nonlinearity with S-maps. S-maps is short for "sequentially weighted global linear maps", and it is similar to simplex projection, except instead of using just the *E* +1 nearest neighbors to make forecasts, S-maps uses all library vectors, and exponentially weights them by their distance to the prediction vector before using linear regression to make a forecast. A parameter, $\it{\bf{\theta}}$, tunes how much greater weight is given to nearby points. If $\it{\bf{\theta}}=0$, all library vectors are weighted equally, and the resulting model is just a vector autogressive (VAR) model of order *E*. However, as $\it{\bf{\theta}}$ is tuned above 0, nearby points in state-space are given more weight in forecasts. Therefore, if model skill $\it{\bf{\rho}}$ increases with increasing $\it{\bf{\theta}}$, it is evidence of nonlinear, state-dependent dynamics. For a more formal description of the S-maps procedure, see @Sugihara1994 and @Deyle2016.

As a side note, with $\it{\bf{\theta}}>0$, although the set library vectors remains constant, the *weights* given to library vectors for regression is specific to each point in state-space, and therefore a separate linear map is created for each predicted vector. This is why the procedure is called "sequentially weighted global linear maps". Conceptually, as the dynamic system moves along the surface of the attractor, S-maps sequentially computes linear maps to the next point based on nearby points.

For prediction of each variable using S-maps, we can again use the optimal embedding dimension, *E*, found through our simplex projection above, and we plot the tuning parameter $\it{\bf{\theta}}$ against $\it{\bf{\rho}}$ to look for evidence of nonlinear dynamics.

```{r univariate s_maps westend,fig.height=10,fig.width=7,echo=F}
westend.smap.list <- list() # list to store output

par(mfrow=c(3,2))
for(i in 1:length(study_spp)) {
  spp <- study_spp[i]
  dat <- westend %>% select(matches(spp)) %>% as.data.frame()
  out <- s_map(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=westend.bestE[spp],silent=T)
  westend.smap.list[[spp]] <- out

  # Plot theta dimension vs. predictive ability (rho)
  plot(out$theta,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],xlab=expression(paste("Nonlinearity (",theta,")")), ylab=expression(paste("Skill, ",rho)))
}
rm(spp,dat,out)
```

All variables show significantly improved predictive ability with increased theta, suggesting nonlinear dynamics. Together, the simplex, prediction horizon, and S-map results suggest our approach is valid--variables are predictable, and most predictable in a nonlinear manner.

## 2. Convergent Cross Mapping

Generalizations of Takens' theorem indicate that if two variables (in our case, species or physical variables) are part of the same dynamic system, their individual dynamics should reflect their relative causal influence [@Sugihara2012; @Deyle2013,@Ye2015,@Clark2015]. In other words, if one variable (for example, giant kelp), is casually forced by another (sea urchins), that forcing should leave a signature on the giant kelp time series. Convergent cross mapping (CCM) tests for causation by using the attractor/manifold built from the time series of one variable to predict another. CCM works just like univariate simplex projection that we did in Step 1, except that separate variables are used for library and prediction vectors. In addition, we normally predict contemporaneous values of the other variable, instead of projecting one step forward (prediction horizon $t_{p}=0$). If the attractor can accurately (based on out-of-sample prediction skill, just as before) predict the dynamics of the second variable, we can claim that the second variable has a causal influence on the first.  In other words, the *causal effect of A on B is determined by how well B cross-maps A*. In this way, the inference from cross-mapping is the converse direction of causation. In our example, if sea urchins drive giant kelp,the dynamic information from the urchin time series should be reflected in the kelp dynamics, and kelp should significantly cross-map the urchins.

Cross-mapping can distinguish unidirectional forcing (A forces B but B does not force A) from bi-directional (A and B force each other). It can also resolve transitive causal chains (A causes B causes C, see Fig. 4 in @Sugihara2012). To look for a causal signal, we plot predictive skill  $\it{\bf{\rho}}$ against library size (the number of embedded vectors used to construct the attractor). There are two criteria for CCM to establish causality: 
* First, and most obviously, predictive cross-map skill using all available data should be significantly greater than zero. 
* Second, that predictability should be convergent.  Convergence means that cross-mapped estimates improve with library length, because the attractor is more fully resolved and therefore estimation error should decline. Convergence is key to distinguishing causation from simple correlation.

The CCM algorithm uses a random sampling method to test multiple "versions" of each library size, sampling a subset from the supplied library vectors to give a sense of the confidence intervals around prediction skill. We again use leave-one-out cross-validation to prevent in-sample fitting.

We can use CCM theory to ensure that the species in our data are actually displaying causal signals.

***

```{r ccm all species and physical vars, echo=F,warning=F}
# Rows hold predicted variables, columns hold forcing variables. There are more columns than rows since the physical variables are included as potential forcing factors.
n_col <- dim(westend)[2]-2
n_row <- dim(westend)[2]-7
col_names <- colnames(westend)[3:(n_col+2)]
row_names <- colnames(westend)[3:(n_row+2)]
westend.xmap_mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))

## CCM causation criterion 1: cross-map skill greater than zero
# matrix to store a bootstrapped p-value, measuring the probability that a given xmap is greater than zero (calculated as 1 minus the number of positive results for rho divided by the number of iterations)
westend.p1.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))

## CCM causation criterion 2: evidence for convergence
# similarly, matrix to store a bootstrapped p-value, this time a t-test value between library size 10 and library size 500, to see if the rho at large library is significantly greater than the rho at small library size (i.e., looking for convergence).
westend.p2.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))

# if both p1 and p2 are positive, indicate overall significant causal signal
westend.ptot.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))

## Run CCM for each combination of variables
for(i in 1:n_row) {
  for(j in 1:n_col) {
    if(i != j) {
      # remember, we use the best embedding dimension for the target variable (the variable we're cross-mapping to, i.e. the putative forcing variable)
      tempE=westend.bestE[col_names[j]]
      temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 2+j,lib_sizes = c(10,500),num_samples=100,replace=T,silent=T)
      # mean rho at library size 500
      rhomean <- temp %>% filter(lib_size==500) %>% ccm_means()
      westend.xmap_mat[i,j] <- rhomean$rho

      # first p-value (greater than zero? at library size 500)
      p1 <- temp %>% filter(lib_size==500) %>% 
        mutate(pos=ifelse(rho>0,1,0)) %>%
        summarise(p=(1-sum(pos)/n()))
      westend.p1.mat[i,j] <- as.numeric(p1)

      # second p-value (rho at lib-size 500 greater than rho at lib-size 10?)
      p2 <- t.test(temp$rho[temp$lib_size==10],temp$rho[temp$lib_size==500])$p.value
      westend.p2.mat[i,j] <- as.numeric(p2)

      # overall significance (both p1 and p2 signficant at alpha 0.05)
      westend.ptot.mat[i,j] <- ifelse(p1<0.05 & p2<0.05,1,0)
    }
  }
}
rm(n_col,n_row,col_names,row_names,p1,p2,tempE,temp,rhomean)

# keep only signficant cross-mappings
westend.xmap_mat <- westend.xmap_mat*westend.ptot.mat %>% as.data.frame()
westend.xmap_rast <- westend.xmap_mat %>% 
  mutate(predictee=row.names(westend.xmap_mat)) %>% 
  gather(key=predictor,value=rho,-predictee)

# If rho is zero, replace with NA (no significant causal signal)
westend.xmap_rast$rho[westend.xmap_rast$rho==0] <- NA

# Names for plotting to distinguish biological and physical variables
westend.xmap_rast <- westend.xmap_rast %>% 
  mutate(force.name=namekey$plotting[match(predictor,namekey$short)]) %>%
  mutate(pred.name=namekey$plotting[match(predictee,namekey$short)])

## plot
westend.xmap_all_plot <- ggplot(westend.xmap_rast,aes(x=force.name,y=pred.name,fill=rho)) +
  geom_raster() +
  scale_fill_gradient(low = "#9AFF9A", high = "#548B54", space = "Lab", na.value = "grey50", 
      guide = "colourbar",name=expression(paste(rho, "(skill)"))) +
  ggtitle("Kelp Forest Convergent Cross Mapping, West End") +
  xlab("Predictor (Forcing Variable)") +
  ylab("Predicted Variable") +
  theme(axis.text.x=element_text(angle = 90, hjust = 1,vjust=0.1))

westend.xmap_all_plot
```

The plot shows cross-mapped (predictor, or forcing) variables in the columns, while rows are predicted variables. Gray indicates a lack of significant causal signal, and darker colors represent higher cross-map skill.

From this plot, we can make a few observations. The five physical forcing variables are represented in the first five columns, followed by the algae and urchin species. Many physical variables show causal relationships with kelp forest species, especially the North Pacific Gyre Oscillation and Pacific Decadal Oscillation, which show causal links to almost all the biological variables. The NPGO is cross-mapped well by the purple urchin *Strongylocentrotus purpuratus*, while SST in turn shows a strong influence on adult *Macrocystis* dynamics. So already we are seeing effects that we might expect based on known interactions between physical variables and biological dynamics [@Reed2011;@Bell2015].

Additionally, adult *Macrocystis* itself seems to be a strong driver of the dynamics of many, if not all the other species in the tropic web, supporting the decades of research showing the importance of giant kelp as a foundational species [@Graham2007c;@Dayton1985], but further suggesting that its dynamics fundamentally drive the dynamics of other species.  Juvenile, or sporophyte *Macrocystis*, on the other hand, is affected by most of the variables in the system, but does not itself drive dynamics of other species, again an effect we would expect to find. In this system, *Pterygophora californica* also seems to be a signicant driver of the dynamics of other species in the trophic web.

Both sea urchin species are causal predictors of *Macrocystis* density, and they also have a causal effect on one another. This analysis alone, however, does not yet tell us whether the bi-directional interaction between *S. purpuratus* and *M. franciscanus* represents apparent mutualism, apparent competition, or a mixture of both at different times. To further investigate the direction and magnitude of the species interactions themselves, we next build multivariate EDM models that explicitly measure dynamic species interaction strengths.

## 3. Multivariate Models Using CCM results

For each site and each species, we'll only use the variables that show a causal link to build multivariate models, also removing those variables that did not pass diagnositic tests (simplex or prediction horizon). We find the causal variables for each species, based on the CCM output for that species.
```{r causal vars fxn,message=F,echo=F}
causal_vars <- function(pmat,vars) {
  out <- list()
  for(i in 1:length(vars)) {
    v <- vars[i]
    causes <- colnames(pmat)[pmat[v,]==1]
    causes <- causes[!is.na(causes)]
    out[[v]] <- c(causes,v)
  }
  return(out)
}
westend.causal.vars <- causal_vars(westend.ptot.mat,vars=c("mac","purp","lam","ymac","red","pter"))
```

## References

### 2018-03-17 data/data preparation scripts/SST_ncdf4_extract.R

@@ -1,0 +1,80 @@
+### Extract appropriate SNI SST data
+# Much code from Luke Miller, but modified to be much simpler and smaller in scope
+# http://lukemiller.org/index.php/2014/11/extracting-noaa-sea-surface-temperatures-with-ncdf4/
+
+
+# Extracts the data from the .nc file, and casts to a long form dataframe
+
+require(ncdf4)
+require(fields)
+require(lubridate)
+require(dplyr)
+
+WD <- getwd()
+
+extr_ncdf_SNI <- function(fname) {
+  nc = nc_open(fname)
+  ncdates = nc$dim$time$vals
+  ncdates = as.Date(ncdates,origin = '1800-1-1')
+  nlon = length(nc$dim$lon$vals)
+  lons= nc$dim$lon$vals
+  nlat = length(nc$dim$lat$vals)
+  lats=nc$dim$lat$vals
+  ndates = length(ncdates)
+  
+  sstout = array(data = NA, dim = c(nlon,nlat,ndates))
+  # Extract the data from the NetCDF file
+  sstout[,,] = ncvar_get(nc, varid = 'sst',
+                         count = c(nlon,nlat,ndates))
+  # The output array sstout will be arranged with longitudes in rows, 
+  # increasing in an easterly direction as you move down a row (larger 
+  # longitude values), and latitudes in columns, increasing in latitude (more 
+  # northerly) as you move across columns. The 3rd dimension represents 
+  # different dates.
+  
+  # If there are missing data in the NetCDF, they should appear as 32767. 
+  # Replace that value with NA if it occurs anywhere.
+  sstout = ifelse(sstout == 32767, NA, sstout)
+  
+  # Rearrange the output matrix or array so that latitudes run from north to
+  # south down the rows, and longitudes run from west to east across columns.
+  dims = dim(sstout) # get size of array
+  # Make new output array to hold rearranged data. The dimension names will
+  # match the newly rearranged latitude and longitude values
+  sst2 = array(data = NA, dim = c(dims[2],dims[1],dims[3]),
+               dimnames = list(Lat = rev(lats),
+                               Long = lons,
+                               Date = as.character(ncdates)))
+  # Step through each page of array and rearrange lat/lon values
+  for (i in 1:dims[3]){
[…]

### 2018-03-17 data/data preparation scripts/benthic_data_prep.R

@@ -1,0 +1,61 @@
+# Benthic swath data preparation
+# Data of benthic densities of key species of interest for EDM
+# Michael C. Kenner, James A. Estes, M. Tim Tinker, James L. Bodkin, Robert K. Cowen, 
+# Christopher Harrold, Brian B. Hatfield, Mark Novak, Andrew Rassweiler, and Daniel C. Reed. 2013. 
+# A multi-decade time series of kelp forest community structure at San Nicolas Island, California (USA). 
+# Ecology 94:2654. http://dx.doi.org/10.1890/13-0561.1
+
+# This function takes the raw benthic density and benthic cover data, edits the species names, 
+# and fills in missing values to create complete timeseries.
+
+library(tidyverse)
+
+#### Species Key ####
+# Kenner et al. published a species key that we use to match species code to species name
+# All species, all surveys filtered to just benthic density survey
+spp.key <- read_csv('data/raw/Table4_Species_sampled.csv',col_types = cols()) %>% 
+  filter(DataSet=="Benthic density")
+
+# better shorthand names
+spp.key$name <- c("pat","red","purp","wavy","astro","derm","halc","halr","limp","paras","pis","pyc","cys","lam","ymac","youn","eis","pter","mac")
+
+
+#species of interest
+study_spp <- c("mac","ymac","lam","pter","red","purp")
+
+#study sites of interest (permanent transects at the "West End" of San Nicolas Island)
+westend.sites <- c("2_10L", '2_22L', "2_32R", "2_39L", "2_45L", "3_10R" ,"3_22R", "3_32L", "3_39R", "3_45L")
+
+##### Load, filter, and normalized the data ####
+westend.norm <- read_csv('data/raw/Benthic density raw data.csv',col_types="iiccid") %>%
+  
+  #fix date column
+  mutate(Date=as.POSIXct(Date,format="%m/%d/%y")) %>%
+
+# Replace species code with species name and rename some variables
+  rename(station=Station,period=Period,date=Date,swath=Swath,dens=Density) %>%
+  left_join(select(spp.key,SpeciesCode,name),by="SpeciesCode") %>%
+  
+  # remove unneeded columns
+  select(-SpeciesCode,-date) %>%
+  rename(spp=name) %>%
+  arrange(spp,period,swath) %>%
+  
+  # join station and swath columns into one unique 'site' identifier
+  unite(site,station,swath) %>% 
+  
[…]

### 2018-03-17 data/data preparation scripts/benthic_swath_data_preparation.R

@@ -1,0 +1,93 @@
+# Benthic swath data preparation
+# Data of benthic densities of key species of interest for EDM
+
+# Load required libraries
+library(tidyverse)
+library(lubridate)
+
+# read in raw integrated quad swath data directly from source; https://environmentaldatainitiative.org/
+dat <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=edi.6.1&entityid=6c16f3b59574b044dd3a44cef29b9b9b",
+                col_types = "cccccDciidiccccdd")
+
+# filter out just the sampling site of interest at San Nicolas Island
+sites <- dat %>%
+  filter(data_source=="sni")%>%
+  distinct(site_id,subsite_id,.keep_all = T)%>%
+  select(data_source,site_id,subsite_id,site_name,longitude,latitude)
+westend <- dat %>%
+  filter(data_source=="sni",site_name==12)
+
+# filter for species of interest
+spp <- c("Mesocentrotus franciscanus","Strongylocentrotus purpuratus","Macrocystis pyrifera","Laminaria","Pterygophora californica")
+westend <- westend %>%
+  filter(taxon_name %in% spp) %>%
+  # short-hand names for species for later reference
+  mutate(spp = case_when(
+    taxon_name=="Mesocentrotus franciscanus" ~ "red",
+    taxon_name=="Strongylocentrotus purpuratus" ~ "purp",
+    # CHECK THAT MAC AND YOUNG MAC ARE CORRECT (ASK ROBERT MILLER)
+    taxon_name=="Macrocystis pyrifera" & proj_taxon_id=="t-s-068" ~ "mac",
+    taxon_name=="Macrocystis pyrifera" & proj_taxon_id=="t-s-080" ~ "ymac",
+    taxon_name=="Laminaria" ~ "lam",
+    taxon_name=="Pterygophora californica" ~ "pter"
+  ))%>%
+  arrange(taxon_name,date) %>%
+  # calculate density
+  mutate(dens=count/area) %>%
+  # remove unneeded variables
+  select(date,proj_taxon_id,auth_taxon_id,taxon_name,spp,transect_id,replicate_id,area,count,dens)
+
+# we need to apply distinct time periods to the data, because surveys happen in Fall and Spring, but each survey occurs 
+# over the course of a few days.
+# for example, for our analysis, the date October 19, 1987 should be in the same "monitoring period" as October 20 or October 21, 1987.
+# if observations are within 50 days of each other, we consider them the same monitoring period
+
[…]

### 2018-03-17 data/data preparation scripts/physical_data_preparation.R

@@ -1,0 +1,239 @@
+# Physical/Oceanographic Data Preparation for San Nicolas Island EDM analysis
+
+#### Introduction ####
+# The following code takes raw physical datasets and matches them to the same timescale as the SNI community benthic monitoring datasets.  
+# There are four physical variables to be matched:
+#   
+# * The Multivariate ENSO index
+# * The Pacific Decadal Oscillation
+# * The North Pacific Gyre Oscillation
+# * Sea Surface Temperature and Significant Wave Height
+# 
+# For each, we:
+# 1. Match them to the monitoring "periods" from the SNI benthic monitoring datasets. For the purpose of using physical variables
+#    to predict biological variables, we offset the periods such that physical variables applied to each period are actually
+#    the mean value for the four months preceding and during the monitoring period. For example, the first monitoring period
+#    in the SNI dataset is Fall, 1980. The physical variables that would apply to this period are the data from summer, i.e. Jun-Sep 1980.
+#    Similarly, for the Spring monitoring periods, the physical variables are averaged from the previous Dec-March.
+# 2. Normalize the time series to zero mean and unit variance, just as we did with biological variables
+
+# Finally, we join the datasets together, aligning all relevant physical data by monitoring period
+
+# Required libraries
+library(tidyverse)
+library(lubridate)
+
+#### Period matching key ####
+# Have to line up the periods such that December is included in the next Spring's period
+period.key <- data_frame(year=rep(1980:2014,each=12),month=rep(1:12,35),period=c(rep(NA,5),rep(1,6),rep(2:69,each=6),NA))
+
+# Set months that are not in Dec-Mar or Jun-Sep as NA so their data aren't included
+period.key$period[!period.key$month %in% c(12,1,2,3,6,7,8,9)] <- NA
+
+#### Multivariate ENSO Index ####
+# source: http://www.esrl.noaa.gov/psd/enso/mei
+
+mei <- read_csv("data/raw/mei.csv",col_types=cols())
+# These data are for every year since 1950, and bimonthly MEI index values. 
[…]

### 2018-03-17 data/data preparation scripts/produce_final_data.R

@@ -1,0 +1,20 @@
+### Join benthic swath and physical variables ###
+
+## This script produces the final dataset ready for EDM analysis
+
+## Normalized time series from benthic swath data and physical variables
+
+library(tidyverse)
+
+## Run the scripts that processed normalized time series
+source("data/data preparation scripts/benthic_data_prep.R")
+source("data/data preparation scripts/physical_data_preparation.R")
+
+## Now we simply join the two datasets by their common monitoring period identifier
+## Physical variables will be duplicated to attach to each of the benthic swath replicates
+
+westend <- westend.norm %>% left_join(phys.dat.norm,by="period")
+
+rm(phys.dat.norm,westend.norm)
+
+### FINISH ###

### 2018-03-17 data/data preparation scripts/wave_height_ncdf4_extract.R

@@ -1,0 +1,49 @@
+# wave height- new source??
+# 
+# http://www.sccoos.org/data/waves/
+# USGS Coastal and Marine Geology Program
+# http://cmgwindwave.usgsportals.net/
+# options chosen in map data: 
+# Model Grid: ENP (Eastern North Pacific)
+# Model output: US Geophysical Fluid Dynamics Laboratory GFDL-ESM2M
+# Location: iooos/USGS station n46219
+# Latitude: 33.221 Longitude -119.882
+
+library(ncdf4)
+library(tidyverse)
+library(lubridate)
+
+modwaves <- nc_open("data/raw/wave_significant_height_enp_gfdlhistorical.nc")
+##[Not run] view structure of NCDF file
+# print(modwaves)
+
+times<-ncvar_get(modwaves,"time")
+
+# NCDF is pretty simple (only one longitude/latitude, so the variable wave height is one-dimensional)
+heights <- ncvar_get(modwaves,"sea_surface_wave_significant_height")
+
+# Maximum wave height defined as 
+# average height of the one third highest waves in the record over the time period
+# function takes a vector of wave heights, calculates the 2/3rds quantile, and then
+# averages the observations greater than that cutoff
+calc_maxHs <- function(Hs) {
+  cutoff <- quantile(Hs,2/3)
+  mean(Hs[Hs>cutoff])
+}
+# Convert to data frame
+wavesdf <- data_frame(time=times,Hs=heights,date=as_date(as.POSIXct(time,origin="1970-01-01 00:00:00"))) %>%
+  #Calculate montly means
+  mutate(year=year(date),month=month(date)) %>%
+  
+  # Calculate the average height of one third highest waves for each month
+  group_by(year,month) %>%
+  mutate(maxHs=calc_maxHs(Hs)) %>%
+  
+  # Finally, filter so there's just one value for each month
+  distinct(year,month,maxHs)
+
+
+# Remove unneeded data from workspace
+rm(calc_maxHs,modwaves,times,heights)
+
+### FINISH ###

### 2018-05-04 EDM_SNI_analysis.Rmd
updated and cleaned data prep

@@
 date: "March 13, 2018"
-output: html_document
+output: pdf_document
 bibliography: citation_bibtex.bib
@@
 require(knitr)
+require(kableExtra)
 require(extrafont)
@@
   # define column "plotting" as a factor so we can later control plotting order on figures
-  mutate(plotting=factor(plotting,levels=c("(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST","L. far","M. pyr","M. pyr (j)","P. cal","S. pur","M. fra","Constant")))
+  mutate(plotting=factor(plotting,levels=c("(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST","M. pyr","M. pyr (j)","L. far","P. cal","S. pur","M. fra","Constant")),
+         long=factor(long,levels=c("Macrocystis >1m","Macrocystis <1m","Pterygophora","Laminaria","Purple urchin","Red urchin","Multivariate ENSO Index","Pacific Decadal Oscillation","North Pacific Gyre Oscillation","Significant Wave Height","Sea Surface Temperature","Constant")))
 
@@
 rm(pd,period.key,periods.first,westend.long,westend.meandens)
+
+#ggplot theme to apply to plots later
+theme_rockwell <- function() {
+  theme_minimal()+
+  theme(text=element_text(family="Rockwell",size=12),
+        axis.title = element_text(family="Rockwell",size=14),
+        strip.background = element_rect(colour="black"),
+        panel.border = element_rect(color="black",fill=NA))
+}
 ```
@@
 ```
-
-***
 
@@
 
-westend.bestE["mac"] <- 6
 westend.bestE["ymac"] <- 8
@@
 
-## 2. Convergent Cross Mapping
+***
+
+## 3. Convergent Cross Mapping
 
@@
 
-## 3. Multivariate Models Using CCM results
+***
 
-For each site and each species, we'll only use the variables that show a causal link to build multivariate models, also removing those variables that did not pass diagnositic tests (simplex or prediction horizon).
-We find the causal variables for each species, based on the CCM output for that species.
+## 4. Multivariate Models Using CCM results
+
+Every EDM method we have used so far involves reconstructing an attractor in state-space and then examining that attractor to make predictions.
+In Step 2, we used single time series and their lags to build attractors, and examined the attractors for evidence of the degree of dimensionality, predictability, and nonlinearity.
+In Step 3, we built attractors from one species and searched for the signature of other species' dynamics encoded in those attractors, leading to inferences about causation.
+
+Multivariate models have the same structure, except instead of using single variables to reconstruct the attractors, we use contemporaneous values of multiple variables [@Deyle2016].
+Instead of library vectors or points in state-space taking the form of, for example, $<x_t,x_{t-1},x_{t-2}>$, they now are formed in true multivariate space, e.g.
+$<Kelp_{t},Urchin_{t},Nutrients_{t}>$.
+Using the normalized time series, we will build library vectors that each include the variable we are trying to predict, as well as other causal variables illuminated by our CCM analysis.
+Then we use S-maps as above to forecast the target variable using leave-one-out cross-validation just as in Step 3.
+
 ```{r causal vars fxn,message=F,echo=F}
@@
 
+***
+
+### Interaction Network
+
+The CCM results translated into an empirical interaction web looks like this:
+
+```{r igraph, echo=F, message=F,warning=F}
+require(igraph) # R Package for network graphing and analysis
+
+# Add NAs for connections in to physical variables (since we want them just to be exogenous)
+connect.mat <- t(westend.ptot.mat) %>% 
+  as.data.frame() %>%
+  mutate(mei=rep(NA,nrow(.)),pdo=rep(NA,nrow(.)),npgo=rep(NA,nrow(.)),waves=rep(NA,nrow(.)),sst=rep(NA,nrow(.)))
+
+g <- graph.adjacency(as.matrix(connect.mat))
+par(mfrow=c(1,1))
+
+# Variable names
+
+names <- as.character(sapply(V(g)$name, function(x) namekey$plotting[match(x,namekey$short)]))
+names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
+
+# Vertex labels
+V(g)$name <-names
+vert_in <- sapply(degree(g,mode="in"),function(x) paste0("(",x,")"))
+vert_out_in <- paste0(degree(g,mode="out"),vert_in)
+V(g)$label <- paste(sep="\n",V(g)$name, vert_out_in)
+V(g)$label[7:11] <- paste(V(g)$name[7:11])
+
+#vertex coordinates
+crds <- cbind(c(300,240,375,313,158,122,430,203,146.3342,309,20),c(10,126,102,206,204,85,162.8175,0,314.3077,318,232.6699))
+
+# vector of node colors for plotting
+vert.cols <- c(rep("tan3",3),rep("orchid3",2),"tan3",rep("gray70",5))
+
+plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.size=40,edge.lty=2,edge.color="black",vertex.color=vert.cols,layout=crds,vertex.frame.color=F,main="",edge.curved=F)
+```
+
+In this network, each node represents a variable, and the numbers associated with each are the number of causal connections in the web, outgoing and (incoming).
+The maximum number of outgoing interactions is 5 (6 total biological variables minus the variable itself), and incoming is 10 (5 biological and 5 physical variables).
+Algae species are in brown, while the two urchin species are in purple and the physical driver are in grey.
+For example, as mentioned above, we can see that young *Macrocystis* is forced by all other variables in the system, but only itself shows a causal forcing on 3 of the 5 other biological variables.
+Again we can see that while the web is not entirely saturated, there are many unidirectional forcings, and a lot of bidirectional causation as well (A forces B *and* B forces A).
+
+***
+
+### Dynamic Species Interaction Models
+
+```{r general smap model fxn, include=F}
+# Generalized function to build a multivariate S-map model.
+Function takes the data, a data frame of the segments (i.e., the denoting the different within-site swaths denoting breaks in the time series), the target species (the one for which we are building a model), and a character vector of the other, causal variables.
+The function returns the fitted model and the coefficients (i.e., the species interaction partials), the model statistics, and a box plot of the interactions coefficients for that species across all the data.
+
+smap_multi <- function(sitedat,sitesegs,species,causalvars) {
+
+  full.mod <- block_lnlp(sitedat,lib=as.matrix(sitesegs),pred=as.matrix(sitesegs),columns=causalvars,target_column = species,theta = c(0, 1e-04, 3e-04, 0.001,0.003, 0.01, 0.1, 0.5, 1, 2, 4, 6,8), num_neighbors=0,method="s-map",silent=T)
+
+  opttheta <- full.mod$theta[full.mod$rho==max(full.mod$rho)]
+
+  best.mod <- block_lnlp(sitedat,lib=as.matrix(sitesegs),pred=as.matrix(sitesegs),columns=causalvars,target_column = species,theta = opttheta, num_neighbors=0,method="s-map",save_smap_coefficients=T,silent=T)
+
+  # Model output-- time, observations, predictions. time is truncated, so re-expand it to fit with original periods
+  t <- best.mod$model_output[[1]]$time
+
+  # Save interaction coefficients
+  coeffs <- best.mod$smap_coefficients[[1]]
+  names(coeffs) <- c(causalvars,"const")
+  coeffs <- coeffs %>% mutate(time=t)%>%complete(time=full_seq(x=c(1,630),1)) %>% mutate(period=rep(1:63,10))
+
+  # long form for plotting
+  coeffs.long <- coeffs %>%
+    gather(key=effect,value=value,-period,-time,na.rm=T) %>%
+    left_join(namekey, by=c("effect"="short"))
+
+  # box plot of interactions
+  sppname <- namekey$long[match(species,namekey$short)]
+  int.box <- ggplot(coeffs.long,aes(x=plotting,y=value))+
+    geom_boxplot() +
+    geom_hline(yintercept=0,linetype=2,col="gray40")+
+    xlab("Interacting Variable") +
+    ylab("Coefficient") +
+    ggtitle(paste("Interactive Effects on Dynamics of ",sppname))
+
+  ## Outputs (to go in a list)
+
+  outlist <- list(model=best.mod,coeffs=coeffs,coeffs.long= coeffs.long,plot=int.box)
+  return(outlist)
+}
+```
+
+```{r apply multivariate models, include=F,warnings=F}
+# Apply to the West End data
+westend.multi.mods <- purrr::map(list("mac","purp","lam","ymac","red","pter"),function(x) {
+  smap_multi(westend,westend.segs,x,westend.causal.vars[[x]])
+})
+names(westend.multi.mods) <- c("mac","purp","lam","ymac","red","pter")
+```
+
+When we construct multivariate models for the six variables in the data based on their causal relationships, we get multivariate S-map models that perform well, displaying a high level of correlation between out-of-sample predictions and observed values, and relatively low mean absolute error (MAE).
+
+```{r model perf,echo=F}
+# collect model performance statistics
+westend.model.perf <- tibble(Species=c("Macrocystis","Purple Urchin","Laminaria","Young Macrocystis","Red Urchin","Pterygophora"),
+                             Predictors=purrr::map(westend.causal.vars,.f=function(x){
+                               y=namekey$long[match(x,namekey$short)];paste(y,sep=",",collapse=", ")}),
+                             Rho = map_dbl(westend.multi.mods,function(x) x$model$rho),
+                             MAE=map_dbl(westend.multi.mods,function(x) x$model$mae),
+                             mod=c("mac","purp","lam","ymac","red","pter"))
+westend.model.perf %>% select(-mod) %>% 
+  kable("latex",col.names = c("Modeled Species","Predictors","$\\rho$","MAE"),digits=3,escape=F) %>%
+  kable_styling(full_width = F) %>%
+  column_spec(1,bold=T)%>%
+  column_spec(2, width = "30em")
+```
+
+```{r all model effects join, echo=F,warning=F}
+# Join all coefficients
+all.mod.coeffs.long <- purrr::map(names(westend.multi.mods),function(x){
+  westend.multi.mods[[x]][["coeffs.long"]] %>% mutate(mod=x)
+  }
+  )%>%
+  bind_rows() %>%
+  mutate(mod_plotting_name = namekey$long[match(mod,namekey$short)])
+
+# assign interaction types: competition, herbivory (effect of urchins on algae), consumption (effect of algae on urchin), or growth
+algaes<-c("lam","mac","pter","ymac")
+urchins <- c("red","purp")
+combinations <- expand.grid(study_spp,study_spp) %>% 
+  rename(mod=Var1,effect=Var2) %>%
+  select(mod,effect)%>%
+  mutate(type = case_when(
+    mod ==   effect ~ "intraspecies",
+    mod == "mac" & effect == "ymac" ~ "intraspecies",
+    mod == "ymac" & effect == "mac" ~ "intraspecies",
+    mod %in% algaes & effect %in% algaes ~ "algal competition",
+    mod %in% urchins & effect %in% urchins ~ "urchin competition",
+    mod %in% algaes & effect %in% urchins ~ "herbivory",
+    mod %in% urchins & effect %in% algaes ~ "consumption",
+    TRUE ~ "NA"
+    ),
+    type=factor(type,levels=c("algal competition","urchin competition","herbivory","consumption","intraspecies")),
+    spp_type = case_when(
+      mod %in% algaes ~ "algae",
+      mod %in% urchins ~ "urchin",
+      TRUE ~ "none"
+    )
+  ) %>%
+  mutate_at(vars(mod,effect),as.character)
+
+# add rho label for facet plotting
+mod_rho_labels <- westend.model.perf %>%
+  mutate(label1="rho==",Rho_round=round(Rho,2))%>%
+  unite(label,label1,Rho_round,sep="",remove=F)%>%
+  select(mod,label)
+
+all.mod.coeffs.long %>%
+  #add the type of interaction
+  left_join(combinations,by=c("mod","effect"))%>%
+  left_join(mod_rho_labels,by="mod") -> all.mod.coeffs.long
+```
+
+***
+
+## 5. Results
+
+### Multivariate Model Output: Species interactions
+
+We can pull from the models every predicted interaction between species.
+These fitted interaction strengths are specifically the estimated effect of the density of one species on itself or another species, when forecasting one monitoring period (approximately six months) ahead.
+For example, each data point for *Pterygophora* in the *Macrocystis* model is the predicted effect of the current density of Pterygophora on the density of Macrocystis in the next monitoring period.
+Mathematically, each interaction measured is a partial derivative, or element of our sequentially calculated Jacobian matrices, estimating the effect of one species on another, $\delta N_{1}/\delta N_{2}$.
+The distributions of those interactions are shown in the figure below.
+
+```{r boxplot of species interactions, echo=F,warning=F,message=F,fig.height=8,fig.width=7.5}
+# boxplot of all species interactions
+all.mod.coeffs.plot.box <- all.mod.coeffs.long %>%
+  #filter for just species (no physical vars yet)
+  filter(effect %in% study_spp) %>%
+  #plot
+  ggplot(aes(x=plotting,y=value,fill=type))+
+  geom_boxplot(outlier.alpha=0.4,outlier.size=1,width=0.8,position = "identity")+
+  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  geom_hline(yintercept=0,linetype=2)+
+  geom_text(aes(x=1.2,y=1.8,label=label),size=6, parse=T,check_overlap=TRUE,show.legend = FALSE)+
+  ylim(-2,2)+
+  coord_flip()+
+  labs(y="Estimated Interaction Coefficient",x="Interacting Species")+
+  facet_grid(mod_plotting_name~.,scales="free",space="free")+
+  theme_rockwell()
+
+all.mod.coeffs.plot.box
+# ggsave("all_spp_coffs.png",plot=all.mod.coeffs.box,height=8,width=7.5)
+```
+
+These are box-and-whisker plots of estimated species interaction strengths from S-map models for the five focal species (panels top to bottom): *Macrocystis* adults, *Macrocystis* juveniles, *Pterygophora*, *Laminaria*, purple urchin, and red urchin.
+Each colored box represents the distribution of all estimated interaction coefficients (x-axis) of an interacting or forcing species (left y-axis) on a modeled species (right y-axis) across all data for a given model (vertical lines: median; box: interquartile range; whiskers extend to data point at most $1.5*IQR$ from the box).
+Correlation coefficient between predictions and observations denoted for each model.
+Color denotes hypothesized interaction type, including interspecific competition (between algae species or between urchin species), herbivory (urchin effect on algae), consumption (algae effect on urchins), and intraspecies interaction (the estimated interaction of a species with itself).
+Abbreviations: S.
+pur, *Strongylocentrotus purpuratus*; P cal., *Pterygophora californica*; M.
+pyr, *Macrocystis pyrifera*; M.
+fra, *Mesocentrotus franciscanus*; L.
+far, *Laminaria farlowii*.
+
+It is clear from the boxplots that fitted intra- and interspecific interactions display evidence of positive, negative, and neutral interactions.
+Over the range of conditions in the time series, each species variable with the exception of young *Macrocystis* has positive intraspecific effects; that is, each species’ density has a positive estimated effect on itself, a combination of survivorship and new recruitment.
+Relative to the magnitude of other interactions, this intraspecific effect is strongest in the two urchin species.
+
+The interactions of the urchin and algal species show evidence of herbivory (green boxes).
+Both red and purple urchin density have predominantly negative effect on adult and juvenile *Macrocystis*, *Laminaria*, and *Pterygophora*, effects that were consistent with our hypotheses.
+However, the converse effects of algal density on urchin density are generally small, and not consistently positive.
+
+Few strong and consistent competitive interactions (red boxes) are observed in model estimates.
+There is evidence of competition between the two urchin species, but the interaction is small relative to other forcing factors in the urchin models, and is occasionally positive (apparent mutualism).
+Among the algal species, the clearest negative competitive effect is that of *Pterygophora* on adult *Macrocystis* (top panel).
+Other interactions are not always negative (competitive), and contrary to expectations, positive values were estimated for many interactions.
+*Pterygophora* has only intermittent negative effects on young *Macrocystis*, and *Pterygophora* and *Laminaria* display an apparent mutualism.
+Perhaps most surprisingly, *Macrocystis* does not display a consistent negative effect on *Pterygophora* or *Laminaria*, contrary to our hypothesis of *Macrocystis* competitive dominance [@Dayton1999].
+
+***
+
+### Multivariate Model Output: Physical Forcing
+
+We can also investigate the effects of the physical forcing variables on the biological dynamics in a similar plot:
+
+```{r phys coeffs,echo=F,message=F,warning=F}
+# with boxplot
+all.phys.coeffs.plot <- all.mod.coeffs.long %>%
+  filter(effect %in% phys.vars)%>%
+  mutate(label = case_when(
+    mod == "mac" ~ "Macrocystis >1m",
+    mod == "ymac" ~ "Macrocystis <1m",
+    mod == "lam" ~ "Laminaria",
+    mod == "pter" ~ "Pterygophora",
+    mod == "red" ~ "Red urchin",
+    mod == "purp" ~ "Purple urchin")
+  ) %>%
+  ggplot(aes(effect,value))+
+  scale_x_discrete(labels=c(
+    "mei"="MEI",
+    "npgo"="NPGO",
+    "pdo"="PDO",
+    "waves"="SWH",
+    "sst"="SST")
+    )+
+  geom_boxplot(fill="dodgerblue2",outlier.alpha=0.4,position = "identity")+
+  geom_hline(yintercept=0,linetype=2)+
+  ylim(-2,2)+
+  coord_flip()+
+  labs(y="Estimated Interaction Coefficient",x="Physical Forcing")+
+  facet_grid(mod_plotting_name~.)+
+  geom_text(aes(x=1.5,y=2,label=label),hjust=1,family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
+  theme_rockwell()+
+  theme(strip.background = element_blank(),
+        strip.text = element_blank())
+all.phys.coeffs.plot
+```
+
+In this figure, each box represents the distribution of estimated effect of a physical driver (left y-axis) on the modeled species across all data for a given model.
+Empty rows indicate physical variables that were not included in specific species models because of a lack of a causal signal from our CCM.
+Abbreviations: SWH, significant wave height; SST, sea surface temperature; PDO, Pacific Decadal Oscillation; NPGO, North Pacific Gyre Oscillation; MEI, Multivariate ENSO Index.
+
+Physical forcing variables affect the dynamics of all species in the study, to varying degrees.
+*Macrocystis* is the only species in the analysis that displays a significant cross-mapping signal (significant causal forcing) with all five included physical variables.
+The negative effects of SWH and SST on adult *Macrocystis*, and the positive effect of the *NPGO*, are consistent with our expectations [@Bell2015].
+However, the PDO index has a positive effect on *Macrocystis*, contrary to expectation because in general, positive values of the PDO are associated with warmer SST and nutrient-poor conditions in the northeast Pacific.
+*Macrocystis* recruits, on the other hand, are positively affected by the MEI, SWH, and SST, and negatively affected by the PDO and NPGO.
+Together, the effects of physical forcing on *Macrocystis* suggest that conditions that are poor for survivorship of adults (higher SST, less nutrient availability, greater disturbance) may produce favorable conditions for recruitment of new sporophytes.
+
+The physical variables are more influential drivers of *Macrocystis* than *Pterygophora* or *Laminaria* dynamics.
+The PDO is the only physical variable casaully linked to *Laminaria* density, and it has predominantly negative but variable effect.
+The SWH, NPGO, MEI, and SST indices do not significantly drive *Laminaria* density.
+Conversely, although the PDO, NPGO, and SST were identified as causal variables for *Pterygophora* density, the magnitudes of their effects are small and inconsistent.
+
+The effects of physical forcing are not limited to the algal species, although the magnitude of physical forcing effects on red urchins are small, especially compared to the effect of their own density (the intraspecific effect).
+For the purple urchin, the PDO, NPGO, and MEI all have negative but variable effects.
+For red urchin, the effects of SWH and SST are small and negative, with the PDO and NPGO have variable effects.
+
+***
+
+### Physical Forcing Effect on Interactions
+
+A major observation from the multivariate models is that many interactions between species and between species and their environments can be variable (positive or negative, depending on conditions).
+For the algal species especially, the combination of variable interspecific interaction strengths and occasionally strong environmental forcing effects suggests that the strength of species interactions may be driven by environmental context.
+
+We can observe this in our data by plotting the same interactions from the previous step against levels of the environmental variables.
+We will investigate the effect of environmental variables on mean interaction strength for the different interaction types (herbivory, competition, consumption).
+
+```{r mean interactions, echo=F,warning=F,message=F}
+# Mean interactions over time
+mean_ints_by_time <- all.mod.coeffs.long %>%
+  select(period,mod,effect,value,type,spp_type) %>%
+  group_by(period,type) %>%
+  summarise(meanint=mean(value,na.rm=T),sdint=sd(value,na.rm=T))%>%
+  mutate(lowerint=meanint-sdint,upperint=meanint+sdint) %>% 
+  group_by(type)%>%
+  complete(period=full_seq(period,1))
+
+# mean_ints_by_time %>%
+#   filter(!is.na(type),type!="intraspecies") %>%
+#   ggplot(aes(x=period,y=meanint,ymax=upperint,ymin=lowerint,color=type))+
+#   geom_line()+
+#   geom_hline(yintercept=0,linetype=2)+
+#   scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="Interaction\ntype")+
+#   coord_cartesian(ylim=c(-0.5,0.5))+
+#   labs(title="Mean species interactions\nby type over time",x="Monitoring Period",y="Interaction Strength")+
+#   theme_rockwell()
+
+## Mean interactions vs. physical conditions
+# join all original density variables to interaction data
+
+all.coeffs.with.phys<- westend %>%
+  select(one_of(c(phys.vars,"period"))) %>%
+  distinct() %>%
+  right_join(all.mod.coeffs.long,by="period")%>%
+  gather("external_forcing","ext_value",one_of(phys.vars))
+
+all.coeffs.with.phys %>%
+  filter(!is.na(type))%>%
+  ggplot(aes(ext_value,value))+
+  geom_point()+
+  geom_smooth()+
+  ylim(-2,2)+
+  labs(title="Physical forcing effect on species interactions",x="Normalized Index Value",y="Interaction Strength")+
+  facet_grid(type~external_forcing)+
+  theme_rockwell()
+
+# absolute value of interactions
+all.coeffs.with.phys %>%
+  filter(!is.na(type))%>%
+  mutate(absvalue=abs(value))%>%
+  ggplot(aes(ext_value,absvalue))+
+  geom_point()+
+  geom_smooth()+
+  labs(title="Physical forcing effect on species interactions",x="Normalized Index Value",y="Interaction Strength")+
+  facet_grid(type~external_forcing)+
+  theme_rockwell()
+```
+
 ## References

### 2018-05-04 markdown_whiteboard.Rmd
updated and cleaned data prep

---
output:
  pdf_document: default
---
We now have a trove of new data: data describing interactions between species, interactions that vary with the state of ecosystem. In general, the multivariate models perform well (i.e. achieve good preditive power).

```{r multivariate model stats,message=F,warning=F}
library(knitr)
library(tidyverse)
require(kableExtra)

x<-data.frame(spp=c(1,2,3),theta=c(3,4,5),rho=c(.89,.99,.32),mae=c(.4,.5,.6),
                lngstr=rep("this is a long string. this is a long string. this is a long string. this is a long string.this is a long string. this is a long string.this is a long string. this is a long string.this is a long string. this is a long string.this is a long string. this is a long string.",3))

knitr::kable(x,format = "latex",col.names=c("Species","Nonlinearity", "$\\rho$","MAE","String"),
             caption="Multivariate S-map Model Statistics, with nonlinearity,
             correlation of predictions with observations, and mean absolute error.",
             escape=F) %>%
  kable_styling(full_width = F) %>%
  column_spec(3,bold=T)%>%
  column_spec(5, width = "30em")

```

### 2018-05-04 data/data preparation scripts/physical_data_preparation.R

@@ -125,72 +125,34 @@ npgo.per <- npgo %>% rename(npgo='NPGO index') %>%
 #### Significant Wave Height ####
 
 # Two sources of wave data
-# first is buoy data from CDIP
-# source: http://cdip.ucsd.edu
-# Wave height and sea surface temperature collected at two nearby buoys
+# Almost all from Tom Bell (newly modeled)
 
-waves.buoy <- read_csv("data/raw/Begg_SNI_CDIP.csv",col_types = cols()) %>%
-
-# Variable Hs denotes significant wave height in meters; 
+# Variable max_Hs denotes significant wave height in meters; 
 # Derived from the zeroeth moment of the reported energy spectrum.
 # Described as the "average height of the one third highest waves in the record over the time period"
-  
-  select(Dataset,Year,Month,Mean_Hs)
-
-waves.per <- waves.buoy %>%
-  
-  # Use the periods key to assign periods to each month, and remove irrelevant data
-  left_join(period.key,by=c("Year"="year","Month"="month")) %>%
-  filter(!is.na(period)) %>%
-  
-  # group by period, and find Maximum Hs during each period
-  # this differs from the averaging we did for MEI, PDO, and NPGO
-  group_by(period) %>%
-  summarise(waves.max=max(Mean_Hs)) %>%
-  
-  # normalize to zero mean, unit variance
-  mutate(waves.norm=(waves.max-mean(waves.max))/sd(waves.max))
 
 # Second source is from
 # http://www.sccoos.org/data/waves/
 # USGS Coastal and Marine Geology Program
 # http://cmgwindwave.usgsportals.net/
 # Modeled waves from the Geophysical Fluid Dynamics Laboratory
-# Complete, but only goes through 2005. We will join the normalized datasets together to form a more complete picture
-# Code that produced the below .csv is available in this directory as "wave_height_ncdf4_extract.R"
+# Complete, but only goes through 2005. We joined the datasets together to form a more complete picture
+# Code that produced the below is available in this directory
 
-source("data/data preparation scripts/wave_height_ncdf4_extract.R")
+source("data/data preparation scripts/wave_height_Bell.R")
 
-waves.modeled <- wavesdf %>% 
+
[…]

### 2018-05-04 data/data preparation scripts/wave_height_Bell.R

@@ -1,0 +1,95 @@
+# Wave height data from Tom Bell
+
+require(tidyverse)
+require(lubridate)
+
+hs <- read_csv("data/raw/SanNic_Hs.csv") %>%
+  # Site 2 is the West End of SNI
+  select(Year,Month,Day,Max_2) %>%
+  unite("date",Year,Month,Day,sep="-",remove=F)%>%
+  mutate(date=as.Date(date)) %>%
+  rename(maxHs=Max_2)%>%
+  # Monthly mean of daily Max Hs
+  # group_by(Year,Month) %>%
+  # summarise(maxHs=mean(Max_2,na.rm=T)) %>%
+  rename(year=Year,month=Month,day=Day)%>%
+  ungroup()
+hs$maxHs[is.na(hs$maxHs)]<-NA
+
+### USGS DATA TO COMPARE ###
+
+# 
+# http://www.sccoos.org/data/waves/
+# USGS Coastal and Marine Geology Program
+# http://cmgwindwave.usgsportals.net/
+# options chosen in map data: 
+# Model Grid: ENP (Eastern North Pacific)
+# Model output: US Geophysical Fluid Dynamics Laboratory GFDL-ESM2M
+# Location: iooos/USGS station n46219
+# Latitude: 33.221 Longitude -119.882
+
+library(ncdf4)
+
+modwaves <- nc_open("data/raw/wave_significant_height_enp_gfdlhistorical.nc")
+##[Not run] view structure of NCDF file
+# print(modwaves)
+
+times<-ncvar_get(modwaves,"time")
+
+# NCDF is pretty simple (only one longitude/latitude, so the variable wave height is one-dimensional)
+heights <- ncvar_get(modwaves,"sea_surface_wave_significant_height")
+
+# Maximum wave height defined as 
+# average height of the one third highest waves in the record over the time period
+# function takes a vector of wave heights, calculates the 2/3rds quantile, and then
+# averages the observations greater than that cutoff
+calc_maxHs <- function(Hs) {
+  cutoff <- quantile(Hs,2/3)
+  mean(Hs[Hs>cutoff])
+}
+# Convert to data frame
+wavesdf <- data_frame(time=times,Hs=heights,date=as_date(as.POSIXct(time,origin="1970-01-01 00:00:00"))) %>%
+  #Calculate montly means
[…]

### 2018-05-04 data/data preparation scripts/wave_height_ncdf4_extract.R
wave height usgs

@@ -10,8 +10,6 @@
 # Latitude: 33.221 Longitude -119.882
 
 library(ncdf4)
-library(tidyverse)
-library(lubridate)
 
 modwaves <- nc_open("data/raw/wave_significant_height_enp_gfdlhistorical.nc")
 ##[Not run] view structure of NCDF file
@@ -33,14 +31,15 @@ calc_maxHs <- function(Hs) {
 # Convert to data frame
 wavesdf <- data_frame(time=times,Hs=heights,date=as_date(as.POSIXct(time,origin="1970-01-01 00:00:00"))) %>%
   #Calculate montly means
-  mutate(year=year(date),month=month(date)) %>%
+  mutate(year=year(date),month=month(date),day=day(date)) %>%
   
   # Calculate the average height of one third highest waves for each month
   group_by(year,month) %>%
   mutate(maxHs=calc_maxHs(Hs)) %>%
   
   # Finally, filter so there's just one value for each month
-  distinct(year,month,maxHs)
+  distinct(year,month,maxHs) %>%
+  ungroup()
 
 
 # Remove unneeded data from workspace

### 2018-05-04 data/data preparation scripts/wave_height_Bell.R
wave height 2

@@ -93,3 +93,10 @@ wavesdf %>%
   geom_line()+
   scale_color_discrete(labels=c("Bell","USGS"))+
   theme_minimal()
+
+wavesdf %>%
+  ggplot(aes(maxHs,maxHs_usgs))+
+  geom_point()+
+  geom_abline(slope=1,intercept=0,linetype=2)+
+  labs(title="Max Daily Wave Height",x="Bell height",y="USGS height")+
+  theme_minimal()

### 2018-05-31 EDM_SNI_analysis.Rmd
added description of effect of physical oceanography on species interactions

@@
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
+
+```{r global_options, include=F,message=F}
+require(knitr)
+knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message = FALSE)
+```
 
@@
 
-westend.bestE["ymac"] <- 8
+westend.bestE["ymac"] <- 3
 westend.bestE["pter"] <- 6
@@
   geom_boxplot(outlier.alpha=0.4,outlier.size=1,width=0.8,position = "identity")+
+  stat_summary(fun.y=mean, colour="white", geom="point", 
+               shape=17, size=2,show.legend = FALSE)+
   scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
   geom_hline(yintercept=0,linetype=2)+
-  geom_text(aes(x=1.2,y=1.8,label=label),size=6, parse=T,check_overlap=TRUE,show.legend = FALSE)+
+  geom_text(aes(x=1.2,y=1.8,label=label),size=4, parse=T,check_overlap=TRUE,show.legend = FALSE)+
   ylim(-2,2)+
@@
 all.mod.coeffs.plot.box
-# ggsave("all_spp_coffs.png",plot=all.mod.coeffs.box,height=8,width=7.5)
+# ggsave("all_spp_coffs.png",plot=all.mod.coeffs.plot.box,height=8,width=7.5)
 ```
@@
 
-```{r phys coeffs,echo=F,message=F,warning=F}
+```{r phys coeffs,echo=F,message=F,warning=F,fig.height=8}
 # with boxplot
@@
   geom_boxplot(fill="dodgerblue2",outlier.alpha=0.4,position = "identity")+
+  stat_summary(fun.y=mean, colour="white", geom="point", 
+               shape=17, size=2,show.legend = FALSE)+
   geom_hline(yintercept=0,linetype=2)+
@@
 
-We can observe this in our data by plotting the same interactions from the previous step against levels of the environmental variables.
-We will investigate the effect of environmental variables on mean interaction strength for the different interaction types (herbivory, competition, consumption).
+In order to observe this, we remove physical forcings and re-run the multivariate models with just the species interactions, and then look at the resulting trends through the lens of various physical forcing (to remove endogeneity).
+We will investigate the effect of environmental variables on interaction strength for the different interaction types (herbivory, competition, and intraspecies effects).
 
-```{r mean interactions, echo=F,warning=F,message=F}
-# Mean interactions over time
-mean_ints_by_time <- all.mod.coeffs.long %>%
-  select(period,mod,effect,value,type,spp_type) %>%
-  group_by(period,type) %>%
-  summarise(meanint=mean(value,na.rm=T),sdint=sd(value,na.rm=T))%>%
-  mutate(lowerint=meanint-sdint,upperint=meanint+sdint) %>% 
-  group_by(type)%>%
-  complete(period=full_seq(period,1))
+```{r model spp_only,warning=F,fig.height=8,fig.width=7.5}
+# remove physical variables from predictors then run models
+causal.vars.spp <- purrr::map(westend.causal.vars,function(x) setdiff(x,phys.vars))
+multi.mods.spp.only <- purrr::map(list("mac","purp","lam","ymac","red","pter"),function(x) {
+  smap_multi(westend,westend.segs,x,causal.vars.spp[[x]])
+})
+names(multi.mods.spp.only) <- c("mac","purp","lam","ymac","red","pter")
 
-# mean_ints_by_time %>%
-#   filter(!is.na(type),type!="intraspecies") %>%
-#   ggplot(aes(x=period,y=meanint,ymax=upperint,ymin=lowerint,color=type))+
-#   geom_line()+
-#   geom_hline(yintercept=0,linetype=2)+
-#   scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="Interaction\ntype")+
-#   coord_cartesian(ylim=c(-0.5,0.5))+
-#   labs(title="Mean species interactions\nby type over time",x="Monitoring Period",y="Interaction Strength")+
-#   theme_rockwell()
+# collect model performance statistics
+spp.only.model.perf <- tibble(Species=c("Macrocystis","Purple Urchin","Laminaria","Young Macrocystis","Red Urchin","Pterygophora"),
+                             Predictors=purrr::map(causal.vars.spp,.f=function(x){
+                               y=namekey$long[match(x,namekey$short)];paste(y,sep=",",collapse=", ")}),
+                             Rho = map_dbl(multi.mods.spp.only,function(x) x$model$rho),
+                             MAE=map_dbl(multi.mods.spp.only,function(x) x$model$mae),
+                             mod=c("mac","purp","lam","ymac","red","pter"))
+spp.only.model.perf %>% select(-mod) %>% 
+  kable("latex",col.names = c("Modeled Species","Predictors","$\\rho$","MAE"),digits=3,escape=F) %>%
+  kable_styling(full_width = F) %>%
+  column_spec(1,bold=T)%>%
+  column_spec(2, width = "30em")
 
+# Join all coefficients
+spp.only.mod.coeffs <- purrr::map(names(multi.mods.spp.only),function(x){
+  multi.mods.spp.only[[x]][["coeffs.long"]] %>% mutate(mod=x)
+  }
+  )%>%
+  bind_rows() %>%
+  mutate(mod_plotting_name = namekey$long[match(mod,namekey$short)])
+
+# assign interaction types: competition, herbivory (effect of urchins on algae), consumption (effect of algae on urchin), or growth
+algaes<-c("lam","mac","pter","ymac")
+urchins <- c("red","purp")
+combinations <- expand.grid(study_spp,study_spp) %>% 
+  rename(mod=Var1,effect=Var2) %>%
+  select(mod,effect)%>%
+  mutate(type = case_when(
+    mod ==   effect ~ "intraspecies",
+    mod == "mac" & effect == "ymac" ~ "intraspecies",
+    mod == "ymac" & effect == "mac" ~ "intraspecies",
+    mod %in% algaes & effect %in% algaes ~ "algal competition",
+    mod %in% urchins & effect %in% urchins ~ "urchin competition",
+    mod %in% algaes & effect %in% urchins ~ "herbivory",
+    mod %in% urchins & effect %in% algaes ~ "consumption",
+    TRUE ~ "NA"
+    ),
+    type=factor(type,levels=c("algal competition","urchin competition","herbivory","consumption","intraspecies")),
+    spp_type = case_when(
+      mod %in% algaes ~ "algae",
+      mod %in% urchins ~ "urchin",
+      TRUE ~ "none"
+    )
+  ) %>%
+  mutate_at(vars(mod,effect),as.character)
+
+# add rho label for facet plotting
+mod_rho_labels <- spp.only.model.perf %>%
+  mutate(label1="rho==",Rho_round=round(Rho,2))%>%
+  unite(label,label1,Rho_round,sep="",remove=F)%>%
+  select(mod,label)
+
+spp.only.mod.coeffs %>%
+  #add the type of interaction
+  left_join(combinations,by=c("mod","effect"))%>%
+  left_join(mod_rho_labels,by="mod") -> spp.only.mod.coeffs
+
+# boxplot of all species interactions
+spp.only.mod.coeffs.plot.box <- spp.only.mod.coeffs %>%
+   filter(effect %in% study_spp) %>%
+  #plot
+  ggplot(aes(x=plotting,y=value,fill=type))+
+  geom_boxplot(outlier.alpha=0.4,outlier.size=1,width=0.8,position = "identity")+
+  stat_summary(fun.y=mean, colour="white", geom="point", 
+               shape=17, size=2,show.legend = FALSE)+
+  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  geom_hline(yintercept=0,linetype=2)+
+  geom_text(aes(x=1.2,y=1.8,label=label),size=4, parse=T,check_overlap=TRUE,show.legend = FALSE)+
+  ylim(-2,2)+
+  coord_flip()+
+  labs(y="Estimated Interaction Coefficient",x="Interacting Species")+
+  facet_grid(mod_plotting_name~.,scales="free_y",space="free")+
+  theme_rockwell()
+
+spp.only.mod.coeffs.plot.box
+# ggsave("spp_only_mods_coeffs.png",plot=spp.only.mod.coeffs.plot.box,height=8,width=7.5)
+```
+
+```{r mean interactions, fig.height=8,fig.width=7.5}
 ## Mean interactions vs. physical conditions
@@
   distinct() %>%
-  right_join(all.mod.coeffs.long,by="period")%>%
+  right_join(spp.only.mod.coeffs,by="period")%>%
   gather("external_forcing","ext_value",one_of(phys.vars))
 
-all.coeffs.with.phys %>%
-  filter(!is.na(type))%>%
-  ggplot(aes(ext_value,value))+
-  geom_point()+
-  geom_smooth()+
-  ylim(-2,2)+
-  labs(title="Physical forcing effect on species interactions",x="Normalized Index Value",y="Interaction Strength")+
-  facet_grid(type~external_forcing)+
-  theme_rockwell()
-
-# absolute value of interactions
-all.coeffs.with.phys %>%
-  filter(!is.na(type))%>%
-  mutate(absvalue=abs(value))%>%
-  ggplot(aes(ext_value,absvalue))+
-  geom_point()+
-  geom_smooth()+
-  labs(title="Physical forcing effect on species interactions",x="Normalized Index Value",y="Interaction Strength")+
-  facet_grid(type~external_forcing)+
-  theme_rockwell()
 ```
 
+### Effects on Individual Species
+
+With 10%/90% quantile regression.
+Columns/x-axis indicates the level of a physical stressor, while rows and y-axis denote the type of species interaction (corresponding to the colors in the boxplots above).
+
+```{r individual_effects,fig.height=6}
+
+phys_effects_on_interactions <- purrr::map(as.list(study_spp),function(spp) {
+  quants <- c(0.10,0.5,0.90)
+  plot_title <- namekey$long[match(spp,namekey$short)]
+  all.coeffs.with.phys %>%
+  filter(!is.na(type),mod==spp)%>%
+  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+    geom_point(col='gray30',alpha=0.7)+
+    geom_hline(yintercept = 0,linetype=2)+
+    geom_quantile(quantiles=quants,lineend = "round",size=1.5,formula=y~x)+
+    scale_color_manual(values=c('mac'="navyblue",'ymac'="gray50",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+
+    labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+
+    facet_grid(type~external_forcing,scales = "free_y")+
+    theme_rockwell()
+})
+phys_effects_on_interactions[[1]]
+phys_effects_on_interactions[[2]]
+phys_effects_on_interactions[[3]]
+phys_effects_on_interactions[[4]]
+phys_effects_on_interactions[[5]]
+phys_effects_on_interactions[[6]]
+```
+
+Here are some notable effects/trends for each species:
+
+* *Mesocentrotus franciscanus*
+    + chance of competition with *Strongylocentrotus purpuratus* seems reduced under higher physical (wave) stress
+    + the same (reduced competition) is marginally true for nutrient-rich conditions (higher NPGO);and the opposite for nutrient- stressed conditions (higher MEI, PDO indices)
+
+* *Strongylocentrotus purpuratus*
+    + "urchin competition" seems like a misnomer for this species; the effect of *Mesocentrotus franciscanus* on *S.
+purpuratus* is primarily positive across a range of physical conditions.
+The effect is dampened somewhat under increased wave disturbance
+    + Effect of *S.
+purpuratus* on itself ("population growth rate" in a way) is affected by oceanographic forcing: it is slightly increased under increased NPGO values and with increased disturbance, and slightly decreased under elevated MEI, PDO, and SST.
+
+* *Laminaria farlowii*
+    + as far as competition goes, most large, negative interactions are associated with juvenile *Macrocystis*, likely as an indicator of impending succesional processes (i.e., the juvenile *Macrocystis* growing into large *Macrocystis*).
+This expected negative effect is reduced under elevated PDO, disturbance, and SST, which makes sense: these are conditions that classically inhibit *Macrocystis* dominance[@Reed2011; @Cavanaugh2011; @Bell2015,@Young2015].
+    + In general, *M franciscanus* has a more negative effect on *Laminaria* than does *S.
+purpuratus*, and that effect is not overly sensitive to physical conditions, although the negative effect is dampened somewhat under elevated disturbance
+    + Physical conditions seem to strongly influence intraspecies *Laminaria* effects: the positive effect of *Laminaria* on its own density is reduced under elevated MEI, PDO, and SST, and increased under elevated NPGO and disturbance.
+
+* *Pterygophora californica*
+    + Algal effects: *Laminaria* is a good indicator of a positive density response of *Pterygophora*, while, as mentioned above (and surprisingly) the effect of *Macrocystis* is small.
+The positive effect of *Laminaria* increases with slightly with elevated NPGO and disturbance, and decreases with elevated MEI, PDO, and SST.
+    + Herbivory effects on *Pterygophora* vary with physical environment.
+*M.
+franciscanus* and *S.
+purpuratus* have similar mean effects on *Pterygophora*, but the effect of *M.
+franciscanus* is much more sensitive to physical conditions.
+Its negative effect is strongest under elevated disturbance and NPGO, while its effect is diminished under elevated MEI, PDO, and SST.
+The same trends hold for the effect of *S.
+purpuratus* on *Pterygophora*, but those effects are less sensitive.
+    + The variation in the intraspecies effect of *Pterygophora* is responsive to physical conditions in a similar manner to *Laminaria* (reduced under elevated MEI, PDO, and SST, and increased under elevated NPGO and disturbance)
+
+* *Macrocystis pyrifera*
+    + Because we have two variables measuring the density of different *Macrocystis* life stages, we can parse out the relative effects of physical conditions on adult vs. juvenile giant kelp.
+    + Environmental stress seems to strengthen the interactions between *Laminaria* and *Pterygophora* and *Macrocystis* recruits.
+Under conditions of elevated MEI, SST, or PDO, or decreased NPGO, the positive effect of *Pterygophora* and the negative effect of *Laminaria* on *Macrocystis* sporophytes are both enhanced.
+Similar is true for the adult *Macrocystis*: interactions between the giant kelp and the other algae species are elevated under nutrient- or temperature-stressed conditions.
+Increased wave height (physical disturbance) seems to enhance the negative effect of *Pterygophora* on *Macrocystis* adults.
+    + The effect of purple urchin on both *Macrocystis* adult and juvenile density is stronger than the red urchin effect, and more variable with changing conditions.
+Purple urchin herbivory as a negative influence on *Macrocystis* is enhanced with elevated disturbance, MEI, and PDO, and slightly reduced with an increase in the NPGO.
+    + The strenghts of intraspecies effects (effects of *Macrocystis* adults on juveniles and vice versa) are also influenced by physical conditions, but perhaps in a different way than might be expected.
+The positive effect of juvenile *Macrocystis* on adult *Macrocystis* seems to be **enhanced** under nutrient-stressed conditions and higher temperatures, and reduced with greater nutrient availability and increased disturbance.
+
+### Summarized effects of different physical conditions
+
+* **Large-scale oceanographic oscillations: NPGO, MEI, and PDO**
+    + Higher nutrient availability often associated with higher values of the NPGO, and lower values of the MEI and PDO.
+    + High nutrient availability leads to higher urchin growth rates and reduced competition between urchin species.
+*Laminaria* and *Pterygophora* intraspecific effects are enhanced, while the *Macrocystis* intraspecies effect is reduced.
+The negative effect of herbivory is strengthened for *Pterygophora*, but reduced for *Macrocystis*.
+The ability of *Pterygophora* and *Laminaria* to compete with *Macrocystis* is reduced under greater nutrient availability, as evidenced by a decrease in interaction strength between the two understory species and the giant kelp, as well as an increase in the positive association of the two with each other, likely as a mutual defense against *Macrocystis* dominance.
+    + The opposite of the above effects are true for nutrient-scarce conditions.
+High values of the MEI and PDO, and low values of the NPGO, are associated with reduced urchin growth and marginally increased urchin competition.
+The negative effect of herbivory is reduced for *Pterygophora*, but exacerbated for *Macrocystis*.
+The understory species are favored in algal competition, in that their negative effects on *Macrocystis* are strengthened, and the converse effects of *Macrocystis* on the understory are reduced.
+Interestingly, *Pterygophora* seems beneficial for *Macrocystis* recruitment under nutrient-stressed conditions, while *Laminaria* is inhibitory.
+The *Laminaria* intraspecies effect is reduced, while the intraspecies *Macrocystis* effect is enhanced.
+    + The qualitative story to be told here is that under nutrient-poor conditions, a *Pterygophora* state is favored in a relative sense over a *Macrocystis* dominant state.
+Under nutrient-stressed conditions, *Pterygophora* is more resistant to herbivory, and more inhibitory to adult *Macrocystis*.
+However, *Pterygophora* is beneficial for *Macrocystis* recruitment, while *Laminaria* inhibits *Macrocystis* recruitment under nutrient stressed conditions.
+When conditions are "better" (more nutrients), *Macrocystis* gains.
+It becomes more able to withstand herbivory, less sensitive to *Pterygophora* or *Laminaria* presence, and more inhibitory to *Pterygophora*.
+
+* **Temperature**
+    + Elevated SST has similar effects to nutrient stress (elevated MEI and PDO, and reduced NPGO), which makes sense: In Southern California, elevated SST is associated with reduced nutrient availability[@Zimmerman1984,@Bell2015].
+Similar effects as described above for nutrient stress hold for elevated temperature.
+
+* **Physical disturbance**
+    + Higher max wave heights was our proxy for physical stress/disturbance.
+    + Higher physical disturbance levels results in reduced interaction strength between the urchin species.
+    + *Pterygophora*'s competitive ability is enhanced with increased disturbance[@Dayton1999], evidenced by an increase in the negative effect of *Pterygophora* on adult *Macrocystis*.
+    + *Pterygophora* and *Macrocystis* are both subject to a greater negative effect of herbivory under increased disturbance.
+    + *Laminaria* benefits as well. The negative effects of herbivory and algal competition are reduced with increased disturbance, and its intraspecies effect is strengthened.
+    + The positive effect of juvenile on adult *Macrocystis* is reduced with increased physical stress.
+
 ## References

### 2018-05-31 data/data preparation scripts/physical_data_preparation.R

@@ -144,7 +144,7 @@ source("data/data preparation scripts/wave_height_Bell.R")
 
 # Join the two datasets together, using the normalized buoy data to "fill in" the modeled data
 wavesboth <- wavesdf %>%
-  mutate(waves.norm=(maxHs-mean(maxHs,na.rm=T)/sd(maxHs,na.rm=T)))%>%
+  mutate(waves.norm=(maxHs-mean(maxHs,na.rm=T))/sd(maxHs,na.rm=T))%>%
   # Use the periods key to assign periods to each month, and remove irrelevant data
   left_join(period.key,by=c("year"="year","month"="month")) %>%
   filter(!is.na(period)) %>%

### 2018-05-31 data/data preparation scripts/wave_height_Bell.R

@@ -3,15 +3,13 @@
 require(tidyverse)
 require(lubridate)
 
-hs <- read_csv("data/raw/SanNic_Hs.csv") %>%
+hs <- read_csv("data/raw/SanNic_Hs_v2.csv",col_types = "iiiidd") %>%
   # Site 2 is the West End of SNI
-  select(Year,Month,Day,Max_2) %>%
+  select(Year,Month,Day,Mean_2,Max_2) %>%
   unite("date",Year,Month,Day,sep="-",remove=F)%>%
   mutate(date=as.Date(date)) %>%
   rename(maxHs=Max_2)%>%
-  # Monthly mean of daily Max Hs
-  # group_by(Year,Month) %>%
-  # summarise(maxHs=mean(Max_2,na.rm=T)) %>%
+
   rename(year=Year,month=Month,day=Day)%>%
   ungroup()
 hs$maxHs[is.na(hs$maxHs)]<-NA
@@ -41,62 +39,61 @@ heights <- ncvar_get(modwaves,"sea_surface_wave_significant_height")
 
 # Maximum wave height defined as 
 # average height of the one third highest waves in the record over the time period
-# function takes a vector of wave heights, calculates the 2/3rds quantile, and then
-# averages the observations greater than that cutoff
-calc_maxHs <- function(Hs) {
-  cutoff <- quantile(Hs,2/3)
-  mean(Hs[Hs>cutoff])
-}
+
 # Convert to data frame
 wavesdf <- data_frame(time=times,Hs=heights,date=as_date(as.POSIXct(time,origin="1970-01-01 00:00:00"))) %>%
-  #Calculate montly means
   mutate(year=year(date),month=month(date),day=day(date)) %>%
   
-  # Calculate the average height of one third highest waves for each month
+  # Calculate the max Hs for each day
   group_by(year,month,day) %>%
   summarise(maxHs_usgs=max(Hs)) %>%
   ungroup()
-  # group_by(year,month) %>%
-  # summarise(maxHs_usgs=mean(maxHs,na.rm=T)) %>%
-  # ungroup()
-
 
 # Remove unneeded data from workspace
-rm(calc_maxHs,modwaves,times,heights)
+rm(modwaves,times,heights)
 
-### FINISH ###
 
-# compare #
+# join and compare 2 datasets #
 wavesdf <- hs %>%
[…]

### 2018-05-31 EDM_SNI_analysis.Rmd
cleaned up simplex and smap; added consistent RNGseed to CCM

@@
 require(extrafont)
+
+# plot theme
+plot_theme <-   theme_minimal()+
+  theme(text=element_text(family="Rockwell",size=12,color="black"),
+        legend.text = element_text(size=14),
+        axis.title=element_text(family="Rockwell",size=14,color="black"),
+        axis.text=element_text(family="Rockwell",size=8,color="black"),
+        strip.background = element_rect(colour="black"),
+        panel.border = element_rect(color="black",fill=NA))
+theme_set(plot_theme)
 ```
@@
       guides(linetype="none")+
-      theme_minimal()+
-      theme(text = element_text(family="Rockwell",size=16),
+      theme(text = element_text(size=16),
             legend.position = c(0.15,0.8),
@@
 
-#ggplot theme to apply to plots later
-theme_rockwell <- function() {
-  theme_minimal()+
-  theme(text=element_text(family="Rockwell",size=12),
-        axis.title = element_text(family="Rockwell",size=14),
-        strip.background = element_rect(colour="black"),
-        panel.border = element_rect(color="black",fill=NA))
-}
 ```
@@
 
-```{r univariate simplex westend,fig.height=10,fig.width=7,echo=F}
+```{r univariate simplex westend,fig.width=7,echo=F}
 # List to store output of simplex projection
@@
 # Run simplex projection for each species at each site, and plot output
-par(mfrow=c(3,2))
 for(i in 1:length(study_spp)) {
@@
   dat <- westend %>% select(matches(spp)) %>% as.data.frame()
-  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=2:15,silent=T)
+  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=2:15,silent=T) %>%
+    mutate(spp=namekey$long[match(spp,namekey$short)])
   westend.simp.list[[spp]] <- out
-
-  # Plot embedding dimension vs. predictive ability (rho)
-  plot(out$E,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],
-       ylim=c(0,1),
-       xlab="Embedding Dimension (E)",ylab=expression(paste("Skill, ",rho)))
-}
+}
+
+# plot
+bind_rows(westend.simp.list) %>%
+  ggplot(aes(E,rho,color=spp))+
+  geom_line(size=2)+
+  facet_wrap(~spp,nrow=2,scales="free_y")+
+  labs(x="Embedding Dimension (E)",y=expression(paste("Skill, ",rho)))+
+  scale_x_continuous(breaks=seq(0,12,by=2))+
+  scale_color_manual(values=c('Macrocystis >1m'="navyblue",'Macrocystis <1m'="gray50",'Pterygophora'="darkgreen",'Laminaria'="darkcyan",'Purple urchin'="mediumpurple4",'Red urchin'="darkred"),name="")+
+  guides(color=F)
 
@@
   dat <- westend %>% select(matches(ind)) %>% as.data.frame()
-  out <- simplex(as.numeric(dat[,1]),lib=c(1,69),E=2:10,silent=T)
+  out <- simplex(as.numeric(dat[,1]),lib=c(1,69),E=2:10,silent=T) %>%
+    mutate(spp=namekey$long[match(ind,namekey$short)])
   phys.simp.list[[ind]] <- out
   westend.simp.list[[ind]] <- out
-
-  # Plot embedding dimension vs. predictive ability (rho)
-  plot(out$E,out$rho,type="l",main=namekey$long[match(ind,namekey$short)],xlab="Embedding Dimension (E)",ylab=expression(paste("Skill, ",rho)))
-}
+}
+
+# plot
+bind_rows(phys.simp.list) %>%
+  ggplot(aes(E,rho))+
+  geom_line(size=2)+
+  facet_wrap(~spp,nrow=2,scales="free_y")+
+  labs(x="Embedding Dimension (E)",y=expression(paste("Skill, ",rho)))+
+  scale_x_continuous(breaks=seq(0,10,by=2))+
+  guides(color=F)+
+  theme(strip.text = element_text(size=8))
+
 rm(ind,dat,out)
@@
 
-```{r prediction horizon test,message=F,echo=F,fig.height=10,fig.width=7,echo=F}
+```{r prediction horizon test,message=F,echo=F,fig.width=7,echo=F}
 # List to hold prediction horizon results
@@
   dat <- westend %>% select(matches(spp)) %>% as.data.frame()
-  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),silent=T,E=westend.bestE[spp],tp=1:10)
+  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),silent=T,E=westend.bestE[spp],tp=1:10) %>%
+    mutate(spp=namekey$long[match(spp,namekey$short)])
   westend.tp.list[[spp]] <- out
-
-  # Plot time horizon vs. predictive ability (rho)
-  plot(out$tp,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],xlab="Time to Prediction",ylab=expression(paste("Skill, ",rho)))
-}
+}
+
+#Plot time horizon vs. rho
+bind_rows(westend.tp.list) %>%
+  ggplot(aes(tp,rho,color=spp))+
+  geom_line(size=2)+
+  facet_wrap(~spp,nrow=2,scales="free_y")+
+  labs(x="Time to Prediction",y=expression(paste("Skill, ",rho)))+
+  scale_color_manual(values=c('Macrocystis >1m'="navyblue",'Macrocystis <1m'="gray50",'Pterygophora'="darkgreen",'Laminaria'="darkcyan",'Purple urchin'="mediumpurple4",'Red urchin'="darkred"),name="")+
+  scale_x_continuous(breaks=seq(0,10,by=2))+
+  guides(color=F)
+
 rm(spp,dat,out)
@@
 The prediction horizon effect check is promising overall, as most variables decline in predictive ability with increasing time horizon.
-For *Pterygophora californica*, there is some evidence of cyclic behavior, in that dynamics are more predictable 6 periods in the future than 3 or 4 periods.
-This is something to be mindful of, but we leave *Pterygophora* in our analysis for now.
+For *Pterygophora californica*, *Macrocystis* juveniles, and *Laminaria*, there is some evidence of cyclic behavior, in that dynamics are more predictable 6 periods in the future than 3 or 4 periods.
 
@@
 
-```{r univariate s_maps westend,fig.height=10,fig.width=7,echo=F}
+```{r univariate s_maps westend,fig.width=7,echo=F}
 westend.smap.list <- list() # list to store output
@@
   dat <- westend %>% select(matches(spp)) %>% as.data.frame()
-  out <- s_map(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=westend.bestE[spp],silent=T)
+  out <- s_map(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=westend.bestE[spp],silent=T) %>%
+    mutate(spp=namekey$long[match(spp,namekey$short)])
   westend.smap.list[[spp]] <- out
-
-  # Plot theta dimension vs. predictive ability (rho)
-  plot(out$theta,out$rho,type="l",main=namekey$long[match(spp,namekey$short)],xlab=expression(paste("Nonlinearity (",theta,")")), ylab=expression(paste("Skill, ",rho)))
-}
+}
+
+bind_rows(westend.smap.list) %>%
+  ggplot(aes(theta,rho,color=spp))+
+  geom_line(size=2)+
+  facet_wrap(~spp,nrow=2,scales="free_y")+
+  labs(x=expression(paste("Nonlinearity (",theta,")")),y=expression(paste("Skill, ",rho)))+
+  scale_color_manual(values=c('Macrocystis >1m'="navyblue",'Macrocystis <1m'="gray50",'Pterygophora'="darkgreen",'Laminaria'="darkcyan",'Purple urchin'="mediumpurple4",'Red urchin'="darkred"),name="")+
+  scale_x_continuous(breaks=seq(0,8,by=2))+
+  guides(color=F)
+
 rm(spp,dat,out)
@@
       tempE=westend.bestE[col_names[j]]
-      temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 2+j,lib_sizes = c(10,500),num_samples=100,replace=T,silent=T)
+      temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 2+j,lib_sizes = c(10,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
       # mean rho at library size 500
@@
   labs(y="Estimated Interaction Coefficient",x="Interacting Species")+
-  facet_grid(mod_plotting_name~.,scales="free",space="free")+
-  theme_rockwell()
+  facet_grid(mod_plotting_name~.,scales="free",space="free")
 
@@
   geom_text(aes(x=1.5,y=2,label=label),hjust=1,family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
-  theme_rockwell()+
   theme(strip.background = element_blank(),
@@
   labs(y="Estimated Interaction Coefficient",x="Interacting Species")+
-  facet_grid(mod_plotting_name~.,scales="free_y",space="free")+
-  theme_rockwell()
+  facet_grid(mod_plotting_name~.,scales="free_y",space="free")
 
@@
     labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+
-    facet_grid(type~external_forcing,scales = "free_y")+
-    theme_rockwell()
+    facet_grid(type~external_forcing,scales = "free_y")
 })
@@
 * *Laminaria farlowii*
-    + as far as competition goes, most large, negative interactions are associated with juvenile *Macrocystis*, likely as an indicator of impending succesional processes (i.e., the juvenile *Macrocystis* growing into large *Macrocystis*).
+    + Important to note that no physical variables in this analysis showed a causal signal for *Laminaria* from CCM
+    + as far as competition goes, most large, negative interactions are associated with juvenile *Macrocystis*, likely as an indicator of impending successional processes (i.e., the juvenile *Macrocystis* growing into large *Macrocystis*).
 This expected negative effect is reduced under elevated PDO, disturbance, and SST, which makes sense: these are conditions that classically inhibit *Macrocystis* dominance[@Reed2011; @Cavanaugh2011; @Bell2015,@Young2015].
@@
 * *Pterygophora californica*
-    + Algal effects: *Laminaria* is a good indicator of a positive density response of *Pterygophora*, while, as mentioned above (and surprisingly) the effect of *Macrocystis* is small.
-The positive effect of *Laminaria* increases with slightly with elevated NPGO and disturbance, and decreases with elevated MEI, PDO, and SST.
+    + Algal effects: *Macrocystis* has a near-zero mean effect on *Pterygophora*.
+However, the quantile regression suggests that *Macrocystis* can have a strong negative effect on *Pterygophora* at lower relative levels of nutrient and temperature stress (high NPGO, low MEI and PDO, lower SST), and potentially after disturbance.
     + Herbivory effects on *Pterygophora* vary with physical environment.
@@
 franciscanus* and *S.
-purpuratus* have similar mean effects on *Pterygophora*, but the effect of *M.
-franciscanus* is much more sensitive to physical conditions.
-Its negative effect is strongest under elevated disturbance and NPGO, while its effect is diminished under elevated MEI, PDO, and SST.
-The same trends hold for the effect of *S.
-purpuratus* on *Pterygophora*, but those effects are less sensitive.
-    + The variation in the intraspecies effect of *Pterygophora* is responsive to physical conditions in a similar manner to *Laminaria* (reduced under elevated MEI, PDO, and SST, and increased under elevated NPGO and disturbance)
+purpuratus* have similar mean effects on *Pterygophora*, and that herbivory is sensitive to physical conditions.
+The effect of herbivory is strongest under elevated disturbance and NPGO, and diminished under elevated MEI, PDO, and SST.
+    + Little variation in the intraspecies effect of *Pterygophora* along physical gradients.
 

### 2018-06-01 EDM_SNI_analysis.Rmd
fixed figure width and network graph

@@
 require(knitr)
-knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message = FALSE)
+knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message = FALSE,results = 'asis')
 ```
@@
 
-```{r ccm all species and physical vars, echo=F,warning=F}
+```{r ccm all species and physical vars, echo=F,warning=F,fig.width=7}
 # Rows hold predicted variables, columns hold forcing variables. There are more columns than rows since the physical variables are included as potential forcing factors.
@@
 
-```{r igraph, echo=F, message=F,warning=F}
+```{r igraph, echo=F, message=F,warning=F,fig.height=6}
 require(igraph) # R Package for network graphing and analysis
@@
 #vertex coordinates
-crds <- cbind(c(300,240,375,313,158,122,430,203,146.3342,309,20),c(10,126,102,206,204,85,162.8175,0,314.3077,318,232.6699))
+crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))
 
@@
 
-plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.size=40,edge.lty=2,edge.color="black",vertex.color=vert.cols,layout=crds,vertex.frame.color=F,main="",edge.curved=F)
+plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.shape="circle", edge.curved=F,margin=0, vertex.size=30,rescale=T,edge.lty=1,edge.color="gray40",layout=crds,vertex.color=vert.cols,vertex.frame.color=NA,main="")
 ```
@@
 The maximum number of outgoing interactions is 5 (6 total biological variables minus the variable itself), and incoming is 10 (5 biological and 5 physical variables).
-Algae species are in brown, while the two urchin species are in purple and the physical driver are in grey.
-For example, as mentioned above, we can see that young *Macrocystis* is forced by all other variables in the system, but only itself shows a causal forcing on 3 of the 5 other biological variables.
-Again we can see that while the web is not entirely saturated, there are many unidirectional forcings, and a lot of bidirectional causation as well (A forces B *and* B forces A).
+Algae species are in brown, while the two urchin species are in purple and the physical drivers are in grey.
+For example, as mentioned above, we can see that young *Macrocystis* (denoted M.
+pyr (j))is forced by all other variables in the system, but only itself shows a causal forcing on 4 of the 5 other biological variables.
+Again we can see that while the web is not entirely saturated, there are many unidirectional forcings, and a prevalence of bidirectional causation as well (A forces B *and* B forces A).
+The adult *Macrocystis* is causally connected to all variables in the analysis.
 
@@
 
-```{r individual_effects,fig.height=6}
+```{r individual_effects,fig.width=7}
 
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-    geom_point(col='gray30',alpha=0.7)+
+    geom_point(col='gray20',alpha=0.6)+
     geom_hline(yintercept = 0,linetype=2)+
-    geom_quantile(quantiles=quants,lineend = "round",size=1.5,formula=y~x)+
+    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
     scale_color_manual(values=c('mac'="navyblue",'ymac'="gray50",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+

### 2018-06-01 EDM_SNI_analysis.Rmd
adding reconstructed attractor visual

@@
             legend.position = c(0.15,0.8),
-            legend.text = element_text(size=12))
+            legend.text = element_text(size=12),
+            panel.border = element_blank())
 westend.mean_dens_plot
@@
 # Run simplex for each physical variable
-par(mfrow=c(3,2))
 phys.simp.list <- list()
@@
   dat <- westend %>% select(matches(ind)) %>% as.data.frame()
-  out <- simplex(as.numeric(dat[,1]),lib=c(1,69),E=2:10,silent=T) %>%
+  out <- simplex(as.numeric(dat[,1]),lib=c(1,63),E=2:10,silent=T) %>%
     mutate(spp=namekey$long[match(ind,namekey$short)])
@@
 ## 5. Results
+
+### Multivariate Attractors
+
+The multivariate models reconstruct attractors for each species based on their causally linked variables from CCM. A visualization of those models in state space might look something like this:
+
+```{r 3d plot,fig.height=6}
+require(plot3D)
+
+x <- westend$lam
+y <- westend$mac
+z <- westend$purp
+
+seg.start <- as.numeric(westend.segs[1,1])
+seg.end <- as.numeric(westend.segs[1,2])
+
+#smooth
+t <- 1:length(seg.start:seg.end)
+tt<-seq(1,length(seg.start:seg.end),len=500)
+
+xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
+ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
+zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
+
+xtemp <- x[seg.start:seg.end]
+ytemp <- y[seg.start:seg.end]
+ztemp <- z[seg.start:seg.end]
+
+# plot
+
+scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
+          type="l",phi=40,theta=20,bty="b",xlab = "Laminaria", ylab = "Macrocystis", zlab = "Purple Urchin")
+points3D(xsmooth,ysmooth,zsmooth,col="black",pch=19,cex=0.5,add=T)
+
+for(i in 2:5) {
+  seg.start <- as.numeric(westend.segs[i,1])
+  seg.end <- as.numeric(westend.segs[i,2])
+
+  #smooth
+  t <- 1:length(seg.start:seg.end)
+  tt<-seq(1,length(seg.start:seg.end),len=500)
+  xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
+  ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
+  zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
+
+  xtemp <- x[seg.start:seg.end]
+  ytemp <- y[seg.start:seg.end]
+  ztemp <- z[seg.start:seg.end]
+
+  # plot
+
+  scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
+            type="l",phi=40,bty="b",add=T)
+  points3D(xtemp,ytemp,ztemp,col="black",pch=19,cex=0.5,add=T)
+}
+
+scatter3D(x[46],y[46],z[46],col="red",pch=19,cex=2,add=T)
+
+#manually find nearest neighbors to point 46
+require(RANN)
+temp <- westend %>% select(lam,mac,purp) %>% slice(1:315) %>% mutate(id=row_number()) %>%
+  filter(!is.na(lam))
+tempnn <-nn2(temp[,1:3],k=4)$nn.idx %>% as.data.frame()
+names(tempnn) <- c("nn1","nn2","nn3","nn4")
+temp <- bind_cols(temp,tempnn)
+
+scatter3D(x[109],y[109],z[109],col="darkgreen",pch=19,cex=2,add=T)
+scatter3D(x[235],y[235],z[235],col="darkgreen",pch=19,cex=2,add=T)
+scatter3D(x[169],y[169],z[169],col="darkgreen",pch=19,cex=2,add=T)
+
+scatter3D(x[110],y[110],z[110],col="blue",pch=19,cex=2,add=T)
+scatter3D(x[236],y[236],z[236],col="blue",pch=19,cex=2,add=T)
+scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
+
+rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)
+```
+
+***
 
@@
 
+***
+
 ### Effects on Individual Species
 
-With 10%/90% quantile regression.
-Columns/x-axis indicates the level of a physical stressor, while rows and y-axis denote the type of species interaction (corresponding to the colors in the boxplots above).
+With 5%/95% quantile regression.
+Columns/x-axis indicates the normalized level of a physical stressor, while rows and y-axis denote the type of species interaction (corresponding to the colors in the boxplots above).
 
@@
 phys_effects_on_interactions <- purrr::map(as.list(study_spp),function(spp) {
-  quants <- c(0.10,0.5,0.90)
+  quants <- c(0.05,0.5,0.95)
   plot_title <- namekey$long[match(spp,namekey$short)]
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-    geom_point(col='gray20',alpha=0.6)+
+    geom_point(col='gray60',alpha=0.6)+
     geom_hline(yintercept = 0,linetype=2)+
     geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-    scale_color_manual(values=c('mac'="navyblue",'ymac'="gray50",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+
+    scale_color_manual(values=c('mac'="navyblue",'ymac'="gray30",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+
     labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+

### 2018-06-07 EDM_SNI_analysis.Rmd
cleaned up language throughout

@@
 author: "Owen Liu"
-date: "March 13, 2018"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: pdf_document
@@
 In the analysis that follows, in each step that involves a new type of EDM method, its logic will be briefly described.
-However, keep in mind that the basic theory behind each individual method is the same: We use some dynamic information (from single or multiple time series) to reconstruct a shadow attractor, and then use the properties of that manifold to make predictions.
+However, keep in mind that the basic theory behind each individual method is the same: We use some dynamic information (from single or multiple time series) to reconstruct a shadow attractor/manifold, and then use the properties of that manifold to make predictions or estimate interactions.
 
@@
 
-And in 2016, the data were again compiled and an updated version was published online:
-
-Miller R., A. Rassweiler, D. Reed, K. Lafferty, L. Kui, M. O'Brien. 2016. Santa Barbara Channel Marine BON: Integrated quad and swath cover. Environmental Data Initiative.
-
 ### Physical Oceanographic Data
@@
 * **Maximum significant wave height (Hs)**
-    + Combined data from the Begg and SNI buoys with modeled data from the [Geophysical Fluid Dynamics Laboratory](http://cmgwindwave.usgsportals.net/)
+    + Modeled wave height data from from the [Geophysical Fluid Dynamics Laboratory](http://cmgwindwave.usgsportals.net/) and from Tom Bell at UCSB
     + Signficant wave height is defined as the average height, in meters, of the one third highest waves in the record
@@
 
-Unlike the biological data, where there are unique spatial replicates, the physical data have only one value for each of the 69 monitoring periods, and hence their values are replicated (copied) for each site to match the total length of the biological data.
+Unlike the biological data, where there are unique spatial replicates, the physical data have only one value for each of the 63 monitoring periods, and hence their values are replicated (copied) for each site to match the total length of the biological data.
 
@@
 To avoid in-sample fitting, we use a leave-one-out cross-validation scheme, removing one vector at a time from the library, and predicting its dynamics from the other library vectors.
+From this exercise, we save the best embedding dimension *E* for each variable, to use in following steps.
 
@@
 Therefore, if model skill $\it{\bf{\rho}}$ increases with increasing $\it{\bf{\theta}}$, it is evidence of nonlinear, state-dependent dynamics.
+
 For a more formal description of the S-maps procedure, see @Sugihara1994 and @Deyle2016.
@@
 Conceptually, as the dynamic system moves along the surface of the attractor, S-maps sequentially computes linear maps to the next point based on nearby points.
-
-For prediction of each variable using S-maps, we can again use the optimal embedding dimension, *E*, found through our simplex projection above, and we plot the tuning parameter $\it{\bf{\theta}}$ against $\it{\bf{\rho}}$ to look for evidence of nonlinear dynamics.
+Mathematically, when making a prediction for a target point $\it{\bf{x^*}}$, each library vector (point on the attractor) $\it{\bf{x}_{k}}$ is given a weight
+$$w_{k}=exp\frac{-\theta\mid\mid \it{\bf{x}_{k}}- \it{\bf{x^*}}\mid\mid}{\bar{d}}$$ 
+where $\mid\mid \it{\bf{x}_{k}}- \it{\bf{x^*}}\mid\mid$ is the Euclidean distance between the library and target vector, and $\bar{d}$ is the average distance to all library vectors.
+This weighting is what allows the tuning of nonlinearity by changing the single value $\theta$.
+
+For prediction of each variable using S-maps, we build an attractor for each species using the optimal embedding dimension, *E*, found through our simplex projection above, and we plot the tuning parameter $\it{\bf{\theta}}$ against $\it{\bf{\rho}}$ to investigate whether local, state-dependent weighting of library vectors increases predicability, which would be evidence of nonlinear dynamics.
 
@@
 
-All variables show significantly improved predictive ability with increased theta, suggesting nonlinear dynamics.
-Together, the simplex, prediction horizon, and S-map results suggest our approach is valid--variables are predictable, and most predictable in a nonlinear manner.
+All variables show significantly improved predictive ability with increased $\theta$, suggesting nonlinear dynamics.
+Together, the simplex, prediction horizon, and S-map results suggest our approach is valid--variables are predictable and nonlinear.
 
@@
 
+So far, we have established the predictability and nonlinearity of single variables.
+In this step, we relate the variables to each other, asking whether there is evidence of causal linkages.
+We do not assume causal interactions--we test for them.
+
 Generalizations of Takens' theorem indicate that if two variables (in our case, species or physical variables) are part of the same dynamic system, their individual dynamics should reflect their relative causal influence [@Sugihara2012; @Deyle2013,@Ye2015,@Clark2015].
-In other words, if one variable (for example, giant kelp), is casually forced by another (sea urchins), that forcing should leave a signature on the giant kelp time series.
+In other words, if one variable (for example, giant kelp), is causally forced by another (sea urchins), that forcing should leave a signature on the giant kelp time series.
 Convergent cross mapping (CCM) tests for causation by using the attractor/manifold built from the time series of one variable to predict another.
@@
 If the attractor can accurately (based on out-of-sample prediction skill, just as before) predict the dynamics of the second variable, we can claim that the second variable has a causal influence on the first.
-In other words, the *causal effect of A on B is determined by how well B cross-maps A*.
+In simple terms, the *causal effect of A on B is determined by how well B cross-maps A*.
 In this way, the inference from cross-mapping is the converse direction of causation.
@@
 There are two criteria for CCM to establish causality:
+
 * First, and most obviously, predictive cross-map skill using all available data should be significantly greater than zero. 
@@
 Convergence means that cross-mapped estimates improve with library length, because the attractor is more fully resolved and therefore estimation error should decline.
-Convergence is key to distinguishing causation from simple correlation.
-
-The CCM algorithm uses a random sampling method to test multiple "versions" of each library size, sampling a subset from the supplied library vectors to give a sense of the confidence intervals around prediction skill.
+Convergence is key to distinguishing causation from simple or spurious correlation [@Sugihara2012].
+If two variables are spuriously correlated and not causally linked, CCM should fail to satisfy this second criterion.
+
+It is also generally true that relative cross-mapping skill of a given variable equates to the strength of causal linkages.
+That is, if species A cross-maps species B with greater skill than it cross-maps species C, it would suggest that species B is the stronger forcing influence on A.
+
+The CCM algorithm uses a random sampling method to test multiple "versions" of each library size, sampling a subset from the supplied library vectors to establish confidence intervals around prediction skill.
 We again use leave-one-out cross-validation to prevent in-sample fitting.
-
-We can use CCM theory to ensure that the species in our data are actually displaying causal signals.
 
@@
   scale_fill_gradient(low = "#9AFF9A", high = "#548B54", space = "Lab", na.value = "grey50", 
-      guide = "colourbar",name=expression(paste(rho, "(skill)"))) +
+      guide = "colourbar",limits=c(0,0.8),breaks=c(0.2,0.4,0.6,0.8),name=expression(paste(rho, "(skill)"))) +
   ggtitle("Kelp Forest Convergent Cross Mapping, West End") +
@@
 
-The plot shows cross-mapped (predictor, or forcing) variables in the columns, while rows are predicted variables.
-Gray indicates a lack of significant causal signal, and darker colors represent higher cross-map skill.
+The plot shows cross-mapped (predictor, or forcing) variables in the columns, while rows are predicted variables (i.e., columns cause rows).
+Gray indicates a lack of significant causal signal--failure to satisfy one of the above criteria--and darker colors represent higher cross-map skill.
 
@@
 The five physical forcing variables are represented in the first five columns, followed by the algae and urchin species.
-Many physical variables show causal relationships with kelp forest species, especially the North Pacific Gyre Oscillation and Pacific Decadal Oscillation, which show causal links to almost all the biological variables.
-The NPGO is cross-mapped well by the purple urchin *Strongylocentrotus purpuratus*, while SST in turn shows a strong influence on adult *Macrocystis* dynamics.
+Many physical variables show causal relationships with kelp forest species, especially the North Pacific Gyre Oscillation and maximum wave height (physical disturbance), which show causal links to almost all the biological variables.
+The NPGO is cross-mapped well by the purple urchin *Strongylocentrotus purpuratus*, while both the NPGO and wave height show a strong influence on adult *Macrocystis* dynamics.
 So already we are seeing effects that we might expect based on known interactions between physical variables and biological dynamics [@Reed2011;@Bell2015].
 
-Additionally, adult *Macrocystis* itself seems to be a strong driver of the dynamics of many, if not all the other species in the tropic web, supporting the decades of research showing the importance of giant kelp as a foundational species [@Graham2007c;@Dayton1985], but further suggesting that its dynamics fundamentally drive the dynamics of other species.
-Juvenile, or sporophyte *Macrocystis*, on the other hand, is affected by most of the variables in the system, but does not itself drive dynamics of other species, again an effect we would expect to find.
-In this system, *Pterygophora californica* also seems to be a signicant driver of the dynamics of other species in the trophic web.
-
-Both sea urchin species are causal predictors of *Macrocystis* density, and they also have a causal effect on one another.
+Additionally, adult *Macrocystis* itself seems to be the first or second most important driver of the dynamics of all the other species in the tropic web, supporting the decades of research showing the importance of giant kelp as a foundational species [@Graham2007c;@Dayton1985], but further suggesting that its dynamics fundamentally drive the dynamics of other species.
+Juvenile, or sporophyte *Macrocystis*, on the other hand, is affected by most of the variables in the system, but does not itself strongly drive dynamics of other species, again an effect we would expect to find.
+In this system, *Pterygophora californica* also seems to be a strong causal driver of the dynamics of other species in the trophic web.
+
+Both sea urchin species are causally linked to *Macrocystis* density, and they also have a causal effect on one another.
 This analysis alone, however, does not yet tell us whether the bi-directional interaction between *S.
@@
 
+### Interaction Network
+
+The CCM results translated into an empirical interaction web looks like this:
+
+```{r igraph, echo=F, message=F,warning=F,fig.height=6}
+require(igraph) # R Package for network graphing and analysis
+
+# Add NAs for connections in to physical variables (since we want them just to be exogenous)
+connect.mat <- t(westend.ptot.mat) %>% 
+  as.data.frame() %>%
+  mutate(mei=rep(NA,nrow(.)),pdo=rep(NA,nrow(.)),npgo=rep(NA,nrow(.)),waves=rep(NA,nrow(.)),sst=rep(NA,nrow(.)))
+
+g <- graph.adjacency(as.matrix(connect.mat))
+par(mfrow=c(1,1))
+
+# Variable names
+
+names <- as.character(sapply(V(g)$name, function(x) namekey$plotting[match(x,namekey$short)]))
+names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
+
+# Vertex labels
+V(g)$name <-names
+vert_in <- sapply(degree(g,mode="in"),function(x) paste0("(",x,")"))
+vert_out_in <- paste0(degree(g,mode="out"),vert_in)
+V(g)$label <- paste(sep="\n",V(g)$name, vert_out_in)
+V(g)$label[7:11] <- paste(V(g)$name[7:11])
+
+#vertex coordinates
+crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))
+
+# vector of node colors for plotting
+vert.cols <- c(rep("tan3",3),rep("orchid3",2),"tan3",rep("gray70",5))
+
+plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.shape="circle", edge.curved=F,margin=0, vertex.size=30,rescale=T,edge.lty=1,edge.color="gray40",layout=crds,vertex.color=vert.cols,vertex.frame.color=NA,main="")
+```
+
+In this network, each node represents a variable, and the numbers associated with each are the number of causal connections in the web, outgoing and (incoming).
+The maximum number of outgoing interactions is 5 (6 total biological variables minus the variable itself), and incoming is 10 (5 biological and 5 physical variables).
+Algae species are in brown, while the two urchin species are in purple and the physical drivers are in grey.
+For example, as mentioned above, we can see that young *Macrocystis* (denoted M.
+pyr (j))is forced by all other variables in the system, but only itself shows a causal forcing on 4 of the 5 other biological variables.
+Again we can see that while the web is not entirely saturated, there are many unidirectional forcings, and a prevalence of bidirectional causation as well (A forces B *and* B forces A).
+The adult *Macrocystis* is causally connected to all variables in the analysis.
+
+***
+
 ## 4. Multivariate Models Using CCM results
@@
 Multivariate models have the same structure, except instead of using single variables to reconstruct the attractors, we use contemporaneous values of multiple variables [@Deyle2016].
-Instead of library vectors or points in state-space taking the form of, for example, $<x_t,x_{t-1},x_{t-2}>$, they now are formed in true multivariate space, e.g.
+That is, instead of library vectors or points in state-space taking the form of, for example, $<x_t,x_{t-1},x_{t-2}>$, they now are formed in true multivariate space, e.g.
 $<Kelp_{t},Urchin_{t},Nutrients_{t}>$.
-Using the normalized time series, we will build library vectors that each include the variable we are trying to predict, as well as other causal variables illuminated by our CCM analysis.
-Then we use S-maps as above to forecast the target variable using leave-one-out cross-validation just as in Step 3.
+Using the normalized time series, we will build library vectors that each include the variable we are trying to predict, as well as other causal variables illuminated by our CCM analysis from Step 3.
+Then we use S-maps as above to forecast the target variable using leave-one-out cross-validation.
 
@@
 
-### Interaction Network
-
-The CCM results translated into an empirical interaction web looks like this:
-
-```{r igraph, echo=F, message=F,warning=F,fig.height=6}
-require(igraph) # R Package for network graphing and analysis
-
-# Add NAs for connections in to physical variables (since we want them just to be exogenous)
-connect.mat <- t(westend.ptot.mat) %>% 
-  as.data.frame() %>%
-  mutate(mei=rep(NA,nrow(.)),pdo=rep(NA,nrow(.)),npgo=rep(NA,nrow(.)),waves=rep(NA,nrow(.)),sst=rep(NA,nrow(.)))
-
-g <- graph.adjacency(as.matrix(connect.mat))
-par(mfrow=c(1,1))
-
-# Variable names
-
-names <- as.character(sapply(V(g)$name, function(x) namekey$plotting[match(x,namekey$short)]))
-names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
-
-# Vertex labels
-V(g)$name <-names
-vert_in <- sapply(degree(g,mode="in"),function(x) paste0("(",x,")"))
-vert_out_in <- paste0(degree(g,mode="out"),vert_in)
-V(g)$label <- paste(sep="\n",V(g)$name, vert_out_in)
-V(g)$label[7:11] <- paste(V(g)$name[7:11])
-
-#vertex coordinates
-crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))
-
-# vector of node colors for plotting
-vert.cols <- c(rep("tan3",3),rep("orchid3",2),"tan3",rep("gray70",5))
-
-plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.shape="circle", edge.curved=F,margin=0, vertex.size=30,rescale=T,edge.lty=1,edge.color="gray40",layout=crds,vertex.color=vert.cols,vertex.frame.color=NA,main="")
-```
-
-In this network, each node represents a variable, and the numbers associated with each are the number of causal connections in the web, outgoing and (incoming).
-The maximum number of outgoing interactions is 5 (6 total biological variables minus the variable itself), and incoming is 10 (5 biological and 5 physical variables).
-Algae species are in brown, while the two urchin species are in purple and the physical drivers are in grey.
-For example, as mentioned above, we can see that young *Macrocystis* (denoted M.
-pyr (j))is forced by all other variables in the system, but only itself shows a causal forcing on 4 of the 5 other biological variables.
-Again we can see that while the web is not entirely saturated, there are many unidirectional forcings, and a prevalence of bidirectional causation as well (A forces B *and* B forces A).
-The adult *Macrocystis* is causally connected to all variables in the analysis.
+### Multivariate Attractors
+
+The multivariate models reconstruct attractors for each species based on their causally linked variables from CCM.
+A partial visualization of one of those attractors in state space might look something like this:
+
+```{r 3d plot,fig.height=7,fig.width=7}
+require(plot3D)
+
+x <- westend$lam
+y <- westend$mac
+z <- westend$purp
+
+seg.start <- as.numeric(westend.segs[1,1])
+seg.end <- as.numeric(westend.segs[1,2])
+
+#smooth
+t <- 1:length(seg.start:seg.end)
+tt<-seq(1,length(seg.start:seg.end),len=500)
+
+xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
+ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
+zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
+
+xtemp <- x[seg.start:seg.end]
+ytemp <- y[seg.start:seg.end]
+ztemp <- z[seg.start:seg.end]
+
+# plot
+op<-par()
+par(mai=c(0.1,0.1,0.1,0.1))
+scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
+          type="l",phi=40,theta=55,bty="b",xlab = "Laminaria", ylab = "Macrocystis", zlab = "Purple Urchin")
+points3D(xtemp,ytemp,ztemp,col="black",pch=19,cex=0.5,add=T)
+
+for(i in 2:5) {
+  seg.start <- as.numeric(westend.segs[i,1])
+  seg.end <- as.numeric(westend.segs[i,2])
+
+  #smooth
+  t <- 1:length(seg.start:seg.end)
+  tt<-seq(1,length(seg.start:seg.end),len=500)
+  xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
+  ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
+  zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
+
+  xtemp <- x[seg.start:seg.end]
+  ytemp <- y[seg.start:seg.end]
+  ztemp <- z[seg.start:seg.end]
+
+  # plot
+
+  scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
+            type="l",bty="b",add=T)
+  points3D(xtemp,ytemp,ztemp,col="black",pch=19,cex=0.5,add=T)
+}
+
+scatter3D(x[46],y[46],z[46],col="red",pch=19,cex=2,add=T)
+
+#manually find nearest neighbors to point 46
+require(RANN)
+temp <- westend %>% select(lam,mac,purp) %>% slice(1:315) %>% mutate(id=row_number()) %>%
+  filter(!is.na(lam))
+tempnn <-nn2(temp[,1:3],k=4)$nn.idx %>% as.data.frame()
+names(tempnn) <- c("nn1","nn2","nn3","nn4")
+temp <- bind_cols(temp,tempnn)
+
+scatter3D(x[109],y[109],z[109],col="darkgreen",pch=19,cex=2,add=T)
+scatter3D(x[235],y[235],z[235],col="darkgreen",pch=19,cex=2,add=T)
+scatter3D(x[169],y[169],z[169],col="darkgreen",pch=19,cex=2,add=T)
+
+scatter3D(x[110],y[110],z[110],col="blue",pch=19,cex=2,add=T)
+scatter3D(x[236],y[236],z[236],col="blue",pch=19,cex=2,add=T)
+scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
+
+par(op)
+rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)
+
+```
+
+Here we can see some apparent "states" of the *Laminaria*-*Macrocystis*-urchin dynamic system.
+There is some evidence here of three orthogonal states: when density of one of the species is well above its mean, the density of the other two are low.
+
+It is important to note, however, that this attractor is not entirely "unfolded" in just these three dimensions-- this is not quite a smooth multivariate attractor.
+This makes sense according to our previous CCM analysis.
+Although we cannot display more than three dimensions visually, our models in the previous step above suggest that any of these variables require more than three dimensions (embedding dimensions from Step 2) to make the most accurate predictions.
+
+However, as a further illustration of how predictions are made in EDM, consider the red dot in the figure above.
+If we are trying to predict the future trajectory for this ecosystem state (i.e., the combination, in this case, of the densities of the three species), we use multivariate projection.
+For simplex projection, we would find the closest neighbors in state-space to the point of interest (the green dots), and then our prediction is a weighted average of those neighbors projected forward one point in time (the blue dots).
+The average is weighted by the Euclidean distance of the green dots from the red dot.
+S-map forecasting works in a similar manner, except it uses all the points on the attractor (not just the nearest neighbors; see description of S-maps in Step 2).
 
@@
 
-### Multivariate Attractors
-
-The multivariate models reconstruct attractors for each species based on their causally linked variables from CCM. A visualization of those models in state space might look something like this:
-
-```{r 3d plot,fig.height=6}
-require(plot3D)
-
-x <- westend$lam
-y <- westend$mac
-z <- westend$purp
-
-seg.start <- as.numeric(westend.segs[1,1])
-seg.end <- as.numeric(westend.segs[1,2])
-
-#smooth
-t <- 1:length(seg.start:seg.end)
-tt<-seq(1,length(seg.start:seg.end),len=500)
-
-xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
-ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
-zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
-
-xtemp <- x[seg.start:seg.end]
-ytemp <- y[seg.start:seg.end]
-ztemp <- z[seg.start:seg.end]
-
-# plot
-
-scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
-          type="l",phi=40,theta=20,bty="b",xlab = "Laminaria", ylab = "Macrocystis", zlab = "Purple Urchin")
-points3D(xsmooth,ysmooth,zsmooth,col="black",pch=19,cex=0.5,add=T)
-
-for(i in 2:5) {
-  seg.start <- as.numeric(westend.segs[i,1])
-  seg.end <- as.numeric(westend.segs[i,2])
-
-  #smooth
-  t <- 1:length(seg.start:seg.end)
-  tt<-seq(1,length(seg.start:seg.end),len=500)
-  xsmooth<-splinefun(t, x[seg.start:seg.end])(tt)
-  ysmooth<-splinefun(t, y[seg.start:seg.end])(tt)
-  zsmooth<-splinefun(t, z[seg.start:seg.end])(tt)
-
-  xtemp <- x[seg.start:seg.end]
-  ytemp <- y[seg.start:seg.end]
-  ztemp <- z[seg.start:seg.end]
-
-  # plot
-
-  scatter3D(xsmooth,ysmooth,zsmooth, col="gray30",
-            type="l",phi=40,bty="b",add=T)
-  points3D(xtemp,ytemp,ztemp,col="black",pch=19,cex=0.5,add=T)
-}
-
-scatter3D(x[46],y[46],z[46],col="red",pch=19,cex=2,add=T)
-
-#manually find nearest neighbors to point 46
-require(RANN)
-temp <- westend %>% select(lam,mac,purp) %>% slice(1:315) %>% mutate(id=row_number()) %>%
-  filter(!is.na(lam))
-tempnn <-nn2(temp[,1:3],k=4)$nn.idx %>% as.data.frame()
-names(tempnn) <- c("nn1","nn2","nn3","nn4")
-temp <- bind_cols(temp,tempnn)
-
-scatter3D(x[109],y[109],z[109],col="darkgreen",pch=19,cex=2,add=T)
-scatter3D(x[235],y[235],z[235],col="darkgreen",pch=19,cex=2,add=T)
-scatter3D(x[169],y[169],z[169],col="darkgreen",pch=19,cex=2,add=T)
-
-scatter3D(x[110],y[110],z[110],col="blue",pch=19,cex=2,add=T)
-scatter3D(x[236],y[236],z[236],col="blue",pch=19,cex=2,add=T)
-scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
-
-rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)
-```
-
-***
-
 ### Multivariate Model Output: Species interactions
@@
 For example, each data point for *Pterygophora* in the *Macrocystis* model is the predicted effect of the current density of Pterygophora on the density of Macrocystis in the next monitoring period.
-Mathematically, each interaction measured is a partial derivative, or element of our sequentially calculated Jacobian matrices, estimating the effect of one species on another, $\delta N_{1}/\delta N_{2}$.
-The distributions of those interactions are shown in the figure below.
+Mathematically, each interaction measured is a partial derivative, or element of our locally calculated Jacobian matrices (the coefficients of our locally-weighted S-maps), estimating the effect of one species on another, $\delta N_{1}/\delta N_{2}$.
+The distributions of those interactions are shown for each separate species model in the figure below.
 
@@
 These are box-and-whisker plots of estimated species interaction strengths from S-map models for the five focal species (panels top to bottom): *Macrocystis* adults, *Macrocystis* juveniles, *Pterygophora*, *Laminaria*, purple urchin, and red urchin.
-Each colored box represents the distribution of all estimated interaction coefficients (x-axis) of an interacting or forcing species (left y-axis) on a modeled species (right y-axis) across all data for a given model (vertical lines: median; box: interquartile range; whiskers extend to data point at most $1.5*IQR$ from the box).
+Each colored box represents the distribution of all estimated interaction coefficients (x-axis) of an interacting or forcing species (left y-axis) on a modeled species (right y-axis) across all data for a given model (white triangles: mean; vertical lines: median; box: interquartile range; whiskers extend to data point at most $1.5*IQR$ from the box).
 Correlation coefficient between predictions and observations denoted for each model.
@@
 
-The interactions of the urchin and algal species show evidence of herbivory (green boxes).
+The interactions of the urchin and algal species show evidence of herbivory (purple boxes).
 Both red and purple urchin density have predominantly negative effect on adult and juvenile *Macrocystis*, *Laminaria*, and *Pterygophora*, effects that were consistent with our hypotheses.
-However, the converse effects of algal density on urchin density are generally small, and not consistently positive.
-
-Few strong and consistent competitive interactions (red boxes) are observed in model estimates.
-There is evidence of competition between the two urchin species, but the interaction is small relative to other forcing factors in the urchin models, and is occasionally positive (apparent mutualism).
+However, the converse effects of algal density on urchin density (green boxes) are generally small, and not consistently positive.
+
+Few strong and consistent competitive interactions (red and orange boxes for urchin and algal competition, respectively) are observed in model estimates.
+Estimates of the interaction between the two urchin species are small and near zero.
+The interactions are small relative to other forcing factors in the urchin models, and are occasionally positive (apparent mutualism).
 Among the algal species, the clearest negative competitive effect is that of *Pterygophora* on adult *Macrocystis* (top panel).
@@
 *Macrocystis* is the only species in the analysis that displays a significant cross-mapping signal (significant causal forcing) with all five included physical variables.
-The negative effects of SWH and SST on adult *Macrocystis*, and the positive effect of the *NPGO*, are consistent with our expectations [@Bell2015].
+The negative effect of SWH on adult *Macrocystis*, and the positive effect of the NPGO, are consistent with our expectations [@Bell2015].
 However, the PDO index has a positive effect on *Macrocystis*, contrary to expectation because in general, positive values of the PDO are associated with warmer SST and nutrient-poor conditions in the northeast Pacific.
@@
 The physical variables are more influential drivers of *Macrocystis* than *Pterygophora* or *Laminaria* dynamics.
-The PDO is the only physical variable casaully linked to *Laminaria* density, and it has predominantly negative but variable effect.
-The SWH, NPGO, MEI, and SST indices do not significantly drive *Laminaria* density.
-Conversely, although the PDO, NPGO, and SST were identified as causal variables for *Pterygophora* density, the magnitudes of their effects are small and inconsistent.
+No physical variable was causally linked to *Laminaria* density.
+Conversely, although the PDO, NPGO, SWH and SST were identified as causal variables for *Pterygophora* density, the magnitudes of their effects are small and inconsistent.
 
@@
 For the purple urchin, the PDO, NPGO, and MEI all have negative but variable effects.
-For red urchin, the effects of SWH and SST are small and negative, with the PDO and NPGO have variable effects.
+For red urchin, the effects of SWH and SST are small and negative, while the effect of the NPGO is slightly positive.
 
@@
     labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+
-    facet_grid(type~external_forcing,scales = "free_y")
+    facet_grid(type~external_forcing,scales = "free_y")+
+    theme(strip.text=element_text(size=8),
+          legend.text = element_text(size=12),
+          legend.position="right",
+          legend.margin=margin(t = -0.5, unit='line'),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"),
+        )
 })

### 2018-06-08 EDM_SNI_analysis.Rmd
edited results language

@@
     + Elevated SST has similar effects to nutrient stress (elevated MEI and PDO, and reduced NPGO), which makes sense: In Southern California, elevated SST is associated with reduced nutrient availability[@Zimmerman1984,@Bell2015].
-Similar effects as described above for nutrient stress hold for elevated temperature.
+Similar effects as described above for nutrient stress (at least as measured by the large oceanographic indices) hold for elevated temperature.
+This suggests both that this type of effect manifests at multiple temporal scales, and that the combination of elevated temperature and stressful nutrient regimes could have compounding effects on species interactions and ecosystem dynamics.
 

### 2018-06-08 EDM_SNI_analysis.Rmd
saved plots

@@
 westend.xmap_all_plot
+# ggsave("plots/ccm_all.png",plot=westend.xmap_all_plot)
 ```
@@
 all.mod.coeffs.plot.box
-# ggsave("all_spp_coffs.png",plot=all.mod.coeffs.plot.box,height=8,width=7.5)
+# ggsave("plots/all_spp_coeffs.png",plot=all.mod.coeffs.plot.box,height=8,width=7.5)
 ```
@@
 all.phys.coeffs.plot
+# ggsave("plots/phys_coeffs.png",plot=all.phys.coeffs.plot)
 ```
@@
 spp.only.mod.coeffs.plot.box
-# ggsave("spp_only_mods_coeffs.png",plot=spp.only.mod.coeffs.plot.box,height=8,width=7.5)
+# ggsave("plots/spp_only_mods_coeffs.png",plot=spp.only.mod.coeffs.plot.box,height=8,width=7.5)
 ```
@@
 })
+ggsave("plots/phys_effects_red.png",plot=phys_effects_on_interactions[[1]])
+ggsave("plots/phys_effects_purp.png",plot=phys_effects_on_interactions[[2]])
+ggsave("plots/phys_effects_lam.png",plot=phys_effects_on_interactions[[3]])
+ggsave("plots/phys_effects_pter.png",plot=phys_effects_on_interactions[[4]])
+ggsave("plots/phys_effects_mac.png",plot=phys_effects_on_interactions[[5]])
+ggsave("plots/phys_effects_ymac.png",plot=phys_effects_on_interactions[[6]])
+
 phys_effects_on_interactions[[1]]
@@
 phys_effects_on_interactions[[6]]
+
 ```

### 2018-06-09 EDM_SNI_analysis.Rmd
saved plots

@@
         legend.text = element_text(size=14),
-        axis.title=element_text(family="Rockwell",size=14,color="black"),
-        axis.text=element_text(family="Rockwell",size=8,color="black"),
+        axis.title=element_text(family="Rockwell",size=14),
+        axis.text=element_text(family="Rockwell",size=8),
         strip.background = element_rect(colour="black"),
         panel.border = element_rect(color="black",fill=NA))
+
+# for black plots
+plot_theme_black <-   theme_minimal()+
+  theme(text=element_text(family="Rockwell",size=12,color="white"),
+        line = element_line(color="gray40"),
+        legend.text = element_text(size=14),
+        axis.title=element_text(family="Rockwell",size=14),
+        axis.text=element_text(family="Rockwell",color="white",size=8),
+        strip.background = element_rect(colour="black"),
+        panel.border = element_rect(color="white",fill=NA),
+        panel.background = element_rect(fill="black"),
+        plot.background = element_rect(fill="black"),
+        panel.grid = element_line(color="gray40"),
+        panel.grid.major = element_line(color="gray40"),
+        panel.grid.minor = element_line(color="gray40"))
 theme_set(plot_theme)
@@
       guides(linetype="none")+
-      theme(text = element_text(size=16),
-            legend.position = c(0.15,0.8),
-            legend.text = element_text(size=12),
+      theme(text = element_text(color="black",size=16),
+            legend.position = c(0.15,0.7),
+            legend.text = element_text(color="black",size=10),
             panel.border = element_blank())
@@
       guide = "colourbar",limits=c(0,0.8),breaks=c(0.2,0.4,0.6,0.8),name=expression(paste(rho, "(skill)"))) +
-  ggtitle("Kelp Forest Convergent Cross Mapping, West End") +
+  ggtitle("Kelp Forest Convergent Cross Mapping") +
   xlab("Predictor (Forcing Variable)") +
   ylab("Predicted Variable") +
-  theme(axis.text.x=element_text(angle = 90, hjust = 1,vjust=0.1))
+  theme(text=element_text(color="black"),
+        plot.background = element_rect(fill="white"),
+        axis.text=element_text(color="black"),
+    axis.text.x=element_text(angle = 90, hjust = 1,vjust=0.1),
+    panel.border = element_blank())
 
 westend.xmap_all_plot
-# ggsave("plots/ccm_all.png",plot=westend.xmap_all_plot)
+# ggsave("plots/ccm_all.png",plot=westend.xmap_all_plot,width=7,height=4)
 ```
@@
 vert.cols <- c(rep("tan3",3),rep("orchid3",2),"tan3",rep("gray70",5))
-
 plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.shape="circle", edge.curved=F,margin=0, vertex.size=30,rescale=T,edge.lty=1,edge.color="gray40",layout=crds,vertex.color=vert.cols,vertex.frame.color=NA,main="")
@@
 scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
+
+dev.copy(pdf,"plots/multi_attractor.pdf")
+dev.off()
 
@@
 Similar is true for the adult *Macrocystis*: interactions between the giant kelp and the other algae species are elevated under nutrient- or temperature-stressed conditions.
-Increased wave height (physical disturbance) seems to enhance the negative effect of *Pterygophora* on *Macrocystis* adults.
+Increased wave height (physical disturbance) does not seem to have a large effect on the effect of *Laminaria* and *Pterygophora* on *Macrocystis*
     + The effect of purple urchin on both *Macrocystis* adult and juvenile density is stronger than the red urchin effect, and more variable with changing conditions.
@@
     + Higher physical disturbance levels results in reduced interaction strength between the urchin species.
-    + *Pterygophora*'s competitive ability is enhanced with increased disturbance[@Dayton1999], evidenced by an increase in the negative effect of *Pterygophora* on adult *Macrocystis*.
     + *Pterygophora* and *Macrocystis* are both subject to a greater negative effect of herbivory under increased disturbance.
@@
 
+### Effect of disturbance on *Macrocystis*
+```{r disturbance and mac,fig.height=7,fig.width=5}
+# with boxplot
+quants <- c(0.05,0.5,0.95)
+waves.mac.pter <- westend %>%
+  select(one_of(c("pter","lam","mac","period"))) %>%
+  distinct() %>%
+  right_join(all.mod.coeffs.long,by="period")%>%
+  gather("external_spp","ext_value",one_of(c('pter','lam','mac')))%>%
+  filter(effect%in% c('waves','pter','lam'),mod=='mac')%>%
+  ggplot(aes(ext_value,value))+
+  geom_point(col='gray60',alpha=0.6)+
+  geom_hline(yintercept = 0,linetype=2)+
+  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+  labs(x="Normalized Density",y="Effect of SWH on Macrocystis")+
+  facet_grid(external_spp~effect,scales="free")
+waves.mac.pter
+
+```
 ## References

### 2018-06-13 EDM_SNI_analysis.Rmd
tailing exploratory analyses

@@
 
+theme_set(plot_theme)
+```
+
+## Introduction
+
+This document accompanies the manuscript "..." The following analyses lay out the building blocks of the empirical dynamic modeling approach [@Deyle2016; @Chang2017] that produced the final version of the analyses in the paper.
+
+In the analysis that follows, in each step that involves a new type of EDM method, its logic will be briefly described.
+However, keep in mind that the basic theory behind each individual method is the same: We use some dynamic information (from single or multiple time series) to reconstruct a shadow attractor/manifold, and then use the properties of that manifold to make predictions or estimate interactions.
+
+## 1. Variables in the Data
+
+### The San Nicolas Island Dataset
+
+San Nicolas Island is a small, remote island situated about 100 kilometers offshore from southern California.
+The island itself is small, about 14 kilometer long and 5 km wide.
+The data in the analysis is from a sampling station on the western end of San Nicolas Island.
+The benthic monitoring data herein have been collected more or less every six months for more than 35 years by the USGS and its Western Ecological Research Center (USGS-WERC), and in 2013 the datasets were made available publicly through Ecological Archives:
+
+Michael C.
+Kenner, James A.
+Estes, M.
+Tim Tinker, James L.
+Bodkin, Robert K.
+Cowen, Christopher Harrold, Brian B.
+Hatfield, Mark Novak, Andrew Rassweiler, and Daniel C.
+Reed.
+2013.
+A multi-decade time series of kelp forest community structure at San Nicolas Island, California (USA).
+Ecology 94:2654.
+http://dx.doi.org/10.1890/13-0561.1
+
+### Physical Oceanographic Data
+
+A body of other research has established that a combination of physical forcing (waves, storms), temperature, and lower frequency climate modes (e.g., El Ninos) have an important influence on the dynamics of kelp forests [@Reed2011; @Cavanaugh2011; @Bell2015,@Young2015].
+With these data, we can draw connections between the physical variables and the species interactions in our constrained trophic web.
+
+We have four datasets, already processed into the same time frame (periods) as the SNI benthic monitoring data:
+
+* **[The Multivariate ENSO index (MEI)](http://www.esrl.noaa.gov/psd/enso/mei)**
+    + The first principal component of a composite set of physical parameters
+    + Positive values of the MEI index are generally associated with El Nino conditions, decreases in wind-driven upwelling, warmer surface waters and nutrient-poor conditions
+    + Variable here is the average index value for the four months preceding each Spring or Fall monitoring period (i.e., December to March or June to September, respectively)
+
+* **[The Pacific Decadal Oscillation index (PDO)](http://research.jisao.washington.edu/pdo/)**
+    + Leading empirical orthogonal function (EOF) of monthly sea surface temperature anomalies (SST-A) over the North Pacific (poleward of 20° N) after the global average sea surface temperature has been removed
+    + Positive PDO values indicate warmer SST, and nutrient-poor conditions along the western coast of the contiguous United States
+    + Aggregated and averaged the same way as MEI
+
+* **[The North Pacific Gyre Oscillation (NPGO)](http://www.o3d.org/npgo/)**
+    + From [@DiLorenzo2008]
+    + Climate pattern that emerges as the 2nd dominant mode of sea surface height variability (2nd EOF SSH) in the Northeast Pacific
+    + Better correlated with salinity, nutrients, and chlorophyll than PDO, showing forcing for the planktonic community
+    + Strong predictor of upwelling cells south of 38 deg N
+    + Aggregated and averaged the same way as MEI and PDO
+
+* **Sea surface temperature (SST)**
+    + Two sources (to fill in data gaps):
+      + Sea surface temperature data directly from Begg Rock and San Nicolas Island buoys, from the [Coastal Data Information Program (CDIP)](http://cdip.ucsd.edu)
+      + NOAA's [Optimally Interpolated Sea Surface Temperature](https://www.ncdc.noaa.gov/oisst)
+    + Similar to the above, value is an average SST for the four months preceding each period
+* **Maximum significant wave height (Hs)**
+    + Modeled wave height data from from the [Geophysical Fluid Dynamics Laboratory](http://cmgwindwave.usgsportals.net/) and from Tom Bell at UCSB
+    + Signficant wave height is defined as the average height, in meters, of the one third highest waves in the record
+    + Instead of an average, value here is the maximum significant wave height of the four months preceding each period.
+This is meant to capture any large storm events, as well as general level of physical disturbance
+
+Unlike the biological data, where there are unique spatial replicates, the physical data have only one value for each of the 63 monitoring periods, and hence their values are replicated (copied) for each site to match the total length of the biological data.
+
+#### Note about data processing
+
+Usual best practice for EDM is to create normalized time series, so that state-space reconstructions are not distorted by differences in orders of magnitude between variables.
+In addition, in this study we concatenate each species/station/swath time series into one long time series for each species, while preserving a time indicator (variable 'period') and a site identifier, to ensure that we do not "cross" the boundaries of replicate time series in analyses.
+Empirical dynamic modeling can use multiple spatial replicates in lieu of increased length of individual time series, to maximize the dynamic information that can be drawn from the system [@Hsieh2008;@Clark2015;@Chang2017].
+
+All raw data is available online or in the included files. We join the monitoring and physical datasets and normalized all time series (see scripts in the "data" folder for details:
+
+```{r load data,message=FALSE,echo=F}
+source("data/data preparation scripts/produce_final_data.R")
+```
+
+```{r list of variables,echo=F,fig.cap="List of variables"}
+tibble(Variable=colnames(westend),Description=c("Monitoring Transect","Monitoring Period","Laminaria Normalized Density",
+                                            "Macrocystis pyrifera >1m Normalized Density","Pterygophora Normalized Density",
+                                            "Strongylocentrotus purpuratus Normalized Density","Mesocentrotus franciscanus Normalized Density",
+                                            "Macrocystis pyrifera <1m Normalized Density","Normalized Multivariate ENSO Index",
+                                            "Normalized Pacific Decadal Oscillation","Normalized North Pacific Gyre Oscillation",
+                                            "Normalized Maximum Significant Wave Height","Normalized Sea Surface Temperature")) %>%
+  knitr::kable()
+```
+
+```{r mean density over time, echo=F,message=F}
+# black plot theme (for presentations)
 # for black plots
@@
         panel.grid.minor = element_line(color="gray40"))
-theme_set(plot_theme)
-```
-
-## Introduction
-
-This document accompanies the manuscript "..." The following analyses lay out the building blocks of the empirical dynamic modeling approach [@Deyle2016; @Chang2017] that produced the final version of the analyses in the paper.
-
-In the analysis that follows, in each step that involves a new type of EDM method, its logic will be briefly described.
-However, keep in mind that the basic theory behind each individual method is the same: We use some dynamic information (from single or multiple time series) to reconstruct a shadow attractor/manifold, and then use the properties of that manifold to make predictions or estimate interactions.
-
-## 1. Variables in the Data
-
-### The San Nicolas Island Dataset
-
-San Nicolas Island is a small, remote island situated about 100 kilometers offshore from southern California.
-The island itself is small, about 14 kilometer long and 5 km wide.
-The data in the analysis is from a sampling station on the western end of San Nicolas Island.
-The benthic monitoring data herein have been collected more or less every six months for more than 35 years by the USGS and its Western Ecological Research Center (USGS-WERC), and in 2013 the datasets were made available publicly through Ecological Archives:
-
-Michael C.
-Kenner, James A.
-Estes, M.
-Tim Tinker, James L.
-Bodkin, Robert K.
-Cowen, Christopher Harrold, Brian B.
-Hatfield, Mark Novak, Andrew Rassweiler, and Daniel C.
-Reed.
-2013.
-A multi-decade time series of kelp forest community structure at San Nicolas Island, California (USA).
-Ecology 94:2654.
-http://dx.doi.org/10.1890/13-0561.1
-
-### Physical Oceanographic Data
-
-A body of other research has established that a combination of physical forcing (waves, storms), temperature, and lower frequency climate modes (e.g., El Ninos) have an important influence on the dynamics of kelp forests [@Reed2011; @Cavanaugh2011; @Bell2015,@Young2015].
-With these data, we can draw connections between the physical variables and the species interactions in our constrained trophic web.
-
-We have four datasets, already processed into the same time frame (periods) as the SNI benthic monitoring data:
-
-* **[The Multivariate ENSO index (MEI)](http://www.esrl.noaa.gov/psd/enso/mei)**
-    + The first principal component of a composite set of physical parameters
-    + Positive values of the MEI index are generally associated with El Nino conditions, decreases in wind-driven upwelling, warmer surface waters and nutrient-poor conditions
-    + Variable here is the average index value for the four months preceding each Spring or Fall monitoring period (i.e., December to March or June to September, respectively)
-
-* **[The Pacific Decadal Oscillation index (PDO)](http://research.jisao.washington.edu/pdo/)**
-    + Leading empirical orthogonal function (EOF) of monthly sea surface temperature anomalies (SST-A) over the North Pacific (poleward of 20° N) after the global average sea surface temperature has been removed
-    + Positive PDO values indicate warmer SST, and nutrient-poor conditions along the western coast of the contiguous United States
-    + Aggregated and averaged the same way as MEI
-
-* **[The North Pacific Gyre Oscillation (NPGO)](http://www.o3d.org/npgo/)**
-    + From [@DiLorenzo2008]
-    + Climate pattern that emerges as the 2nd dominant mode of sea surface height variability (2nd EOF SSH) in the Northeast Pacific
-    + Better correlated with salinity, nutrients, and chlorophyll than PDO, showing forcing for the planktonic community
-    + Strong predictor of upwelling cells south of 38 deg N
-    + Aggregated and averaged the same way as MEI and PDO
-
-* **Sea surface temperature (SST)**
-    + Two sources (to fill in data gaps):
-      + Sea surface temperature data directly from Begg Rock and San Nicolas Island buoys, from the [Coastal Data Information Program (CDIP)](http://cdip.ucsd.edu)
-      + NOAA's [Optimally Interpolated Sea Surface Temperature](https://www.ncdc.noaa.gov/oisst)
-    + Similar to the above, value is an average SST for the four months preceding each period
-* **Maximum significant wave height (Hs)**
-    + Modeled wave height data from from the [Geophysical Fluid Dynamics Laboratory](http://cmgwindwave.usgsportals.net/) and from Tom Bell at UCSB
-    + Signficant wave height is defined as the average height, in meters, of the one third highest waves in the record
-    + Instead of an average, value here is the maximum significant wave height of the four months preceding each period.
-This is meant to capture any large storm events, as well as general level of physical disturbance
-
-Unlike the biological data, where there are unique spatial replicates, the physical data have only one value for each of the 63 monitoring periods, and hence their values are replicated (copied) for each site to match the total length of the biological data.
-
-#### Note about data processing
-
-Usual best practice for EDM is to create normalized time series, so that state-space reconstructions are not distorted by differences in orders of magnitude between variables.
-In addition, in this study we concatenate each species/station/swath time series into one long time series for each species, while preserving a time indicator (variable 'period') and a site identifier, to ensure that we do not "cross" the boundaries of replicate time series in analyses.
-Empirical dynamic modeling can use multiple spatial replicates in lieu of increased length of individual time series, to maximize the dynamic information that can be drawn from the system [@Hsieh2008;@Clark2015;@Chang2017].
-
-All raw data is available online or in the included files. We join the monitoring and physical datasets and normalized all time series (see scripts in the "data" folder for details:
-
-```{r load data,message=FALSE,echo=F}
-source("data/data preparation scripts/produce_final_data.R")
-```
-
-```{r list of variables,echo=F,fig.cap="List of variables"}
-tibble(Variable=colnames(westend),Description=c("Monitoring Transect","Monitoring Period","Laminaria Normalized Density",
-                                            "Macrocystis pyrifera >1m Normalized Density","Pterygophora Normalized Density",
-                                            "Strongylocentrotus purpuratus Normalized Density","Mesocentrotus franciscanus Normalized Density",
-                                            "Macrocystis pyrifera <1m Normalized Density","Normalized Multivariate ENSO Index",
-                                            "Normalized Pacific Decadal Oscillation","Normalized North Pacific Gyre Oscillation",
-                                            "Normalized Maximum Significant Wave Height","Normalized Sea Surface Temperature")) %>%
-  knitr::kable()
-```
-
-```{r mean density over time, echo=F,message=F}
+
 # Species naming key for use in plotting
@@
 westend.mean_dens_plot
-rm(pd,period.key,periods.first,westend.long,westend.meandens)
+rm(pd,westend.long,westend.meandens)
+
+```
+
+```{r physical variables timeseries}
+mei.pdo.npgo.ts.plot <-westend %>%
+  ungroup()%>%
+  slice(1:63)%>%
+  select(period,mei,pdo,npgo)%>%
+  gather("driver","value",mei:npgo,-period)%>%
+  left_join(namekey, by=c("driver"="short"))%>%
+  left_join(periods.first,by="period") %>%
+  ggplot(aes(date_str,value,col=driver))+
+  geom_line(lwd=1.5)+
+  geom_hline(yintercept=0,linetype=2,col="black")+
+  labs(title="Physical Drivers: Climate Modes",x="Year",y="Normalized Index Value")+
+  plot_theme_black
+sst.waves.ts.plot <-westend %>%
+  ungroup()%>%
+  slice(1:63)%>%
+  select(period,sst,waves)%>%
+  gather("driver","value",sst,waves,-period)%>%
+  left_join(namekey, by=c("driver"="short"))%>%
+  left_join(periods.first,by="period") %>%
+  ggplot(aes(date_str,value,col=driver))+
+  geom_line(lwd=1.5)+
+  geom_hline(yintercept=0,linetype=2,col="black")+
+  labs(title="Physical Drivers: Temperature and Disturbance",x="Year",y="Normalized Index Value")+
+  plot_theme_black
+mei.pdo.npgo.ts.plot
+sst.waves.ts.plot
+# ggsave("plots/climate_ts_plot.png",plot=mei.pdo.npgo.ts.plot,width=7,height=4)
+# ggsave("plots/sst_waves_ts_plot.png",plot=sst.waves.ts.plot,width=7,height=4)
 
@@
 ***
+
+### Example CCM
+
+Each individual CCM analysis involves looks something like this:
+
+```{r ccm example}
+tempE <- westend.bestE['purp']
+temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 'mac',target_column = 'purp',lib_sizes = c(10,25,50,75,100,125,150,200,300,400,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
+
+mac_xmap_purp <- temp %>%
+  group_by(lib_size)%>%
+  summarise(rhomean=mean(rho,na.rm=T),upper=rhomean+sd(rho,na.rm=T),lower=rhomean-sd(rho,na.rm=T))%>%
+  ungroup()%>%
+  ggplot(aes(lib_size,rhomean))+
+  geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3,fill="red")+
+  geom_line(color="darkorchid3")+
+  labs(x="Library Size",y=expression(paste(rho, " (predictive skill)")),title="Macrocystis xmap Purple Urchin")
+
+mac_xmap_purp
+# ggsave("plots/ccm_example.png",plot=mac_xmap_purp,width=6,height=4)
+```
 
@@
 phys_effects_on_interactions[[6]]
+```
+
+```{r some single plots}
+# macrocystis, pterygophora, and temperature
+quants<-c(0.05,0.5,0.95)
+all.coeffs.with.phys %>%
+  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%
+  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+   geom_point(col='gray60',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="white")+
+   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
+   labs(x="Normalized SST",y="Interaction Strength")+
+   facet_grid(~mod,scales = "free_y")+
+  plot_theme_black+
+    theme(strip.text=element_blank(),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+
+# macrocystis, pterygophora, and NPGO
+all.coeffs.with.phys %>%
+  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="npgo")%>%
+  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+   geom_point(col='gray60',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="white")+
+   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
+   labs(x="Normalized NPGO",y="Interaction Strength")+
+   facet_grid(~mod,scales = "free_y")+
+  plot_theme_black+
+    theme(strip.text=element_blank(),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+
+# macrocystis, pterygophora, herbivory, and NPGO
+all.coeffs.with.phys %>%
+  filter(type=="herbivory",mod%in% c('mac','pter'),effect %in% c('red','purp'),external_forcing=="npgo")%>%
+  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+   geom_point(col='gray60',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="white")+
+   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
+   labs(x="Normalized NPGO",y="Interaction Strength")+
+   facet_grid(~mod,scales = "free_y")+
+  plot_theme_black+
+    theme(strip.text=element_blank(),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+
+# macrocystis, pterygophora, and waves
+all.coeffs.with.phys %>%
+  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="waves")%>%
+  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+   geom_point(col='gray60',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="white")+
+   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
+   labs(x="Normalized SWH",y="Interaction Strength")+
+   facet_grid(~mod,scales = "free_y")+
+  plot_theme_black+
+    theme(strip.text=element_blank(),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
 
@@
 ```{r disturbance and mac,fig.height=7,fig.width=5}
-# with boxplot
 quants <- c(0.05,0.5,0.95)
@@
 waves.mac.pter
-
+```
+
+### Effect of Macrocystis on Laminaria/Pterygophora interaction
+```{r disturbance and mac,fig.height=7,fig.width=5}
+quants <- c(0.05,0.5,0.95)
+mac.lam.pter <- westend %>%
+  select(one_of(c("mac","period"))) %>%
+  distinct() %>%
+  right_join(all.mod.coeffs.long,by="period")%>%
+  gather("external_spp","ext_value",mac)%>%
+  filter(effect=='pter',mod=='lam')%>%
+  ggplot(aes(ext_value,value))+
+  geom_point(col='gray60',alpha=0.6)+
+  geom_hline(yintercept = 0,linetype=2)+
+  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+  labs(x="Normalized Density",y="Effect of Pterygophora on Laminaria")
+mac.lam.pter
 ```

### 2018-07-04 EDM_SNI_analysis.Rmd
experimental analyses at the end

@@
 
-  # Model output-- time, observations, predictions. time is truncated, so re-expand it to fit with original periods
+  # gather and organize model output
   t <- best.mod$model_output[[1]]$time
 
+  coeff.names <- c(causalvars,"const")
   # Save interaction coefficients
   coeffs <- best.mod$smap_coefficients[[1]]
-  names(coeffs) <- c(causalvars,"const")
+  names(coeffs) <- coeff.names
+
+  # # Model output-- time, observations, predictions. match time indicator with observations (one period ahead)
   coeffs <- coeffs %>% mutate(time=t)%>%complete(time=full_seq(x=c(1,630),1)) %>% mutate(period=rep(1:63,10))
 
-  # long form for plotting
+  # Save coefficient variances (sensu Deyle 2016; we have a nxn variance-covariance matrix for each estimated point, where n is the total number of causal variables for whom coefficients are estimated, plus the constant term)
+  # Variances for individual coefficients are the diagonals of each covariance matrix
+  coeffs.variances <- best.mod$smap_coefficient_covariances[[1]] %>%
+    purrr::map_dfr(function(x) {
+      # if no estimate, no variance associated with it (NULL)
+      if(is.null(x)) x <- matrix(NaN,nrow=length(coeff.names),ncol=length(coeff.names))
+      # pull out diagonal elements, add variable names
+      out <- t(diag(x)) %>% as.data.frame()
+      names(out) <- coeff.names
+      out
+      })%>%
+
+    # add period identifier and overall time stamp
+    # mutate(time=t) %>% complete(time=full_seq(c(1,630),1)) %>% mutate(period=rep(1:63,10))
+    mutate(time=t)%>%complete(time=full_seq(x=c(1,630),1)) %>% mutate(period=rep(1:63,10))
+
+  # long form for plotting, including variance
   coeffs.long <- coeffs %>%
@@
     left_join(namekey, by=c("effect"="short"))
+
+  # long form variances
+  coeffs.variances.long <- coeffs.variances %>%
+    gather(key=effect,value=variance,-period,-time,na.rm=T) %>%
+    left_join(namekey,by=c("effect"="short"))
+
+  coeffs.long <- coeffs.long %>%
+    left_join(coeffs.variances.long,by=c("time","period","effect","dataset","long","plotting")) %>%
+    # add standard deviation and +/- 95% CI (1.96*SD)
+    mutate(sd=sqrt(variance),upper=value+1.96*sd,lower=value-1.96*sd)
 
@@
 
-  outlist <- list(model=best.mod,coeffs=coeffs,coeffs.long= coeffs.long,plot=int.box)
+  outlist <- list(model=best.mod,coeffs=coeffs,coeffs.variances=coeffs.variances,coeffs.long= coeffs.long,plot=int.box)
   return(outlist)
@@
 
-```{r phys coeffs,echo=F,message=F,warning=F,fig.height=8}
+```{r phys coeffs,echo=F,message=F,warning=F,fig.height=8,fig.width=8}
 # with boxplot
@@
 ```{r individual_effects,fig.width=7}
-
 phys_effects_on_interactions <- purrr::map(as.list(study_spp),function(spp) {
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
     geom_point(col='gray60',alpha=0.6)+
@@
           legend.margin=margin(t = -0.5, unit='line'),
-          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"),
+          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line")
         )
@@
   filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%
+
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
     geom_point(col='gray60',alpha=0.6)+
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+ 
   geom_point(col='gray60',alpha=0.6)+
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
     geom_point(col='gray60',alpha=0.6)+
@@
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
+  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
   geom_point(col='gray60',alpha=0.6)+
@@
 ```
+### Effect of urchin density on Macrocystis/Pterygophora interaction
+
+```{r urchin dens mac pter}
+quants <- c(0.05,0.5,0.95)
+urchin.mac.pter <- westend %>%
+  select(one_of(c("red","period"))) %>%
+  distinct() %>%
+  mutate(time=row_number())%>%
+  right_join(all.mod.coeffs.long,by=c("period",'time'))%>%
+  gather("external_spp","ext_value",red)%>%
+  filter(effect %in% c('pter','mac'),mod %in% c('pter','mac'),!is.na(ext_value))%>%
+  ggplot(aes(ext_value,value))+
+  geom_point(col='gray60',alpha=0.6)+
+  geom_hline(yintercept = 0,linetype=2)+
+  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
+  labs(x="Normalized Red Urchin Density",y="Interaction Strength")+
+  facet_wrap(~mod)
+urchin.mac.pter
+```
+
+### Do conditions affect average interaction strength?
+
+```{r conditions and interaction strength}
+# disturbance. top quartile vs. bottom quartile
+all.coeffs.with.phys %>%
+  mutate(av_value=abs(value))%>%
+  filter(external_forcing=='pdo', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
+  mutate(stress_level=ifelse(ext_value>=quantile(westend$pdo,0.75,na.rm=T),'high','low'))%>%
+  ggplot(aes(av_value,..density..,fill=stress_level))+
+  geom_density(alpha=0.6)+
+  geom_vline(xintercept = 0,linetype=2,color='black')+
+  labs(x='Interaction Strength',y='Probability Density')
+
+```
+
 ## References

### 2018-07-14 EDM_SNI_analysis.Rmd
exploratory analyes at the end

@@
 
-westend.bestE["ymac"] <- 3
-westend.bestE["pter"] <- 6
+# westend.bestE["ymac"] <- 3
+# westend.bestE["pter"] <- 6
 
@@
 ```{r causal vars fxn,message=F,echo=F}
-causal_vars <- function(pmat,vars) {
-  out <- list()
-  for(i in 1:length(vars)) {
-    v <- vars[i]
-    causes <- colnames(pmat)[pmat[v,]==1]
-    causes <- causes[!is.na(causes)]
-    out[[v]] <- c(causes,v)
-  }
-  return(out)
-}
-westend.causal.vars <- causal_vars(westend.ptot.mat,vars=c("mac","purp","lam","ymac","red","pter"))
+# Pull out causal vars from CCM for each species
+
+westend.causal.vars <- purrr::map(c("mac","purp","lam","ymac","red","pter"), function(x) {
+  westend.xmap_rast %>%
+    filter(predictee==x,!is.na(rho))->out
+  out <- c(x,out$predictor)
+  out
+})
+
+names(westend.causal.vars)=c("mac","purp","lam","ymac","red","pter")
+
+### NOT RUN: VERSION OF MODELS WHERE MULTIVARIATE S-MAPS EQUALS BEST EMBEDDING DIMENSION
+# For each, take only the top n causal vars, where n is equal to the best embedding dimension from above
+# If n > total number of significant variables from CCM, substitute lagged versions of the variable for dimensions
+# westend.causal.vars <- purrr::map(c("mac","purp","lam","ymac","red","pter"),function(x){
+#   # number of predictor variables (minus 1 because we use the species' own time series)
+#   nvars <- westend.bestE[[x]]
+#   westend.xmap_rast %>%
+#     select(predictee,predictor,rho)%>%
+#     filter(predictee==x)%>%
+#     top_n(nvars-1,wt=rho) -> out
+#   out <- c(x,out$predictor)
+#   if(length(out)<nvars) {
+#     ndiff <- nvars-length(out)
+#     out <- c(out,paste0(x,1:ndiff))
+#   }
+#   out
+# })
+# names(westend.causal.vars) <- c("mac","purp","lam","ymac","red","pter")
+# Have to add lagged variables for ymac
+# westend <- westend %>% group_by(site) %>% mutate(ymac1=lag(ymac,1)) %>% ungroup()
+
 ```
@@
                                y=namekey$long[match(x,namekey$short)];paste(y,sep=",",collapse=", ")}),
+                             Theta = map_dbl(westend.multi.mods,function(x) x$model$theta),
                              Rho = map_dbl(westend.multi.mods,function(x) x$model$rho),
@@
 westend.model.perf %>% select(-mod) %>% 
-  kable("latex",col.names = c("Modeled Species","Predictors","$\\rho$","MAE"),digits=3,escape=F) %>%
+  kable("latex",col.names = c("Modeled Species","Predictors","$\\theta$","$\\rho$","MAE"),digits=3,escape=F) %>%
   kable_styling(full_width = F) %>%
@@
 all.coeffs.with.phys %>%
-  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%
+  filter(type=="algal competition",mod%in% c('ymac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%
 
@@
     geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-    geom_point(col='gray60',alpha=0.6)+
-    geom_hline(yintercept = 0,linetype=2,col="white")+
+    geom_point(col='black',alpha=0.6)+
+    geom_hline(yintercept = 0,linetype=2,col="black")+
     geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
@@
     facet_grid(~mod,scales = "free_y")+
-  plot_theme_black+
-    theme(strip.text=element_blank(),
-          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
 
@@
   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+ 
-  geom_point(col='gray60',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="white")+
+  geom_point(col='black',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="black")+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
@@
    facet_grid(~mod,scales = "free_y")+
-  plot_theme_black+
-    theme(strip.text=element_blank(),
-          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+  theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
 
@@
    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-    geom_point(col='gray60',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="white")+
+    geom_point(col='black',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="black")+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
@@
    facet_grid(~mod,scales = "free_y")+
-  plot_theme_black+
-    theme(strip.text=element_blank(),
-          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
+    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
 
@@
   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-  geom_point(col='gray60',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="white")+
+  geom_point(col='black',alpha=0.6)+
+   geom_hline(yintercept = 0,linetype=2,col="black")+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
@@
    facet_grid(~mod,scales = "free_y")+
-  plot_theme_black+
     theme(strip.text=element_blank(),
@@
 Purple urchin herbivory as a negative influence on *Macrocystis* is enhanced with elevated disturbance, MEI, and PDO, and slightly reduced with an increase in the NPGO.
-    + The strenghts of intraspecies effects (effects of *Macrocystis* adults on juveniles and vice versa) are also influenced by physical conditions, but perhaps in a different way than might be expected.
+    + The strengths of intraspecies effects (effects of *Macrocystis* adults on juveniles and vice versa) are also influenced by physical conditions, but perhaps in a different way than might be expected.
 The positive effect of juvenile *Macrocystis* on adult *Macrocystis* seems to be **enhanced** under nutrient-stressed conditions and higher temperatures, and reduced with greater nutrient availability and increased disturbance.
@@
   mutate(av_value=abs(value))%>%
-  filter(external_forcing=='pdo', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
+  filter(external_forcing=='waves', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
   mutate(stress_level=ifelse(ext_value>=quantile(westend$pdo,0.75,na.rm=T),'high','low'))%>%
-  ggplot(aes(av_value,..density..,fill=stress_level))+
+  ggplot(aes(log(value),..density..,fill=stress_level))+
   geom_density(alpha=0.6)+
@@
 
+### Mean vs. variance in interactions
+
+```{r mean vs variance}
+int_mean_sd <-spp.only.mod.coeffs %>%
+  filter(!is.na(type))%>%
+  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
+  group_by(mod,effect,type) %>%
+  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),cv_int=sd_int/abs_mean*100,
+            perc_neutral=sum(neutral)/n()*100)%>%
+  ungroup()
+
+# Mean interaction strength by interaction type
+int_mean_sd %>%
+  filter(mod != "ymac",effect != 'ymac')%>%
+  group_by(type)%>%
+  summarise(mean_str=mean(abs_mean),ci95=sd(abs_mean)/sqrt(n())*1.96)%>%
+  ungroup()%>%
+  ggplot(aes(type,mean_str))+
+  geom_col()+
+  geom_errorbar(aes(ymax=mean_str+ci95,ymin=mean_str-ci95),position='dodge',width=0.2)+
+  labs(x='Interaction Type',y='Interaction Strength')
+
+# scatterplot mean vs. sd (removing intraspecies interactions)
+int_mean_sd %>%
+  filter(mod != "ymac",effect != 'ymac')%>%
+  ggplot(aes(abs_mean,sd_int))+
+  geom_point(aes(color=type),size=5)+
+  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  geom_smooth(method="lm",se=F,col="black")+
+  labs(x="Mean Interaction Strength",y="SD Interaction Strength")
+
+# CV calculated using the correction for sample size (cv*=1+1/4n * cv)
+int_mean_sd %>%
+  filter(mod != 'ymac',effect!='ymac')%>%
+  ggplot(aes(type,cv_int))+
+  geom_col()+
+
+  labs(x="Interaction Type",y="CV of Measured Interactions")
+
+  # summarise(mean_cv=mean(cv_int),sd_cv=sd(cv_int),upper=mean_cv+sd_cv/sqrt(n())*1.96,lower=mean_cv-sd_cv/sqrt(n())*1.96)%>%
+# geom_errorbar(aes(ymax=upper,ymin=lower),position="dodge",width=0.25)+
+```
+
+### Bidirectional interactions (is the effect of A on B the same as B on A)
+```{r bidirectional}
+int_mean_sd %>%
+  mutate(rev_effect=mod,rev_mod=effect) %>%
+  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))%>%
+  mutate(upper1=mean_int.x+sd_int.x,lower1=mean_int.x-sd_int.x,upper2=mean_int.y+sd_int.y,lower2=mean_int.y-sd_int.y)-> int_mean_sd2
+
+int_mean_sd2 %>%
+  filter(type.x %in% c('algal competition','urchin competition'),type.y %in% c('algal competition','urchin competition'))%>%
+  ggplot(aes(mean_int.x,mean_int.y))+
+  geom_point()+
+  geom_smooth(method="lm")+
+  geom_abline(slope=1,intercept=0,linetype=2)+
+  geom_errorbar(aes(ymin=lower2,ymax=upper2))+
+  geom_errorbarh(aes(xmin=lower1,xmax=upper1))+
+  labs(x='Effect of A on B',y='Effect of B on A')
+```
+
+### Within-model cross correlation of effects
+
+```{r cross correlation of effect}
+cc_effects <- purrr::map(c("mac","purp","lam","ymac","red","pter"),function(x){
+  spp.only.mod.coeffs %>% filter(mod==x,effect!='const') %>%
+    select(time,period,effect,value)%>%spread(effect,value) %>%
+    left_join(select(westend,one_of(phys.vars),period),by="period")%>%distinct()%>%
+    select(-time,-period)%>%
+    cor(use="complete.obs")
+})
+names(cc_effects) <- c("mac","purp","lam","ymac","red","pter")
+cc_effects_plot <- function(model,effect1,effect2){
+  spp.only.mod.coeffs %>% filter(mod==model,effect%in%c(effect1,effect2)) %>%
+    select(time,period,effect,value)%>%spread(effect,value)%>%
+    left_join(select(westend,one_of(phys.vars),period),by="period")%>%distinct()%>%
+    ggplot(aes_string(effect1,effect2))+geom_point()+geom_smooth(method='lm',se=F)
+}
+```
+
 ## References

### 2018-07-14 EDM_SNI_analysis.Rmd
added CCM between interactions and physical vars

@@
       temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 2+j,lib_sizes = c(10,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
+
      # mean rho at library size 500
-      rhomean <- temp %>% filter(lib_size==500) %>% ccm_means()
-      westend.xmap_mat[i,j] <- rhomean$rho
+      rhomeans <- temp %>% ccm_means()
+      westend.xmap_mat[i,j] <- rhomeans$rho[rhomeans$lib_size==500]
 
@@
 
-      # second p-value (rho at lib-size 500 greater than rho at lib-size 10?)
+      # second p-value (rho at lib-size 500 greater than rho at lib-size 10? By more than 0.1?)
       p2 <- t.test(temp$rho[temp$lib_size==10],temp$rho[temp$lib_size==500])$p.value
+      if((rhomeans$rho[rhomeans$lib_size==500]-rhomeans$rho[rhomeans$lib_size==10])<0.1) p2 <- 0
       westend.p2.mat[i,j] <- as.numeric(p2)
@@
 
+## Convergent Cross Mapping interactions with Physical Forcing
+
+Now that we have reconstructed species interactions, we can try using CCM to establish the relationship between the strength of those interactions and physical drivers.
+
+We first create a time series block of the reconstructed attractors, then find their best embedding dimension, followed by cross mapping the interactions to the physical variables
+
+```{r ints block}
+# create time series block of all interactions (later we could come back to this to deal with error)
+interactions.block <- spp.only.mod.coeffs %>%
+  filter(type != "intraspecies", !is.na(type))%>%
+  select(time,period,mod,effect,value)%>%
+  unite("effect_mod",effect,mod)%>%
+  spread(effect_mod,value) %>%
+  complete(time=full_seq(x=c(1,630),1))%>%
+  # join physical variables
+  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
+  distinct()
+```
+
+```{r simplex ints}
+# Run simplex projection for each species interaction
+interaction_names <- colnames(interactions.block)[3:29]
+interactions.simp.list <- list()
+
+for(i in 2:length(interaction_names)) {
+  x <- interaction_names[i]
+  dat <- interactions.block %>% select(matches(x)) %>% as.data.frame()
+  out <- simplex(as.numeric(dat[,1]),lib=as.matrix(westend.segs),E=2:15,silent=T) %>%
+    mutate(int=x)
+ interactions.simp.list[[x]] <- out
+}
+
+# Save best embedding dimensions
+ints.bestE <- purrr::map_chr(interactions.simp.list,function(x) {
+  temp <- x %>% filter(!is.na(rho))
+  temp$E[temp$rho==max(temp$rho)]
+})
+```
+
+```{r ccm}
+## CCM checking physical variables against interactions
+# Rows hold predicted variables, columns hold forcing variables. In this case, the columns (forcing factors) are the physical variables, while the rows are the interactions
+n_col <- length(phys.vars)
+n_row <- dim(interactions.block)[2]-7
+col_names <- phys.vars
+row_names <- interaction_names
+interactions.xmap_mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))
+
+## CCM causation criterion 1: cross-map skill greater than zero
+# matrix to store a bootstrapped p-value, measuring the probability that a given xmap is greater than zero (calculated as 1 minus the number of positive results for rho divided by the number of iterations)
+interactions.p1.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))
+
+## CCM causation criterion 2: evidence for convergence
+# similarly, matrix to store a bootstrapped p-value, this time a t-test value between library size 10 and library size 500, to see if the rho at large library is significantly greater than the rho at small library size (i.e., looking for convergence).
+interactions.p2.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))
+
+# if both p1 and p2 are positive, indicate overall significant causal signal
+interactions.ptot.mat <- array(NA,dim=c(n_row,n_col),dimnames=list(row_names,col_names))
+
+## Run CCM for each combination of variables
+for(i in 1:n_row) {
+  for(j in 1:n_col) {
+      # remember, we use the best embedding dimension for the target variable (the variable we're cross-mapping to, i.e. the putative forcing variable, in this case the physical variables)
+      tempE= westend.bestE[col_names[j]]
+      temp <- ccm(interactions.block,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 27+j,
+                  lib_sizes = c(10,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
+      # mean rho at library size 500
+      rhomeans <- temp %>% ccm_means()
+      interactions.xmap_mat[i,j] <- rhomeans$rho[rhomeans$lib_size==500]
+
+      # first p-value (greater than zero? at library size 500)
+      p1 <- temp %>% filter(lib_size==500) %>% 
+        mutate(pos=ifelse(rho>0,1,0)) %>%
+        summarise(p=(1-sum(pos)/n()))
+      interactions.p1.mat[i,j] <- as.numeric(p1)
+
+      # second p-value (rho at lib-size 500 greater than rho at lib-size 10? By more than 0.1?)
+      p2 <- t.test(temp$rho[temp$lib_size==10],temp$rho[temp$lib_size==500])$p.value
+      if((rhomeans$rho[rhomeans$lib_size==500]-rhomeans$rho[rhomeans$lib_size==10])<0.1) p2 <- 0
+      interactions.p2.mat[i,j] <- as.numeric(p2)
+
+      # overall significance (both p1 and p2 signficant at alpha 0.05)
+      interactions.ptot.mat[i,j] <- ifelse(p1<0.05 & p2<0.05,1,0)
+   }
+}
+
+rm(n_col,n_row,col_names,row_names,p1,p2,tempE,temp,rhomean)
+
+# keep only signficant cross-mappings
+interactions.xmap_mat <- interactions.xmap_mat*interactions.ptot.mat %>% as.data.frame()
+interactions.xmap_rast <- interactions.xmap_mat %>% 
+  mutate(predictee=row.names(interactions.xmap_mat)) %>% 
+  gather(key=predictor,value=rho,-predictee)
+
+# If rho is zero, replace with NA (no significant causal signal)
+interactions.xmap_rast$rho[interactions.xmap_rast$rho==0] <- NA
+
+# Names for plotting to distinguish biological and physical variables
+interactions.xmap_rast <- interactions.xmap_rast %>% 
+  mutate(force.name=namekey$plotting[match(predictor,namekey$short)])
+
+## plot
+interactions.xmap_all_plot <- ggplot(interactions.xmap_rast,aes(x=force.name,y=predictee,fill=rho)) +
+  geom_raster() +
+  scale_fill_gradient(low = "#9AFF9A", high = "#548B54", space = "Lab", na.value = "grey50", 
+      guide = "colourbar",limits=c(0,0.8),breaks=c(0.2,0.4,0.6,0.8),name=expression(paste(rho, "(skill)"))) +
+  ggtitle("Interaction Convergent Cross Mapping") +
+  xlab("Predictor (Forcing Variable)") +
+  ylab("Predicted Variable") +
+  theme(text=element_text(color="black"),
+        plot.background = element_rect(fill="white"),
+        axis.text=element_text(color="black"),
+    axis.text.x=element_text(angle = 90, hjust = 1,vjust=0.1),
+    panel.border = element_blank())
+
+interactions.xmap_all_plot
+# ggsave("plots/ccm_ints_phys.png",plot=interactions.xmap_all_plot,width=6,height=5)
+```
+
 ## References

### 2018-07-17 EDM_SNI_analysis.Rmd
ints multi

@@
   # box plot of interactions
-  sppname <- namekey$long[match(species,namekey$short)]
+  sppname <- species
   int.box <- ggplot(coeffs.long,aes(x=plotting,y=value))+
@@
 
+### Multivariate S-maps with Interactions
+
+```{r multivariate smaps}
+ints.causal.vars <- purrr::map(interaction_names, function(x) {
+  interactions.xmap_rast %>%
+    filter(predictee==x,!is.na(rho))->out
+  out <- c(x,out$predictor)
+  out
+})
+
+names(ints.causal.vars)=interaction_names
+
+# apply s-maps
+ints.multi.mods <- purrr::map(interaction_names,function(x) {
+  smap_multi(interactions.block,westend.segs,x,ints.causal.vars[[x]])
+})
+names(ints.multi.mods) <- interaction_names
+
+```
+
 ## References

### 2018-07-18 EDM_SNI_analysis.Rmd
fixed physical normalization, added density plots, new network graph

--- 2018-07-14 EDM_SNI_analysis.Rmd
@@
 require(extrafont)
+require(GGally)
+require(sna)
+require(network)
 
@@
 ```{r ccm example}
-tempE <- westend.bestE['purp']
-temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 'mac',target_column = 'purp',lib_sizes = c(10,25,50,75,100,125,150,200,300,400,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
-
-mac_xmap_purp <- temp %>%
+tempE <- westend.bestE['red']
+temp <- ccm(westend,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 'mac',target_column = 'red',lib_sizes = c(10,25,50,75,100,125,150,200,300,400,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
+
+mac_xmap_red <- temp %>%
   group_by(lib_size)%>%
-  summarise(rhomean=mean(rho,na.rm=T),upper=rhomean+sd(rho,na.rm=T),lower=rhomean-sd(rho,na.rm=T))%>%
+  summarise(rhomean=mean(rho,na.rm=T),upper=quantile(rho, 0.975),lower=quantile(rho, 0.025))%>%
   ungroup()%>%
@@
   geom_line(color="darkorchid3")+
-  labs(x="Library Size",y=expression(paste(rho, " (predictive skill)")),title="Macrocystis xmap Purple Urchin")
-
-mac_xmap_purp
+  labs(x="Library Size",y=expression(paste(rho, " (predictive skill)")),title="Macrocystis xmap Red Urchin")
+
+mac_xmap_red
 # ggsave("plots/ccm_example.png",plot=mac_xmap_purp,width=6,height=4)
@@
       p2 <- t.test(temp$rho[temp$lib_size==10],temp$rho[temp$lib_size==500])$p.value
-      if((rhomeans$rho[rhomeans$lib_size==500]-rhomeans$rho[rhomeans$lib_size==10])<0.1) p2 <- 0
+      if((rhomeans$rho[rhomeans$lib_size==500]-rhomeans$rho[rhomeans$lib_size==10])<0.1) p2 <- 1
       westend.p2.mat[i,j] <- as.numeric(p2)
@@
 
-```{r igraph, echo=F, message=F,warning=F,fig.height=6}
-require(igraph) # R Package for network graphing and analysis
-
-# Add NAs for connections in to physical variables (since we want them just to be exogenous)
+```{r network,fig.height=6,fig.width=6}
 connect.mat <- t(westend.ptot.mat) %>% 
   as.data.frame() %>%
-  mutate(mei=rep(NA,nrow(.)),pdo=rep(NA,nrow(.)),npgo=rep(NA,nrow(.)),waves=rep(NA,nrow(.)),sst=rep(NA,nrow(.)))
-
-g <- graph.adjacency(as.matrix(connect.mat))
-par(mfrow=c(1,1))
-
-# Variable names
-
-names <- as.character(sapply(V(g)$name, function(x) namekey$plotting[match(x,namekey$short)]))
+  mutate(mei=rep(0,nrow(.)),pdo=rep(0,nrow(.)),npgo=rep(0,nrow(.)),waves=rep(0,nrow(.)),sst=rep(0,nrow(.)))%>%
+  as.matrix()
+names <- map_chr(colnames(connect.mat), function(x) as.character(namekey$plotting[match(x,namekey$short)]))
 names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
-
-# Vertex labels
-V(g)$name <-names
-vert_in <- sapply(degree(g,mode="in"),function(x) paste0("(",x,")"))
-vert_out_in <- paste0(degree(g,mode="out"),vert_in)
-V(g)$label <- paste(sep="\n",V(g)$name, vert_out_in)
-V(g)$label[7:11] <- paste(V(g)$name[7:11])
-
-#vertex coordinates
-crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))
-
-# vector of node colors for plotting
-vert.cols <- c(rep("tan3",3),rep("orchid3",2),"tan3",rep("gray70",5))
-plot(g,asp=0.8, vertex.label.color="black", vertex.label.cex=0.8, edge.arrow.size=0.6,vertex.label.font=2, vertex.shape="circle", edge.curved=F,margin=0, vertex.size=30,rescale=T,edge.lty=1,edge.color="gray40",layout=crds,vertex.color=vert.cols,vertex.frame.color=NA,main="")
+g <- network(connect.mat,vertex.attrnames = names,loops=F)
+g %v% "type" <- c(rep("darkgreen",3),rep("darkorchid1",2),"darkgreen",rep("dodgerblue",5))
+g %v% "labels" <- names
+int_net<-ggnet2(g,arrow.size=8,arrow.gap=0.05,color = "type")+
+  geom_point(aes(color = color), size = 12, alpha = 0.5) +
+  geom_point(aes(color = color), size = 9)+
+  geom_text(aes(label=label),family="Rockwell",size=4)+
+  theme(panel.border = element_rect(colour="white",fill=NA))
+
+ggsave("plots/int_net.png",plot=int_net)
+int_net
 ```
@@
   # box plot of interactions
-  sppname <- namekey$long[match(species,namekey$short)]
-  int.box <- ggplot(coeffs.long,aes(x=plotting,y=value))+
+  sppname <- species
+  int.box <- coeffs.long %>% filter(effect!='const')%>%
+    ggplot(aes(x=plotting,y=value))+
     geom_boxplot() +
@@
 
-all.coeffs.with.phys<- westend %>%
+spp.only.coeffs.with.phys<- westend %>%
   select(one_of(c(phys.vars,"period"))) %>%
   distinct() %>%
-  right_join(spp.only.mod.coeffs,by="period")%>%
+  right_join(filter(spp.only.mod.coeffs,!is.na(type)),by="period")%>%
   gather("external_forcing","ext_value",one_of(phys.vars))
@@
   plot_title <- namekey$long[match(spp,namekey$short)]
-  all.coeffs.with.phys %>%
+  spp.only.coeffs.with.phys %>%
   filter(!is.na(type),mod==spp)%>%
   ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
+    # geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
     geom_point(col='gray60',alpha=0.6)+
@@
 
-```{r some single plots}
-# macrocystis, pterygophora, and temperature
-quants<-c(0.05,0.5,0.95)
-all.coeffs.with.phys %>%
-  filter(type=="algal competition",mod%in% c('ymac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%
-
-  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-    geom_point(col='black',alpha=0.6)+
-    geom_hline(yintercept = 0,linetype=2,col="black")+
-    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-    scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
-    labs(x="Normalized SST",y="Interaction Strength")+
-    facet_grid(~mod,scales = "free_y")+
-    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
-
-# macrocystis, pterygophora, and NPGO
-all.coeffs.with.phys %>%
-  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="npgo")%>%
-  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+ 
-  geom_point(col='black',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="black")+
-   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
-   labs(x="Normalized NPGO",y="Interaction Strength")+
-   facet_grid(~mod,scales = "free_y")+
-  theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
-
-# macrocystis, pterygophora, herbivory, and NPGO
-all.coeffs.with.phys %>%
-  filter(type=="herbivory",mod%in% c('mac','pter'),effect %in% c('red','purp'),external_forcing=="npgo")%>%
-  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-    geom_point(col='black',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="black")+
-   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
-   labs(x="Normalized NPGO",y="Interaction Strength")+
-   facet_grid(~mod,scales = "free_y")+
-    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
-
-# macrocystis, pterygophora, and waves
-all.coeffs.with.phys %>%
-  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="waves")%>%
-  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
-  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
-  geom_point(col='black',alpha=0.6)+
-   geom_hline(yintercept = 0,linetype=2,col="black")+
-   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
-   labs(x="Normalized SWH",y="Interaction Strength")+
-   facet_grid(~mod,scales = "free_y")+
-    theme(strip.text=element_blank(),
-          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))
-
-```
-
 Here are some notable effects/trends for each species:
@@
 
-### Effect of disturbance on *Macrocystis*
-```{r disturbance and mac,fig.height=7,fig.width=5}
-quants <- c(0.05,0.5,0.95)
-waves.mac.pter <- westend %>%
-  select(one_of(c("pter","lam","mac","period"))) %>%
-  distinct() %>%
-  right_join(all.mod.coeffs.long,by="period")%>%
-  gather("external_spp","ext_value",one_of(c('pter','lam','mac')))%>%
-  filter(effect%in% c('waves','pter','lam'),mod=='mac')%>%
-  ggplot(aes(ext_value,value))+
-  geom_point(col='gray60',alpha=0.6)+
-  geom_hline(yintercept = 0,linetype=2)+
-  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-  labs(x="Normalized Density",y="Effect of SWH on Macrocystis")+
-  facet_grid(external_spp~effect,scales="free")
-waves.mac.pter
-```
-
-### Effect of Macrocystis on Laminaria/Pterygophora interaction
-```{r disturbance and mac,fig.height=7,fig.width=5}
-quants <- c(0.05,0.5,0.95)
-mac.lam.pter <- westend %>%
-  select(one_of(c("mac","period"))) %>%
-  distinct() %>%
-  right_join(all.mod.coeffs.long,by="period")%>%
-  gather("external_spp","ext_value",mac)%>%
-  filter(effect=='pter',mod=='lam')%>%
-  ggplot(aes(ext_value,value))+
-  geom_point(col='gray60',alpha=0.6)+
-  geom_hline(yintercept = 0,linetype=2)+
-  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-  labs(x="Normalized Density",y="Effect of Pterygophora on Laminaria")
-mac.lam.pter
-```
-### Effect of urchin density on Macrocystis/Pterygophora interaction
-
-```{r urchin dens mac pter}
-quants <- c(0.05,0.5,0.95)
-urchin.mac.pter <- westend %>%
-  select(one_of(c("red","period"))) %>%
-  distinct() %>%
-  mutate(time=row_number())%>%
-  right_join(all.mod.coeffs.long,by=c("period",'time'))%>%
-  gather("external_spp","ext_value",red)%>%
-  filter(effect %in% c('pter','mac'),mod %in% c('pter','mac'),!is.na(ext_value))%>%
-  ggplot(aes(ext_value,value))+
-  geom_point(col='gray60',alpha=0.6)+
-  geom_hline(yintercept = 0,linetype=2)+
-  geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
-  labs(x="Normalized Red Urchin Density",y="Interaction Strength")+
-  facet_wrap(~mod)
-urchin.mac.pter
-```
-
-### Do conditions affect average interaction strength?
-
-```{r conditions and interaction strength}
-# disturbance. top quartile vs. bottom quartile
-all.coeffs.with.phys %>%
-  mutate(av_value=abs(value))%>%
-  filter(external_forcing=='waves', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
-  mutate(stress_level=ifelse(ext_value>=quantile(westend$pdo,0.75,na.rm=T),'high','low'))%>%
-  ggplot(aes(log(value),..density..,fill=stress_level))+
-  geom_density(alpha=0.6)+
-  geom_vline(xintercept = 0,linetype=2,color='black')+
-  labs(x='Interaction Strength',y='Probability Density')
+### Example: Effect of Macrocystis on Pterygophora
+
+```{r mac pter}
+mac_effect_on_pter <-spp.only.coeffs.with.phys %>%
+  ungroup()%>%
+  filter(mod=="pter",effect=="mac")%>%
+  mutate(phys_high_low=case_when(
+    ext_value > 1 ~ "high",
+    ext_value < -1 ~ "low")) %>%
+  filter(!is.na(phys_high_low))%>%
+  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
+  group_by(type,external_forcing,phys_high_low)%>%
+  mutate(mean_int=mean(value,na.rm=T))%>%
+  distinct()%>%
+  ungroup()
+
+mac_effect_on_pter %>%
+ggplot(aes(value,..density..,fill=phys_high_low))+
+  geom_density(col=NA,alpha=0.6)+
+  geom_vline(xintercept=0,linetype=2)+
+  geom_vline(aes(xintercept=mean_int,color=phys_high_low))+
+  scale_fill_manual(values=c("red","blue"),name="Physical\nForcing Value")+
+  scale_color_manual(values=c("red","blue"))+
+  guides(color="none")+
+  labs(x="Effect of Macrocystis on Pterygophora",y="Probability Density")+
+  facet_grid(rows=vars(external_forcing))
+
+```
+
+### Distribution of interaction strengths by type
+
+```{r interactions density plot by type,fig.height=6}
+# Density plot
+interaction_distribution_by_type <- spp.only.mod.coeffs %>%
+  filter(!is.na(type))%>%
+  group_by(type) %>%
+  mutate(mean_int=mean(value,na.rm=T))%>%
+  ungroup()%>%
+  ggplot(aes(value,..density..))+
+  geom_density(aes(fill=type),alpha=0.8,col=NA)+
+  geom_vline(xintercept=0,linetype=2)+
+  geom_vline(aes(xintercept=mean_int))+
+  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+   guides(fill=F)+
+  xlim(-2,2)+
+  labs(x="Interaction strength",y="Kernel density")+
+  geom_text(aes(x=-2,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
+  facet_grid(rows=vars(type))+
+  theme(strip.text = element_blank())
+
+interaction_distribution_by_type
+ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6)
+```
+
+```{r ints dist by type and phys}
+int_dist_by_type_phys <- spp.only.coeffs.with.phys %>%
+  ungroup()%>%
+  mutate(phys_high_low=case_when(
+    ext_value > 1 ~ "high",
+    ext_value < -1 ~ "low")) %>%
+  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
+  group_by(type,external_forcing,phys_high_low)%>%
+  mutate(mean_int=mean(value,na.rm=T))%>%
+  distinct()%>%
+  ungroup()
+
+int_dist_by_type_phys_plots <- purrr::map(phys.vars,function(x) {
+  int_dist_by_type_phys %>%
+    filter(external_forcing==x,!is.na(phys_high_low))%>%
+    ggplot(aes(value,..density..,fill=phys_high_low,col=phys_high_low))+
+    geom_density(alpha=0.6,col=NA)+
+    geom_vline(xintercept=0,linetype=2)+
+    geom_vline(aes(xintercept=mean_int,col=phys_high_low))+
+    xlim(-2,2)+
+    scale_fill_manual(values=c("red","blue"),name=paste0(x,"\nLevel"),labels=c("high (> 1)","low (< 1)"))+
+    scale_color_manual(values=c('red','blue'))+
+    guides(color="none")+
+    geom_text(aes(x=-2,y=0.5,label=type),col="black",hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
+    labs(x="Interaction strength",y="Kernel density")+
+    facet_grid(rows=vars(type))+
+    theme(strip.text=element_blank())
+  })
+names(int_dist_by_type_phys_plots) <- phys.vars
+int_dist_by_type_phys_plots$mei
+int_dist_by_type_phys_plots$sst
+int_dist_by_type_phys_plots$pdo
+int_dist_by_type_phys_plots$waves
+int_dist_by_type_phys_plots$npgo
 
@@
   left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
-  distinct()
+  distinct()%>%
+  ungroup()%>%
+  # normalize
+  mutate_at(vars(lam_mac:ymac_red),function(dens) {(dens-mean(dens,na.rm=T))/sd(dens,na.rm=T)})
+```
+
+```{r ints time series plot}
+spp.only.mod.coeffs %>%
+  filter(time %in% 1:63,type=="algal competition")%>%
+  unite("interaction",effect,mod,sep="_")%>%
+  ggplot(aes(time,value,color=interaction))+
+  geom_line()+
+  geom_point()+
+  labs(x="period",y="interaction strength")
 ```
@@
 # Run simplex projection for each species interaction
-interaction_names <- colnames(interactions.block)[3:29]
+interaction_names <- colnames(interactions.block)[3:22]
 interactions.simp.list <- list()
@@
   temp$E[temp$rho==max(temp$rho)]
+})
+purrr::map_chr(interactions.simp.list,function(x) {
+  max(x$rho,na.rm=T)
 })
@@
       tempE= westend.bestE[col_names[j]]
-      temp <- ccm(interactions.block,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 27+j,
+      temp <- ccm(interactions.block,lib=as.matrix(westend.segs),pred=as.matrix(westend.segs),E=tempE,lib_column= 2+i,target_column = 2+length(interaction_names)+j,
                   lib_sizes = c(10,500),num_samples=100,replace=T,silent=T,RNGseed = 41389)
@@
   scale_fill_gradient(low = "#9AFF9A", high = "#548B54", space = "Lab", na.value = "grey50", 
-      guide = "colourbar",limits=c(0,0.8),breaks=c(0.2,0.4,0.6,0.8),name=expression(paste(rho, "(skill)"))) +
+      guide = "colourbar",limits=c(0,0.75),breaks=c(0.25,0.5,0.75),name=expression(paste(rho, "(skill)"))) +
   ggtitle("Interaction Convergent Cross Mapping") +
@@
 
+### Multivariate S-maps with Interactions
+
+```{r multivariate smaps}
+ints.causal.vars <- purrr::map(interaction_names, function(x) {
+  interactions.xmap_rast %>%
+    filter(predictee==x,!is.na(rho))->out
+  out <- out$predictor
+  out
+})
+
+names(ints.causal.vars)=interaction_names
+
+# apply s-maps
+ints.multi.mods <- purrr::map(interaction_names,function(x) {
+  if(length(ints.causal.vars[[x]])==0) return(NULL)
+  else(smap_multi(interactions.block,westend.segs,x,ints.causal.vars[[x]]))
+})
+names(ints.multi.mods) <- interaction_names
+
+purrr::map_dbl(ints.multi.mods, function(x) {
+x$model$rho
+})
+
+```
+
 ## References

### 2018-07-18 data/data preparation scripts/physical_data_preparation.R

@@ -25,7 +25,7 @@ library(lubridate)
 
 #### Period matching key ####
 # Have to line up the periods such that December is included in the next Spring's period
-period.key <- data_frame(year=rep(1980:2014,each=12),month=rep(1:12,35),period=c(rep(NA,5),rep(1,6),rep(2:69,each=6),NA))
+period.key <- data_frame(year=rep(1980:2011,each=12),month=rep(1:12,32),period=c(rep(NA,5),rep(1,6),rep(2:63,each=6),NA))
 
 # Set months that are not in Dec-Mar or Jun-Sep as NA so their data aren't included
 period.key$period[!period.key$month %in% c(12,1,2,3,6,7,8,9)] <- NA
@@ -144,14 +144,14 @@ source("data/data preparation scripts/wave_height_Bell.R")
 
 # Join the two datasets together, using the normalized buoy data to "fill in" the modeled data
 wavesboth <- wavesdf %>%
-  mutate(waves.norm=(maxHs-mean(maxHs,na.rm=T))/sd(maxHs,na.rm=T))%>%
   # Use the periods key to assign periods to each month, and remove irrelevant data
   left_join(period.key,by=c("year"="year","month"="month")) %>%
   filter(!is.na(period)) %>%
   
-  # For each period, average the SST value to make just one value for each period
+  # For each period, average the SWH value to make just one value for each period
   group_by(period) %>%
-  summarise(waves.norm=max(waves.norm,na.rm=T)) %>%
+  summarise(waves=max(maxHs,na.rm=T)) %>%
+  mutate(waves.norm=(waves-mean(waves,na.rm = T))/sd(waves,na.rm=T))%>%
   ungroup()
 
 #### Sea Surface Temperature ####

### 2018-07-19 EDM_SNI_analysis.Rmd
ported manuscript to markdown

@@
 require(GGally)
+require(ggalt)
 require(sna)
@@
 ```{r network,fig.height=6,fig.width=6}
+# g <- network(connect.mat,vertex.attrnames = names,loops=F)
+# g %v% "type" <- c(rep("darkgreen",3),rep("darkorchid1",2),"darkgreen",rep("dodgerblue",5))
+# g %v% "labels" <- names
+# set.edge.value(g,"weights",edge.weight.mat*3)
+# 
+# int_net<-ggnet2(g,arrow.size=6,arrow.gap=0.04,color = "type",label=names,label.alpha=0,edge.size="weights")+
+#   geom_point(aes(color = color), size = 14, alpha = 0.5) +
+#   geom_point(aes(color = color), size = 12)+
+#   geom_text(aes(label=label),colour='white',family="Rockwell",size=3)+
+#   theme(panel.border = element_rect(colour="white",fill=NA))
+# 
+# # ggsave("plots/int_net.pdf",plot=int_net)
+# int_net
+
+## i graph
+require(igraph)
+
+# connections matrix for the network based on the CCM results above
 connect.mat <- t(westend.ptot.mat) %>% 
@@
   as.matrix()
+
+# edges (arrows) weighted based on CCM rho value
+edge.weight.mat <- t(westend.xmap_mat)%>%
+  as.data.frame() %>%
+  mutate(mei=rep(0,nrow(.)),pdo=rep(0,nrow(.)),npgo=rep(0,nrow(.)),waves=rep(0,nrow(.)),sst=rep(0,nrow(.)))%>%
+  as.matrix()
+edge.weight.mat[is.na(edge.weight.mat)]<-0
+edge.weight.mat
+
 names <- map_chr(colnames(connect.mat), function(x) as.character(namekey$plotting[match(x,namekey$short)]))
+names[6] <- "M. pyr\n(j)"
 names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
-g <- network(connect.mat,vertex.attrnames = names,loops=F)
-g %v% "type" <- c(rep("darkgreen",3),rep("darkorchid1",2),"darkgreen",rep("dodgerblue",5))
-g %v% "labels" <- names
-int_net<-ggnet2(g,arrow.size=8,arrow.gap=0.05,color = "type")+
-  geom_point(aes(color = color), size = 12, alpha = 0.5) +
-  geom_point(aes(color = color), size = 9)+
-  geom_text(aes(label=label),family="Rockwell",size=4)+
-  theme(panel.border = element_rect(colour="white",fill=NA))
-
-ggsave("plots/int_net.png",plot=int_net)
-int_net
+
+# initiate the igraph
+g2 <- graph.adjacency(connect.mat)
+
+# Vertex labels
+V(g2)$name <-names
+
+# label with the number of incoming(outgoing) connections
+vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
+vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
+V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
+V(g2)$label[7:11] <- paste(V(g2)$name[7:11])
+
+# #vertex coordinates (layout of plot
+# crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))
+
+# vector of node colors for plotting
+vert.cols <- c(rep("#238B45",3),rep("#BEBADA",2),"#238B45",rep("#8DD3C7",5))
+
+# weighted edges, with width and color
+edge.weights <- as.numeric(t(edge.weight.mat))[t(edge.weight.mat) != 0]*5
+E(g2)$width <- edge.weights
+col.1 <- adjustcolor("#99D8C9", alpha=0.8)
+col.2 <- adjustcolor("#006D2C", alpha=0.8)
+edge.pal <- colorRampPalette(c(col.1, col.2), alpha = TRUE)
+edge.pal <- edge.pal(100)
+edge.cols<-round((edge.weights-min(edge.weights))/(max(edge.weights)-min(edge.weights))*100) +1
+edge.cols <- edge.pal[edge.cols]
+
+# some edges must be curved to look okay in the graph
+curves <- rep(0,40)
+curves[c(1,4,13,17,18,21)] <-0.4
+curves[c(10,19)] <-0.3
+curves[20] <- -0.1
+curves[c(6,39)] <- 0.2
+
+#plot to file if uncommented
+# png("plots/ccm_network.png",height=600,width=600)
+plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
+# dev.off()
 ```
@@
 
-```{r 3d plot,fig.height=7,fig.width=7}
+```{r 3d plot,fig.height=4,fig.width=4,dpi=300}
 require(plot3D)
@@
 scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
-
-dev.copy(pdf,"plots/multi_attractor.pdf")
-dev.off()
+# 
+# dev.copy(png,"plots/multi_attractor.png",width=700,height=700,res=96)
 
@@
 rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)
-
 ```
@@
 })
-ggsave("plots/phys_effects_red.png",plot=phys_effects_on_interactions[[1]])
-ggsave("plots/phys_effects_purp.png",plot=phys_effects_on_interactions[[2]])
-ggsave("plots/phys_effects_lam.png",plot=phys_effects_on_interactions[[3]])
-ggsave("plots/phys_effects_pter.png",plot=phys_effects_on_interactions[[4]])
-ggsave("plots/phys_effects_mac.png",plot=phys_effects_on_interactions[[5]])
-ggsave("plots/phys_effects_ymac.png",plot=phys_effects_on_interactions[[6]])
+# ggsave("plots/phys_effects_red.png",plot=phys_effects_on_interactions[[1]])
+# ggsave("plots/phys_effects_purp.png",plot=phys_effects_on_interactions[[2]])
+# ggsave("plots/phys_effects_lam.png",plot=phys_effects_on_interactions[[3]])
+# ggsave("plots/phys_effects_pter.png",plot=phys_effects_on_interactions[[4]])
+# ggsave("plots/phys_effects_mac.png",plot=phys_effects_on_interactions[[5]])
+# ggsave("plots/phys_effects_ymac.png",plot=phys_effects_on_interactions[[6]])
 
@@
 
-```{r mac pter}
+```{r mac pter,fig.width=6,fig.height=6}
 mac_effect_on_pter <-spp.only.coeffs.with.phys %>%
@@
 
-mac_effect_on_pter %>%
+# physical forcings labeller
+phys.labeller <- c(
+  mei= "MEI",
+  npgo= "NPGO",
+  pdo= "PDO",
+  sst="SST",
+  waves="SWH"
+)
+
+mac_effect_on_pter_plot <-mac_effect_on_pter %>%
 ggplot(aes(value,..density..,fill=phys_high_low))+
@@
   geom_vline(xintercept=0,linetype=2)+
-  geom_vline(aes(xintercept=mean_int,color=phys_high_low))+
-  scale_fill_manual(values=c("red","blue"),name="Physical\nForcing Value")+
-  scale_color_manual(values=c("red","blue"))+
+  geom_vline(aes(xintercept=mean_int,color=phys_high_low),size=1.5)+
+  scale_fill_manual(values=c("darkorange","dodgerblue4"),name="Physical\nForcing Value",labels=c("high","low"))+
+  scale_color_manual(values=c("darkorange","dodgerblue4"))+
   guides(color="none")+
   labs(x="Effect of Macrocystis on Pterygophora",y="Probability Density")+
-  facet_grid(rows=vars(external_forcing))
-
+  facet_grid(rows=vars(external_forcing),labeller=labeller(external_forcing=phys.labeller))+
+  theme(strip.text=element_text(size=10))
+
+# ggsave("plots/mac_effect_on_pter_phys.png",plot=mac_effect_on_pter_plot,h=6,w=6,dpi=500)
+mac_effect_on_pter_plot
 ```
@@
 
-```{r interactions density plot by type,fig.height=6}
+```{r interactions density plot by type,fig.height=6,fig.width=6}
 # Density plot
@@
    guides(fill=F)+
-  xlim(-2,2)+
+  xlim(-1.5,1.5)+
   labs(x="Interaction strength",y="Kernel density")+
-  geom_text(aes(x=-2,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
+  geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
   facet_grid(rows=vars(type))+
@@
 interaction_distribution_by_type
-ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6)
+# ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6,w=6,dpi=500)
 ```
@@
   group_by(mod,effect,type) %>%
-  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),cv_int=sd_int/abs_mean*100,
+  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
+            cv_int=sd_int/abs_mean*100,upper=quantile(value,0.975),lower=quantile(value,0.025),
             perc_neutral=sum(neutral)/n()*100)%>%
@@
 # Mean interaction strength by interaction type
-int_mean_sd %>%
+spp.only.mod.coeffs %>%
+  filter(!is.na(type))%>%
+  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
   filter(mod != "ymac",effect != 'ymac')%>%
-  group_by(type)%>%
-  summarise(mean_str=mean(abs_mean),ci95=sd(abs_mean)/sqrt(n())*1.96)%>%
+  group_by(type) %>%
+  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
   ungroup()%>%
-  ggplot(aes(type,mean_str))+
+  ggplot(aes(type,mean_int))+
   geom_col()+
-  geom_errorbar(aes(ymax=mean_str+ci95,ymin=mean_str-ci95),position='dodge',width=0.2)+
+  geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
   labs(x='Interaction Type',y='Interaction Strength')
@@
 # scatterplot mean vs. sd (removing intraspecies interactions)
-int_mean_sd %>%
-  filter(mod != "ymac",effect != 'ymac')%>%
-  ggplot(aes(abs_mean,sd_int))+
+int_mean_sd_plot <- int_mean_sd %>%
+  filter(type != "intraspecies")%>%
+  ggplot(aes(mean_int,sd_int))+
   geom_point(aes(color=type),size=5)+
+  geom_smooth(method="lm",se=F,col="black",show.legend = F)+
+  geom_vline(xintercept=0,linetype=2)+
+  # geom_encircle(aes(fill=type),col=NA,alpha=0.3,expand=0.1)+
   scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
-  geom_smooth(method="lm",se=F,col="black")+
-  labs(x="Mean Interaction Strength",y="SD Interaction Strength")
-
-# CV calculated using the correction for sample size (cv*=1+1/4n * cv)
-int_mean_sd %>%
-  filter(mod != 'ymac',effect!='ymac')%>%
+  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  guides(fill='none')+
+  labs(x="Mean Interaction Strength",y="SD Interaction Strength")+
+  theme(legend.position = c(0.2,0.8))
+# ggsave("plots/int_mean_sd.png",plot=int_mean_sd_plot,dpi=500)
+int_mean_sd_plot
+
+# CV
+spp.only.mod.coeffs %>%
+  filter(!is.na(type))%>%
+  filter(type != "intraspecies")%>%
+  group_by(type) %>%
+  summarise(abs_mean=mean(abs(value)),sd_int=sd(value,na.rm=T),cv_int=abs_mean/sd_int*100)%>%
+  ungroup()%>%
   ggplot(aes(type,cv_int))+
   geom_col()+
-
   labs(x="Interaction Type",y="CV of Measured Interactions")
@@
   mutate(rev_effect=mod,rev_mod=effect) %>%
-  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))%>%
-  mutate(upper1=mean_int.x+sd_int.x,lower1=mean_int.x-sd_int.x,upper2=mean_int.y+sd_int.y,lower2=mean_int.y-sd_int.y)-> int_mean_sd2
+  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))-> int_mean_sd2
 
@@
   geom_abline(slope=1,intercept=0,linetype=2)+
-  geom_errorbar(aes(ymin=lower2,ymax=upper2))+
-  geom_errorbarh(aes(xmin=lower1,xmax=upper1))+
+  geom_errorbar(aes(ymin=lower.y,ymax=upper.y))+
+  geom_errorbarh(aes(xmin=lower.x,xmax=upper.x))+
   labs(x='Effect of A on B',y='Effect of B on A')

### 2018-07-19 manuscript.Rmd
ported manuscript to markdown

---
title: "Environmental Context Dependency in Species Interactions"
author: "Owen Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
fontsize: 12pt
bibliography: citation_bibtex.bib
csl: nature.csl
editor_options: 
 chunk_output_type: inline
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \renewcommand\linenumberfont{\normalfont\bfseries\small}
---

```{r global_options, include=F,message=F}
require(knitr)
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message = FALSE,results = 'asis',dpi=96)
```

**Primary target journal**: Nature Ecology and Evolution

**Journal Guidelines**:

* Maximum 3500 words main text (excluding introductory paragraph/abstract)
* Maximum 6 display items (figures/tables)
* The introductory paragraph is typically 150 words and is unreferenced; it contains a brief account of the background and rationale of the work, followed by a statement of the main conclusions introduced by the phrase "Here we show" or some equivalent.An introduction (without heading) of up to 500 words of referenced text expands on the background of the work (some overlap with the summary is acceptable), and is followed by a concise, focused account of the findings (headed 'Results'), and one or two short paragraphs of discussion (headed 'Discussion').

**Work confirmed/supported/debates**:

* Tegner et al 1997, Dayton et al 1999
* Bottom-up vs. top-down forcing? Disturbance or herbivory?
* Competition between urchin species?
* Physical drivers increase predictability of algae models!

**Suggested future research**

* Predominance of mutualism- why the positive interactions of Pterygophora and Macrocystis?
* Urchin grazing pressure as most important??
* Multiple stressors (e.g. herbivory, sst, wave height)

**FIGURES**

Time series of study species
Example attractor?
CCM results
Empirical interaction web, in “normal”, “high disturbance”, and “high nutrient stress” environments??

***

## Introductory paragraph (abstract)

Ecological interactions are not uniform across time, and instead vary with environmental conditions. Interactions among species are often measured with short-term controlled experiments, but these experiments are subject to the particular environmental conditions under which they are performed. As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interaction strengths across a wide range of environmental conditions in a coastal marine ecosystem. By including large-scale climate indices, sea surface temperature, and a measure of physical disturbance in analyses, we show that environmental context influences the strength and direction of species interactions. In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied interactions. The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of ecosystem dynamics.

## Introduction

Interactions between species drive patterns of diversity, stability, resilience, and productivity in nature [@Connell1978; @Suttle2007a; @Tilman1999d; @Berlow1999b]. In any ecosystem, the collection of species interactions determines community dynamics. Until recently, most studies viewed these dynamics---e.g., the bleaching and recovery of a coral reef, or the assembly and disassembly of terrestrial plant communities---as processes resulting from static, predictable species interactions. However, the observation that species interactions are not spatiotemporally uniform [@Odadi2011; @Pennings2005; @Menge1994; @Chamberlain2014] calls into question assumptions of interaction stability.

Ecologists recognize now that important species interactions may vary over time, but this context dependency remains difficult to measure and describe. Experiments that measure interactions are generally performed over a limited spatiotemporal range, and are therefore subject to a specific environmental context that may not encompass the range of conditions experienced by that ecosystem over longer time scales [@Vaughn2010]. This is worrying, since environmental context can profoundly influence the outcome of species interactions ranging anywhere from keystone predation [@Menge1994] to competition [@Lima2008; @Holzapfel1999a; @Odadi2011], to protective symbioses [@Daskin2012; @Glynn1996; @White2017a].

Moreover, the focus of the search for context dependency has been on mean interaction strengths, at the expense of specific examinations of interaction variance [@Chamberlain2014]. This focus may be misguided, as it has been shown that interactions that are variable in magnitude and direction—and therefore “weak” when averaged—may actually be some of the most important in driving community dynamics [@Berlow1999b]. If key species interactions are variable in this way across environmental gradients, then many studies may be attributing important ecological phenomena to observational noise.

A solution to these difficulties is to a) utilize ecological observations collected over a long time period, across a large range of environmental contexts, with b) an analytical method to directly estimate context-dependent species interactions from those observations. Such an approach could help to characterize environmental contingencies in species interactions and explicitly examine interaction variability. Here, we use empirical dynamic modelling (EDM, [@Deyle2016]) to estimate a varying species interaction network and establish environmental context dependency in interaction strength and direction. Empirical dynamic modelling uses information from single or multiple time series to empirically model relationships between variables through the reconstruction of dynamic attractors (https://youtu.be/8DikuwwPWsY). The general modelling framework for all EDM methods is readily adaptable to many different sorts of time series variables, including environmental variables manifesting at different scales [@Deyle2013; @Chang2017; @Ye2015]. Because the methods are specifically designed for nonlinear dynamic systems, EDM---in theory---should be able to illuminate context-dependent patterns in species interactions.

Recently-developed EDM methods exist for uncovering dynamic species interactions from time series data [@Deyle2016], but these methods have insofar been applied only to simulated and planktonic communities, and their utility to the study of other ecological systems remains untested. We focus here instead on giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented [@Foster1985a; @Graham2007c; @Teagle2017]. The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011]. Recently, however, findings from long-term kelp forest research programs have begun to challenge many long-held beliefs about the drivers of kelp forest ecosystem dynamics [@Reed2016]. In particular, a longer-term perspective has led to a recognition of the critical importance of environmental context—such as level of physical disturbance or the current state of El Niño conditions—for understanding kelp forest processes [@Parnell2010; @Byrnes2011; @Bell2015; @Young2015].

In this study, we utilize monitoring data from one such effort at San Nicolas Island, a small, remote member of the California Channel Islands in the northeast Pacific [@Kenner2013]. Our analyses focus on the dynamics of five common southern California kelp forest species, whose interactions are thought to be important in structuring kelp forest ecosystems [@Foster1985a; @Foster1990a; @Teagle2017]. The giant kelp *Macrocystis pyrifera* is the eponymous foundation species, the primary canopy- and habitat-forming kelp along most of the central and southern coast of California [@Graham2007c]. We explore its dynamics and interactions with two presumptive competitors and two abundant herbivores. The understory kelp species *Laminaria farlowii* and *Pterygophora californica* compete with Macrocystis for space, light, and nutrients [@Dayton1984; @Dayton1999; @Reed1984]. The two herbivores---the purple sea urchin *Strongylocentrotus purpuratus* and the red sea urchin *Mesocentrotus franciscanus*---are thought in many places to control Macrocystis density and can sometimes wipe out entire giant kelp forests, leading to the alternative ecosystem state known as an urchin barren [@Ling2014].

To characterize environmental context dependency in kelp forest interactions between these five species, we take three general steps (see Methods). First, we use empirical dynamic modeling causality tests called convergent cross-mapping [@Sugihara2012] to construct a kelp forest species interaction network directly from time series data. Even though these common kelp forest species are all thought to interact, we tested that assumption with convergent cross mapping, which can separate correlation from causation [@Sugihara2012].

Second, we use multivariate S-maps (sequential locally weighted global linear maps) [@Deyle2016; @Sugihara1994] to reconstruct the interactions between species for all of the causal links established in the previous step. S-maps reconstructs dynamic “attractors” by casting the abundances of causally-related species into state space. That is, for a set of causally-related species $A$, $B$, and $C$, a point in multivariate space $\{A_t, B_t, C_t\}$ can be plotted, using each species’ abundance at time t. The attractor is then created by tracing this multivariate trajectory forward in time for all t (FIG?). S-maps then computes sequential Jacobian matrices for each point along the attractor, where the elements of each matrix are the partial derivatives between species. These estimated partial derivatives are our measure of species interactions.

The last step is to investigate if and how species interactions varies with environmental context. Because multivariate S-maps are computed for every point along reconstructed attractors, we can extract distributions of interactions that can be related back to environmental context. We explore patterns of species interactions in relation to sea surface temperature, physical disturbance (measured by wave height), and three indices of low-frequency climate modes: the Multivariate ENSO Index, the Pacific Decadal Oscillation, and the North Pacific Gyre Oscillation [@Wolter2011; @Mantua2002; @DiLorenzo2008].

## Results and discussion

```{r fig.width=4, fig.height=4,echo=FALSE,fig.cap="Example reconstructed dynamic attractor"}
library(png)
library(grid)
include_graphics("plots/multi_attractor.pdf")
```

```{r fig.width=4, fig.height=4,echo=FALSE,fig.cap="Reconstructed interaction web using results of convergent cross mapping"}
include_graphics("plots/ccm_network.pdf")
```

```{r echo=FALSE,fig.cap="Distribution of estimated interaction by type"}
img <- readPNG("plots/int_dist_by_type_density.png")
grid.raster(img)
```

```{r echo=FALSE, fig.cap="Mean and standard deviation of each estimated unidirectional species interaction"}
img <- readPNG("plots/int_mean_sd.png")
grid.raster(img)
```

```{r echo=FALSE,fig.cap="Macrocystis effect on Pterygophora"}
img <- readPNG("plots/mac_effect_on_pter_phys.png")
grid.raster(img)
```

* CCM results and variable interaction network- centrality of Macrocystis
* Mean vs. variance in certain algal interactions?
* Bidirectional mean interactions (across two dimensions, A on B and B on A)?
* Frequency of positive vs. negative interactions? Across environmental gradients?

## References

### 2018-07-20 EDM_SNI_analysis.Rmd
updated manuscript and added .csl

@@
 require(network)
+require(gridExtra)
+require(RANN)
 
@@
 fra","M.
-pyr (j)","(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST"))
+pyr (j)","(P) MEI","(P) PDO","(P) NPGO","(P) SWH","(P) SST"))
 
@@
   # define column "plotting" as a factor so we can later control plotting order on figures
-  mutate(plotting=factor(plotting,levels=c("(P) MEI","(P) PDO","(P) NPGO","(P) Max Hs","(P) SST","M. pyr","M. pyr (j)","L. far","P. cal","S. pur","M. fra","Constant")),
+  mutate(plotting=factor(plotting,levels=c("(P) MEI","(P) PDO","(P) NPGO","(P) SWH","(P) SST","M. pyr","M. pyr (j)","L. far","P. cal","S. pur","M. fra","Constant")),
          long=factor(long,levels=c("Macrocystis >1m","Macrocystis <1m","Pterygophora","Laminaria","Purple urchin","Red urchin","Multivariate ENSO Index","Pacific Decadal Oscillation","North Pacific Gyre Oscillation","Significant Wave Height","Sea Surface Temperature","Constant")))
@@
       theme(text = element_text(color="black",size=16),
+            axis.title=element_text(size=10),
             legend.position = c(0.15,0.7),
@@
 rm(pd,westend.long,westend.meandens)
-
-```
-
-```{r physical variables timeseries}
+# ggsave("plots/normalized_time_series.pdf",h=5,w=7)
+```
+
+```{r physical variables timeseries,fig.height=9,fig.width=7}
 mei.pdo.npgo.ts.plot <-westend %>%
@@
   labs(title="Physical Drivers: Climate Modes",x="Year",y="Normalized Index Value")+
-  plot_theme_black
+  theme(axis.title=element_text(size=10))
 sst.waves.ts.plot <-westend %>%
@@
   labs(title="Physical Drivers: Temperature and Disturbance",x="Year",y="Normalized Index Value")+
-  plot_theme_black
+  theme(axis.title=element_text(size=10))
 mei.pdo.npgo.ts.plot
@@
 # ggsave("plots/sst_waves_ts_plot.png",plot=sst.waves.ts.plot,width=7,height=4)
-
+lay <- rbind(1,2,3,3)
+norm_ts<-grid.arrange(mei.pdo.npgo.ts.plot,sst.waves.ts.plot,westend.mean_dens_plot,layout_matrix=lay)
+norm_ts
+# ggsave("plots/normalized_time_series.pdf",plot=norm_ts,h=9,w=7)
+rm(lay)
 ```
@@
 ```{r time series segments, include=F}
+
 # segments of the time series (which rows of the data are individual transects?)
@@
 }
-rm(n_col,n_row,col_names,row_names,p1,p2,tempE,temp,rhomean)
+rm(n_col,n_row,col_names,row_names,p1,p2,tempE,temp,rhomeans)
 
@@
       guide = "colourbar",limits=c(0,0.8),breaks=c(0.2,0.4,0.6,0.8),name=expression(paste(rho, "(skill)"))) +
+  geom_text(aes(label=round(rho,2)),family="Rockwell")+
   ggtitle("Kelp Forest Convergent Cross Mapping") +
@@
 names[6] <- "M. pyr\n(j)"
-names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
+names [7:11] <- c("MEI","PDO","NPGO","SWH","SST")
 
@@
 #plot to file if uncommented
-# png("plots/ccm_network.png",height=600,width=600)
+# pdf("plots/ccm_network.pdf",height=7,width=7)
 plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
@@
 #manually find nearest neighbors to point 46
-require(RANN)
 temp <- westend %>% select(lam,mac,purp) %>% slice(1:315) %>% mutate(id=row_number()) %>%
@@
 # ggsave("plots/mac_effect_on_pter_phys.png",plot=mac_effect_on_pter_plot,h=6,w=6,dpi=500)
+# ggsave("plots/mac_effect_on_pter_phys.pdf",plot=mac_effect_on_pter_plot,h=6,w=6)
 mac_effect_on_pter_plot
@@
 interaction_distribution_by_type <- spp.only.mod.coeffs %>%
-  filter(!is.na(type))%>%
+  filter(!is.na(type),type != "intraspecies")%>%
   group_by(type) %>%
@@
 # ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6,w=6,dpi=500)
+# ggsave("plots/int_dist_by_type_density.pdf",plot=interaction_distribution_by_type,h=6,w=6)
 ```
@@
 
-# scatterplot mean vs. sd (removing intraspecies interactions)
+# scatterplot mean vs. sd (removing intraspecies interactions and young macrocystis)
 int_mean_sd_plot <- int_mean_sd %>%
-  filter(type != "intraspecies")%>%
+  filter(type != "intraspecies", effect != "ymac", mod != "ymac")%>%
   ggplot(aes(mean_int,sd_int))+
-  geom_point(aes(color=type),size=5)+
+  geom_point(aes(color=type,shape=type,fill=type),size=5)+
   geom_smooth(method="lm",se=F,col="black",show.legend = F)+
@@
   scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  scale_shape_manual(values=c(15,16,17,18),name="")+
   guides(fill='none')+
@@
 # ggsave("plots/int_mean_sd.png",plot=int_mean_sd_plot,dpi=500)
+# ggsave("plots/int_mean_sd.pdf",plot=int_mean_sd_plot)
 int_mean_sd_plot
@@
 
-```{r simplex ints}
+```{r simplex ints,eval=F}
 # Run simplex projection for each species interaction
@@
 
-```{r ccm}
+```{r ccm, eval=F}
 ## CCM checking physical variables against interactions
@@
 
-```{r multivariate smaps}
+```{r multivariate smaps, eval=F}
 ints.causal.vars <- purrr::map(interaction_names, function(x) {

### 2018-07-20 manuscript.Rmd
updated manuscript and added .csl

@@
 
+```{r packages and workspace}
+require(tidyverse)
+require(extrafont)
+load("ws/wrkspce_07.19.18.RData")
+
+# plot theme
+plot_theme <-   theme_minimal()+
+  theme(text=element_text(family="Rockwell",size=12,color="black"),
+        legend.text = element_text(size=14),
+        axis.title=element_text(family="Rockwell",size=14),
+        axis.text=element_text(family="Rockwell",size=8),
+        strip.background = element_rect(colour="black"),
+        panel.border = element_rect(color="black",fill=NA))
+
+theme_set(plot_theme)
+```
+
 **Primary target journal**: Nature Ecology and Evolution
@@
 
-* Maximum 3500 words main text (excluding introductory paragraph/abstract)
+* Maximum 3500 words main text (excluding introductory pa ragraph/abstract)
 * Maximum 6 display items (figures/tables)
@@
 Time series of study species
-Example attractor?
+Example attractor? (in supplementary)
 CCM results
-Empirical interaction web, in “normal”, “high disturbance”, and “high nutrient stress” environments??
 
@@
 We focus here instead on giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented [@Foster1985a; @Graham2007c; @Teagle2017].
-The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011].
+The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces in ecosystems [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011].
 Recently, however, findings from long-term kelp forest research programs have begun to challenge many long-held beliefs about the drivers of kelp forest ecosystem dynamics [@Reed2016].
 In particular, a longer-term perspective has led to a recognition of the critical importance of environmental context—such as level of physical disturbance or the current state of El Niño conditions—for understanding kelp forest processes [@Parnell2010; @Byrnes2011; @Bell2015; @Young2015].
+In this study we utilize monitoring data from one such effort at San Nicolas Island, a small, remote member of the California Channel Islands in the northeast Pacific [@Kenner2013].
 
-In this study, we utilize monitoring data from one such effort at San Nicolas Island, a small, remote member of the California Channel Islands in the northeast Pacific [@Kenner2013].
-Our analyses focus on the dynamics of five common southern California kelp forest species, whose interactions are thought to be important in structuring kelp forest ecosystems [@Foster1985a; @Foster1990a; @Teagle2017].
-The giant kelp *Macrocystis pyrifera* is the eponymous foundation species, the primary canopy- and habitat-forming kelp along most of the central and southern coast of California [@Graham2007c].
-We explore its dynamics and interactions with two presumptive competitors and two abundant herbivores.
-The understory kelp species *Laminaria farlowii* and *Pterygophora californica* compete with Macrocystis for space, light, and nutrients [@Dayton1984; @Dayton1999; @Reed1984].
-The two herbivores---the purple sea urchin *Strongylocentrotus purpuratus* and the red sea urchin *Mesocentrotus franciscanus*---are thought in many places to control Macrocystis density and can sometimes wipe out entire giant kelp forests, leading to the alternative ecosystem state known as an urchin barren [@Ling2014].
-
-To characterize environmental context dependency in kelp forest interactions between these five species, we take three general steps (see Methods).
+To characterize environmental context dependency in kelp forest interactions between species, we take three general steps (see Methods).
 First, we use empirical dynamic modeling causality tests called convergent cross-mapping [@Sugihara2012] to construct a kelp forest species interaction network directly from time series data.
-Even though these common kelp forest species are all thought to interact, we tested that assumption with convergent cross mapping, which can separate correlation from causation [@Sugihara2012].
-
-Second, we use multivariate S-maps (sequential locally weighted global linear maps) [@Deyle2016; @Sugihara1994] to reconstruct the interactions between species for all of the causal links established in the previous step.
-S-maps reconstructs dynamic “attractors” by casting the abundances of causally-related species into state space.
-That is, for a set of causally-related species $A$, $B$, and $C$, a point in multivariate space $\{A_t, B_t, C_t\}$ can be plotted, using each species’ abundance at time t.
-The attractor is then created by tracing this multivariate trajectory forward in time for all t (FIG?).
-S-maps then computes sequential Jacobian matrices for each point along the attractor, where the elements of each matrix are the partial derivatives between species.
-These estimated partial derivatives are our measure of species interactions.
-
-The last step is to investigate if and how species interactions varies with environmental context.
-Because multivariate S-maps are computed for every point along reconstructed attractors, we can extract distributions of interactions that can be related back to environmental context.
-We explore patterns of species interactions in relation to sea surface temperature, physical disturbance (measured by wave height), and three indices of low-frequency climate modes: the Multivariate ENSO Index, the Pacific Decadal Oscillation, and the North Pacific Gyre Oscillation [@Wolter2011; @Mantua2002; @DiLorenzo2008].
+In so doing, we test for all unidirectional causal signals between five common kelp forest species, as well as between five exogenous environmental variables and those species.
+Second, for each identified causal link between species, we reconstruct the actual species interactions over time, using another EDM tool called multivariate S-maps (sequential locally weighted global linear maps) [@Deyle2016; @Sugihara1994].
+Finally, we show how variability in key species interactions can be related back to the environmental context under which they took place.
 
@@
 
-```{r fig.width=4, fig.height=4,echo=FALSE,fig.cap="Example reconstructed dynamic attractor"}
-library(png)
-library(grid)
-include_graphics("plots/multi_attractor.pdf")
-```
+Our analyses focus on the dynamics of five common southern California kelp forest species, whose interactions are thought to be important in structuring kelp forest ecosystems [@Foster1985a; @Foster1990a; @Teagle2017].
+The giant kelp *Macrocystis pyrifera* is the eponymous foundation species[@Dayton1972], the primary canopy- and habitat-forming kelp along most of the central and southern coast of California [@Graham2007c].
+The monitoring data include young *Macrocystis* recruits (sporophytes identified as *Macrocystis* but less than 1 meter tall [@Kenner2013]).
+We explore *Macrocystis* dynamics and its interactions with two presumptive competitors and two abundant herbivores.
+The understory kelp species *Laminaria farlowii* and *Pterygophora californica* compete with Macrocystis for space, light, and nutrients [@Dayton1984; @Dayton1999; @Reed1984].
+The two herbivores---the purple sea urchin *Strongylocentrotus purpuratus* and the red sea urchin *Mesocentrotus franciscanus*---are thought in many places to control *Macrocystis* density and can sometimes wipe out entire giant kelp forests, leading to the alternative ecosystem state known as an urchin barren [@Ling2014; @Harrold1985a].
 
-```{r fig.width=4, fig.height=4,echo=FALSE,fig.cap="Reconstructed interaction web using results of convergent cross mapping"}
+In southern California, *Macrocystis* population dynamics can be driven by nutrient availability and physical disturbance [@Bell2015; @Reed2011].
+The availability of nitrate is inversely related to seawater temperature[@Zimmerman1984] and, over longer time scales, is associated with oscillations in patterns of upwelling and oceanic currents.
+Accordingly, we include five environmental variables in our analyses to test their relationship to kelp forest species dynamics and interactions: sea surface temperature (SST), physical disturbance (measured by maximum seasonal wave height, SWH), and three indices of low-frequency climate modes: the Multivariate El Niño Index (MEI) [@Wolter2011], the Pacific Decadal Oscillation (PDO) [@Mantua2002], and the North Pacific Gyre Oscillation (NPGO) [@DiLorenzo2008].
+
+Applying convergent cross mapping[@Sugihara2012] (CCM) to the set of six biological and five physical variables, we find a relatively dense interaciton network (Figure \ref{fig:ccm}).
+Out of 90 possible unidirectional links between species and between species and the environmental variables, 40 are significant.
+Adult *Macrocystis* density is driven by all five environmental variables, with SWH, SST, and the NPGO showing the strongest causal signals.
+This finding confirms recent work by others using different methods [@Bell2015] that showed that these same three variables were the primary controls of giant kelp biomass dynamics across the California coast.
+More generally, although the included physical variables show significant links to many of the biological variables, the NPGO and SWH show the strongest links to almost all of the biological variables.
+Our analysis suggests that more attention should be focused on the effects of the NPGO.
+
+```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping\\label{fig:ccm}"}
 include_graphics("plots/ccm_network.pdf")
@@
 
-```{r echo=FALSE,fig.cap="Distribution of estimated interaction by type"}
-img <- readPNG("plots/int_dist_by_type_density.png")
-grid.raster(img)
+Interestingly, in turn, adult *Macrocystis* shows strong causal links to every other biological variable.
+This is despite the fact that the study site at San Nicolas Island does not have a stable giant kelp forest (see raw time series, Figure \ref{fig:normts}); rather, the site has transitioned from an urchin barren[@Harrold1985a] to a *Pterygophora* and *Laminaria*-dominated state, to a *Macrocystis* forest at various times throughout the 30-year time series.
+The implication is that, despite not maintaining dominance in the typical ecological sense of word (large abundance and biomass), *Macrocystis* is "dynamically dominant", and remains the key foundation species in this ecosystem.
+Its dynamics are fundamentally important in driving the dynamics of all the other kelp forest species[@Graham2007c].
+
+CCM analysis confirms that the system studied represents a complex array of significant interactions between algal species and their herbivores.
+However, CCM alone does not elucidate the direction and magnitude of species interactions.
+To obtain estimates of the interactions themselves, we use multivariate S-maps[@Deyle2016; @Sugihara1994] (see Methods).
+S-maps reconstructs dynamic “attractors” by casting the abundances of causally-related species into state space.
+For a set of causally-related species, a point in multivariate space can be plotted using each species’ abundance as an axis.
+The attractor is then created by tracing this multispecies trajectory forward in time (see example attractor in Figure \ref{fig:attractor}).
+For each point along the attractor, S-maps computes a Jacobian matrix, the elements of which are the estimated partial derivatives between species.
+These interaction matrix elements are our measure of species interactions.
+Because Jacobians are computed sequentially for every point along reconstructed attractors, we obtain estimates of interaction strength that vary with ecosystem state.
+
+There are two important aspects of S-maps that deserve mention.
+First, the S-map estimation procedure, like all EDM methods, is specifically designed for nonlinear systems, and is therefore an appropriate tool for investigating ecosystems exhibiting nonlinear dynamics such as alternative stable states or hysteresis[@Deyle2016].
+Secondly, because S-maps utilizes reconstructed multispecies attractors, each estimated interaction is fundamentally based on observations of similar past ecosystem states---in our case represented as multivariate collections of causally-linked species' densites---rather than a phenomenological extrapolation of the most recent dynamics.
+In simple terms, instead of asking, "What is our prediction for the strength of herbivory based on last year's observed dynamics," S-maps is concerned with, "What is our best estimate for the strength of herbivory, based on our knowledge of times in the past when the ecosystem was most similar to today?"
+
+```{r fig.cap="Distribution of estimated interaction by type\\label{fig:int_dist}"}
+interaction_distribution_by_type
 ```
 
-```{r echo=FALSE, fig.cap="Mean and standard deviation of each estimated unidirectional species interaction"}
-img <- readPNG("plots/int_mean_sd.png")
-grid.raster(img)
+In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}).
+After grouping species interactions by type, only herbivory (the effect of urchins on algae species) is predominantly negative.
+Conversely, the effect of the algae species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions.
+Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic.
+These results suggest that facilitation is common in kelp forests when context is considered.
+Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species may sometimes outweigh their direct competition.
+Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c].
+These types of indirect facilitation in kelp forests have received little attention[@Arkema2009a; Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006].
+Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an important structuring factor in kelp forest ecosystems.
+
+```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction"}
+int_mean_sd_plot
 ```
 
-```{r echo=FALSE,fig.cap="Macrocystis effect on Pterygophora"}
-img <- readPNG("plots/mac_effect_on_pter_phys.png")
-grid.raster(img)
+```{r fig.cap="Macrocystis effect on Pterygophora"}
+mac_effect_on_pter_plot
 ```
@@
 
+## Methods
+
+```{r fig.width=4, fig.height=4,fig.cap="Example reconstructed dynamic attractor\\label{fig:attractor}"}
+include_graphics("plots/multi_attractor.pdf")
+```
+
+```{r fig.cap="Raw data for species and physical drivers, normalized to zero mean and unit variance\\label{fig:normts}"}
+include_graphics("plots/normalized_time_series.pdf")
+```
+
 ## References

### 2018-07-23 EDM_SNI_analysis.Rmd
complete rough draft manuscript

@@
 #plot to file if uncommented
-# pdf("plots/ccm_network.pdf",height=7,width=7)
+# png("plots/ccm_network.png",height=500,width=600)
 plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
@@
 
+```{r pter effect ymac,fig.height=6,fig.width=6}
+pter_effect_ymac <-spp.only.coeffs.with.phys %>%
+  ungroup()%>%
+  filter(mod=="ymac",effect=="pter")%>%
+  mutate(phys_high_low=case_when(
+    ext_value > 1 ~ "high",
+    ext_value < -1 ~ "low")) %>%
+  filter(!is.na(phys_high_low))%>%
+  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
+  group_by(type,external_forcing,phys_high_low)%>%
+  mutate(mean_int=mean(value,na.rm=T))%>%
+  distinct()%>%
+  ungroup()
+
+# physical forcings labeller
+phys.labeller <- c(
+  mei= "MEI",
+  npgo= "NPGO",
+  pdo= "PDO",
+  sst="SST",
+  waves="SWH"
+)
+
+pter_effect_ymac_plot <-pter_effect_ymac %>%
+ggplot(aes(value,..density..,fill=phys_high_low))+
+  geom_density(col=NA,alpha=0.6)+
+  geom_vline(xintercept=0,linetype=2)+
+  geom_vline(aes(xintercept=mean_int,color=phys_high_low),size=1.5)+
+  scale_fill_manual(values=c("darkorange","dodgerblue4"),name="Physical\nForcing Value",labels=c("high","low"))+
+  scale_color_manual(values=c("darkorange","dodgerblue4"))+
+  guides(color="none")+
+  labs(x="Effect of Pterygophora on Macrocystis Recruits",y="Probability Density")+
+  facet_grid(rows=vars(external_forcing),labeller=labeller(external_forcing=phys.labeller))+
+  theme(strip.text=element_text(size=10))
+
+# ggsave("plots/pter_effect_ymac_phys.png",plot=pter_effect_ymac_plot,h=6,w=6,dpi=500)
+# ggsave("plots/pter_effect_ymac_phys.pdf",plot=pter_effect_ymac_plot,h=6,w=6)
+pter_effect_ymac_plot
+```
+
 ### Distribution of interaction strengths by type
@@
   xlim(-1.5,1.5)+
-  labs(x="Interaction strength",y="Kernel density")+
+  labs(x="Interaction strength",y="Probability density")+
   geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
@@
 
+# Mean vs. range (95%)
+
+int_mean_range_plot <- int_mean_sd %>%
+  filter(type != "intraspecies", effect != "ymac", mod != "ymac")%>%
+  ggplot(aes(abs_mean,y=upper-lower))+
+  geom_point(aes(color=type,shape=type,fill=type),size=5)+
+  geom_smooth(method="lm",se=F,col="black",linetype=2,show.legend = F)+
+  # geom_vline(xintercept=0,linetype=2)+
+  # geom_encircle(aes(fill=type),col=NA,alpha=0.3,expand=0.1)+
+  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  scale_shape_manual(values=c(15,16,17,18),name="")+
+  guides(fill='none')+
+  labs(x="Mean Interaction Strength",y="Range Interaction Strength")+
+  theme(legend.position = c(0.8,0.8),
+        legend.background = element_rect(color="black"))
+# ggsave("plots/int_mean_range.png",plot=int_mean_range_plot,h=5,w=6)
+# ggsave("plots/int_mean_range.pdf",plot=int_mean_sd_plot)
+int_mean_range_plot
+
 # CV

### 2018-07-23 manuscript.Rmd
complete rough draft manuscript

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: pdf_document
-fontsize: 12pt
-bibliography: citation_bibtex.bib
-csl: nature.csl
+output:
+  word_document: default
+  pdf_document: default
 editor_options:
   chunk_output_type: inline
+fontsize: 12pt
 header-includes:
@@
 - \renewcommand\linenumberfont{\normalfont\bfseries\small}
+csl: nature.csl
+bibliography: citation_bibtex.bib
 ---
@@
 ```{r packages and workspace}
+# # header-includes:
+#   - \usepackage[left]{lineno}
+#   - \linenumbers
+#   - \renewcommand\linenumberfont{\normalfont\bfseries\small}
 require(tidyverse)
 require(extrafont)
-load("ws/wrkspce_07.19.18.RData")
+load("ws/wrkspce_07.20.18.RData")
 
@@
 
-* Maximum 3500 words main text (excluding introductory pa ragraph/abstract)
+* Maximum 3500 words main text (excluding introductory paragraph/abstract)
 * Maximum 6 display items (figures/tables)
 * The introductory paragraph is typically 150 words and is unreferenced; it contains a brief account of the background and rationale of the work, followed by a statement of the main conclusions introduced by the phrase "Here we show" or some equivalent.An introduction (without heading) of up to 500 words of referenced text expands on the background of the work (some overlap with the summary is acceptable), and is followed by a concise, focused account of the findings (headed 'Results'), and one or two short paragraphs of discussion (headed 'Discussion').
-
-**Work confirmed/supported/debates**:
-
-* Tegner et al 1997, Dayton et al 1999
-* Bottom-up vs. top-down forcing? Disturbance or herbivory?
-* Competition between urchin species?
-* Physical drivers increase predictability of algae models!
-
-**Suggested future research**
-
-* Predominance of mutualism- why the positive interactions of Pterygophora and Macrocystis?
-* Urchin grazing pressure as most important??
-* Multiple stressors (e.g. herbivory, sst, wave height)
-
-**FIGURES**
-
-Time series of study species
-Example attractor? (in supplementary)
-CCM results
 
@@
 Interactions among species are often measured with short-term controlled experiments, but these experiments are subject to the particular environmental conditions under which they are performed.
-As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interaction strengths across a wide range of environmental conditions in a coastal marine ecosystem.
-By including large-scale climate indices, sea surface temperature, and a measure of physical disturbance in analyses, we show that environmental context influences the strength and direction of species interactions.
-In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied interactions.
+As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interactions across a wide range of environmental conditions.
+We show that environmental context influences the strength and direction of species interactions.
+In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied dynamics, including the importance of specifically studying variation rather than mean interaction outcomes.
 The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of ecosystem dynamics.
@@
 Such an approach could help to characterize environmental contingencies in species interactions and explicitly examine interaction variability.
-Here, we use empirical dynamic modelling (EDM, [@Deyle2016]) to estimate a varying species interaction network and establish environmental context dependency in interaction strength and direction.
+Here, we use empirical dynamic modelling (EDM[@Deyle2016]) to estimate a varying species interaction network and establish environmental context dependency in interaction strength and direction.
 Empirical dynamic modelling uses information from single or multiple time series to empirically model relationships between variables through the reconstruction of dynamic attractors (https://youtu.be/8DikuwwPWsY).
@@
 
-## Results and discussion
+## Results
 
@@
 
-```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping\\label{fig:ccm}"}
+```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping.
+Each arrow represents a significant causal signal, and link width and opacity scale with the strength of causal forcing (see full results in Supplementary material).
+Species abbreviations: M.
+pyr: \\textit{Macrocystis pyrifera}; L.far: \\textit{Laminaria farlowii}; P.cal: \\textit{Pterygophora californica}; M.fra: \\textit{Mesocentrotus franciscanus}; S.pur: \\textit{Strongylocentrotus purpuratus}.
+Physical drivers: NPGO: North Pacific Gyre Oscillation; MEI: Multivariate El Nino Index; PDO: Pacific Decadal Oscillation; SST: Sea surface temperature; SWH: Significant wave height.\\label{fig:ccm}"}
 include_graphics("plots/ccm_network.pdf")
@@
 
-```{r fig.cap="Distribution of estimated interaction by type\\label{fig:int_dist}"}
+```{r fig.cap="Smoothed kernel density histograms of estimated interactions by type. Solid lines denote means across all estimated interactions of that type.\\label{fig:int_dist}"}
 interaction_distribution_by_type
@@
 Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic.
-These results suggest that facilitation is common in kelp forests when context is considered.
-Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species may sometimes outweigh their direct competition.
+These results suggest that facilitation---direct or indirect---can arise in kelp forests in multiple contexts[@Bennett2015].
+Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species through shared herbivory avoidance[@Harris1984] may sometimes outweigh their direct competition.
 Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c].
-These types of indirect facilitation in kelp forests have received little attention[@Arkema2009a; Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006].
-Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an important structuring factor in kelp forest ecosystems.
+These types of indirect facilitation in kelp forests have received relatively little attention[@Arkema2009a; @Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006].
+Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an additional important structuring force.
 
-```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction"}
-int_mean_sd_plot
+```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction (excluding intraspecies effects)\\label{fig:int_mean_range}"}
+int_mean_range_plot
 ```
 
-```{r fig.cap="Macrocystis effect on Pterygophora"}
+There is an inverse relationship between mean and range in interaction strengths (Figure \ref{fig:int_mean_range}).
+Importantly, these reported ranges in interaction strength are distinct from estimation or experimental errors.
+They represent observed ranges across all individually-estimated S-map interactions from each reconstructed attractor.
+While inverse mean-range relationship was true overall, it varied by interaction type.
+For the three significant interactions between algae species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series.
+The effects of algal on urchin species were the most variable in general, and the herbivory effects the least.
+Additionally, the range was greater than the mean for all of these interactions, further supporting the idea that context dependency may be the rule rather than the exception in this ecosystem.
+The observed mean-range relationship echoes work done across a range of experimental systems[@Berlow1999b], and implies that interactions that seem weak on average may simply vary in sign and magnitude across contexts but still be important in each of those contexts nonetheless.
+
+```{r fig.cap="Distributions of \\textit{Macrocystis} effects on \\textit{Pterygophora}, under high (greater than 1) and low (less than -1) values of five normalized environmental indices (abbreviations as in Figure \\ref{fig:ccm}.
+Solid lines: mean interactions under each regime.\\label{fig:mac_pter}"}
 mac_effect_on_pter_plot
 ```
+
+As a case in point, consider the interaction of adult *Macrocystis* and the understory kelp *Pterygophora californica*,  the interaction with the "weakest" mean strength of all algal interactions in Figure \ref{fig:int_mean_range}.
+*Macrocystis* is often the dominant competitor in kelp forests for nutrients and light[@Dayton1984; @Reed1984], but in this site on average has a neutral to positive effect on *Pterygophora* (Figure \ref{fig:mac_pter}), a result seemingly incongruous with the established competitive hierarchy.
+However, in southern California, *Macrocystis* is also known to be a better competitor under cold-water, nutrient-rich conditions[@Dayton1999].
+Although negative interactions of *Macrocystis* with *Pterygophora* are rare in our study, they align with those expectations: *Macrocystis* has a greater chance of flipping its interaction from positive to negative---and asserting interspecies dominance---under low values of the MEI and PDO, and high values of the NPGO (Figure \ref{fig:mac_pter}).
+These are all climate regimes associated with increased availability of nutrients in southern California[@Mantua2002; @Wolter2011; @DiLorenzo2008].
+Hence, the rare, but strong negative effects of *Macrocystis* on *Pterygophora* conform to expected patterns across decadal-scale climate shifts[@Dayton1999; @Tegner1997], despite the two species' weak mean interaction at this site.
+In this way, the prevailing environmental context at any point in time is the landscape upon which species interactions occur, potentially tipping the scales in favor of one species over another.
+Moreover, these rare competitive events may be precisely what precipitates nonlinear ecosystem shifts between alternative states.
+
+## Discussion
+
+Ecosystem dynamics are composed of nonlinear species relationships, played out within shifting environmental contexts.
+A significant challenge in the study of ecosystem dynamics has been the difficulty in appropriately extrapolating experimental results to real ecosystems, where multiple species-species and species-environment interactions are operating simulateously.
+We have shown in this study that empirical dynamic modeling can help to tackle this challenge, using time series data to accurately reconstruct nonlinear ecosystem trajectories.
+Beginning with a published monitoring data set from a kelp forest ecosystem, EDM methods helped to elucidate causation, build interaction networks, and investigate the influence of large-scale environmental drivers on interaction strength.
+In this particular ecosystem, our analyses confirmed decades of experimental work regarding the foundation species *Macrocystis pyrifera*, but also were able to contextualize those classic interactions as they apply at the study site.
+A classic algal competitive dominance hierarchy[@Dayton1999] is seemingly weak at this site under average conditions, but not absent---under predictable nutrient, temperature, and disturbance regimes, *Macrocystis* can be dominant at this site.
+
+More generally, we report evidence here of an inverse mean-range relationship in interaction strengths and a strong presence of apparent facilitation under multiple environmental contexts, results that agree with previous theoretical and experimental findings across multiple ecosystems[@McCann1998; @Berlow1999b; @Bruno2003; @Bulleri2009].
+An important implication is that if a goal is understanding the dynamics of entire ecosystems, studying the mean outcome of single species interactions may not be adequate.
+Indirect associations between multiple species and shifting environmental contexts may give rise to rare, critical moments when fleeting strong interactions determine ecosystem shifts.
+This idea needs further investigation in multiple systems, but if widely applicable, it means that context dependency---and its role in mediating varying species interaction strengths---deserves more attention than the identification of context-averaged mean interactions[@Chamberlain2014].
+In a growing number of ecosystems, EDM is helping in this endeavour[@Deyle2016; @Ye2015a; @McGowan2017].
+
+EDM does not take the place of experimentation, but we argue that it can help to both contextualize and guide it.
+Our analysis was a proof of concept: we started with simple time series from a monitoring dataset in a well-studied but complex ecosystem, and showed how previous experimental results play out over a longer time period.
+We hope that this approach can be applied to many other ecosystems, where time series data exist but where important interactions may not be as well-established.
+Where important interactions are known, EDM can help to explore whether environmental context matters in interaction variance.
+Where those interactions are not as well known, EDM may be a helpful first step in identification of ecosystem links whose mechanisms can then be further established through other methods.
+
+## Extra notes (scratch space)
+
+The disturbance effect is interesting as well[@Tegner1997].
+Large disturbances can rip out entire forests (*Macrocystis* and *Pterygophora* alike) and can be the catalyst for phase shifts in ecosystem state.
+That is evident here in the negative tail of the effect of *Macrocystis* on *Pterygophora*.
+If conditions are beneficial for rapid growth, *Macrocystis* can establish more quickly than *Pterygophora* after disturbance, thereby achieving dominance and having a negative effect on *Pterygoph
+ora*.
 
@@
 * Frequency of positive vs. negative interactions? Across environmental gradients?
+
+The realized competitive niche of *Macrocystis*[@Dayton1999] may be altered at the study site, which is comprised of a low-rugosity reef highly exposed to wave disturbance, habitat characteristics that can inhibit or even reverse *Macrocystis* dominance[@Graham1997].
+The simple observation that the San Nicolas Island site has not maintained a stable kelp forest for significant periods of time further supports the lack of consistent *Macrocystis* dominance.
+
+**Work confirmed/supported/debates**:
+
+* Tegner et al 1997, Dayton et al 1999
+* Bottom-up vs. top-down forcing? Disturbance or herbivory?
+* Competition between urchin species?
+* Physical drivers increase predictability of algae models!
+
+**Suggested future research**
+
+* Predominance of mutualism- why the positive interactions of Pterygophora and Macrocystis?
+* Urchin grazing pressure as most important??
+* Multiple stressors (e.g. herbivory, sst, wave height)
 

### 2018-08-21 EDM_SNI_analysis.Rmd
molly's icons and manuscript abstract

@@
 require(tidyverse)
+require(here)
 require(rEDM)
@@
 require(extrafont)
+require(png)
+require(ggridges)
 require(GGally)
 require(ggalt)
+require(ggsci)
 require(sna)
@@
 # Density plot
+require(png)
+a1 <- readPNG(here::here("plots","icons","macrocystis-transparent.png")) %>% rasterGrob()
 interaction_distribution_by_type <- spp.only.mod.coeffs %>%
@@
   ungroup()%>%
-  ggplot(aes(value,..density..))+
-  geom_density(aes(fill=type),alpha=0.8,col=NA)+
+  ggplot(aes(value,type,fill=type))+
+  geom_density_ridges(alpha=0.8,scale=1.4)+
+  # stat_density_ridges(geom="density_ridges_gradient",calc_ecdf='true',quantiles=4,alpha=0.8,scale=1)+
+  # scale_fill_viridis_d(name="quantiles",alpha=0.8)+
   geom_vline(xintercept=0,linetype=2)+
-  geom_vline(aes(xintercept=mean_int))+
+  # geom_vline(aes(xintercept=mean_int,color=type))+
   scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
-   guides(fill=F)+
+  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
+  guides(fill=F,color=F)+
   xlim(-1.5,1.5)+
   labs(x="Interaction strength",y="Probability density")+
-  geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
-  facet_grid(rows=vars(type))+
+  annotation_custom(a1,ymin=2,ymax=3,xmin=-1,xmax=-0.5)+
+  # geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
+  # facet_grid(rows=vars(type))+
   theme(strip.text = element_blank())

### 2018-08-21 manuscript.Rmd
molly's icons and manuscript abstract

@@
 require(extrafont)
+require(ggridges)
 load("ws/wrkspce_07.20.18.RData")
@@
 
-Ecological interactions are not uniform across time, and instead vary with environmental conditions.
-Interactions among species are often measured with short-term controlled experiments, but these experiments are subject to the particular environmental conditions under which they are performed.
-As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series to estimate species interactions across a wide range of environmental conditions.
-We show that environmental context influences the strength and direction of species interactions.
-In so doing, we are able to confirm and extend results from previous studies, as well as identify potentially important but understudied dynamics, including the importance of specifically studying variation rather than mean interaction outcomes.
-The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of ecosystem dynamics.
+Ecological interactions are not uniform across time.
+They commonly vary with environmental conditions.
+Yet, interactions among species are often measured with short-term controlled experiments whose outcome can depend greatly on the particular environmental conditions under which they are performed.
+Running experiments repeatedly to address this context challenge is an inefficient solution at best.
+As an alternative, we utilize empirical dynamic modeling applied to a 30-year time series from coastal kelp forests to estimate species interactions across a wide range of environmental conditions.
+We show that environmental context greatly alters the strength and direction of species interactions.
+In so doing, we confirm and extend results from previous studies on the power of empirical dynamic modeling.
+We identify potentially important but understudied kelp forest dynamics, including the importance of specifically studying variation in interaction strength rather than mean interaction outcomes.
+The significant context dependency in species interactions found in this study argues for a greater utilization of long-term data and empirical dynamic modeling in studies of the dynamics of other ecosystems.
 

### 2018-08-23 EDM_SNI_analysis.Rmd
manuscript and figure edits

@@
 # label with the number of incoming(outgoing) connections
-vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
-vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
-V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
-V(g2)$label[7:11] <- paste(V(g2)$name[7:11])
+# vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
+# vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
+# V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
+# V(g2)$label[7:11] <- paste(V(g2)$name[7:11])
 
@@
 # png("plots/ccm_network.png",height=500,width=600)
-plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
+# pdf("plots/ccm_network.pdf")
+plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.8, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
 # dev.off()
@@
   group_by(mod,effect,type) %>%
-  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
+  summarise(median_int=median(value,na.rm=T),mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
             cv_int=sd_int/abs_mean*100,upper=quantile(value,0.975),lower=quantile(value,0.025),
@@
   geom_col()+
-  geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
+  # geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
   labs(x='Interaction Type',y='Interaction Strength')
+
+# Median interaction strength by interaction type
+spp.only.mod.coeffs %>%
+  filter(!is.na(type))%>%
+  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
+  filter(mod != "ymac",effect != 'ymac')%>%
+  group_by(type) %>%
+  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),median_int=median(value,na.rm=T),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
+  ungroup()%>%
+  ggplot(aes(type,median_int))+
+  geom_col(fill="gray60")+
+  geom_point(data=int_mean_sd,aes(type,median_int,color=type),position="dodge")+
+  scale_color_npg()+
+  guides(color=F)+
+  # geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
+  labs(x='Interaction Type',y='Median Interaction Strength')
 

### 2018-08-23 manuscript.Rmd
manuscript and figure edits

@@
 output:
+  pdf_document: default
   word_document: default
-  pdf_document: default
 editor_options:
@@
 
-Recently-developed EDM methods exist for uncovering dynamic species interactions from time series data [@Deyle2016], but these methods have insofar been applied only to simulated and planktonic communities, and their utility to the study of other ecological systems remains untested.
-We focus here instead on giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented [@Foster1985a; @Graham2007c; @Teagle2017].
+Recently-developed EDM methods exist for uncovering dynamic species interactions from time series data [@Deyle2016], but these methods have to date been applied only to simulated and planktonic communities, and their utility to the study of other ecological systems remains untested.
+Here we extend the exploration of dynamic species interactions using EDM to giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented through experimental and comparative studies [@Foster1985a; @Graham2007c; @Teagle2017].
 The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces in ecosystems [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011].
@@
 Adult *Macrocystis* density is driven by all five environmental variables, with SWH, SST, and the NPGO showing the strongest causal signals.
-This finding confirms recent work by others using different methods [@Bell2015] that showed that these same three variables were the primary controls of giant kelp biomass dynamics across the California coast.
+This finding aligns with recent work by others using different methods [@Bell2015] that showed that these same three variables were the primary controls of giant kelp biomass dynamics across the California coast.
 More generally, although the included physical variables show significant links to many of the biological variables, the NPGO and SWH show the strongest links to almost all of the biological variables.
@@
 Physical drivers: NPGO: North Pacific Gyre Oscillation; MEI: Multivariate El Nino Index; PDO: Pacific Decadal Oscillation; SST: Sea surface temperature; SWH: Significant wave height.\\label{fig:ccm}"}
-include_graphics("plots/ccm_network.pdf")
+include_graphics("plots/ccm_network_w_icons.pdf")
 ```
@@
 This is despite the fact that the study site at San Nicolas Island does not have a stable giant kelp forest (see raw time series, Figure \ref{fig:normts}); rather, the site has transitioned from an urchin barren[@Harrold1985a] to a *Pterygophora* and *Laminaria*-dominated state, to a *Macrocystis* forest at various times throughout the 30-year time series.
-The implication is that, despite not maintaining dominance in the typical ecological sense of word (large abundance and biomass), *Macrocystis* is "dynamically dominant", and remains the key foundation species in this ecosystem.
-Its dynamics are fundamentally important in driving the dynamics of all the other kelp forest species[@Graham2007c].
+The implication is that, despite not maintaining dominance in the typical ecological sense of word (large abundance and biomass), *Macrocystis* remains a key foundation species in this ecosystem, because its dynamics are fundamentally important in driving the dynamics of all the other kelp forest species[@Graham2007c].
 
@@
 To obtain estimates of the interactions themselves, we use multivariate S-maps[@Deyle2016; @Sugihara1994] (see Methods).
-S-maps reconstructs dynamic “attractors” by casting the abundances of causally-related species into state space.
+S-maps reconstruct dynamic “attractors” by casting the abundances of causally-related species into state space.
 For a set of causally-related species, a point in multivariate space can be plotted using each species’ abundance as an axis.
@@
 
-There are two important aspects of S-maps that deserve mention.
+Dynamic ecosystems are analogous to a landscape of variable topography, where our position on the landscape represents the current ecosystem state.
+In this analogy, each cardinal direction represents the density of a different species (i.e., axis in state-space), and our movement across the landscape through time represents movement along the multivariate attractor as species densities change.
+S-maps are simply our reconstructed topographic maps of these ecosystem landscapes.
+At a given point in time as we move across the landscape, we may be on top of a steep pinnacle, on a flat plain, or in a shallow valley.
+Regardless, it is the slope of the landscape in each direction (the partial derivative) that defines the local interaction strength between species: a steep upward slope represents a strong positive interaction between species, while a flat surface is a neutral interaction and a slight downward slope is a weak negative interaction.
+
+There are two important characteristics of S-maps that deserve mention.
 First, the S-map estimation procedure, like all EDM methods, is specifically designed for nonlinear systems, and is therefore an appropriate tool for investigating ecosystems exhibiting nonlinear dynamics such as alternative stable states or hysteresis[@Deyle2016].
-Secondly, because S-maps utilizes reconstructed multispecies attractors, each estimated interaction is fundamentally based on observations of similar past ecosystem states---in our case represented as multivariate collections of causally-linked species' densites---rather than a phenomenological extrapolation of the most recent dynamics.
-In simple terms, instead of asking, "What is our prediction for the strength of herbivory based on last year's observed dynamics," S-maps is concerned with, "What is our best estimate for the strength of herbivory, based on our knowledge of times in the past when the ecosystem was most similar to today?"
+Secondly, because S-maps utilizes reconstructed multispecies attractors, each estimated interaction is fundamentally based on observations of similar past ecosystem states---where each state is represented as a multivariate vector of causally-linked species' densites---rather than a phenomenological extrapolation of the most recent dynamics.
+For example, instead of asking, "What is our prediction for the strength of herbivory based on last year's observed dynamics," S-maps is concerned with, "What is our best estimate for the strength of herbivory, based on our knowledge of times in the past when the ecosystem was most similar to today?"
 
 ```{r fig.cap="Smoothed kernel density histograms of estimated interactions by type. Solid lines denote means across all estimated interactions of that type.\\label{fig:int_dist}"}
-interaction_distribution_by_type
+include_graphics("plots/int_dist_by_type_density.pdf")
+# interaction_distribution_by_type
 ```
@@
 In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}).
-After grouping species interactions by type, only herbivory (the effect of urchins on algae species) is predominantly negative.
-Conversely, the effect of the algae species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions.
-Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic.
+After grouping species interactions by type, only herbivory (the effect of urchins on algal species) is predominantly negative.
+Conversely, the effect of the algal species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions.
+Likewise, contrary to our expectations, interactions between the algal species and between the urchin species are not always antagonistic.
 These results suggest that facilitation---direct or indirect---can arise in kelp forests in multiple contexts[@Bennett2015].
-Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species through shared herbivory avoidance[@Harris1984] may sometimes outweigh their direct competition.
+For example, since there is evidence here of strong herbivory, then there may be indirect facilitation between algal species because of a shared herbivore[@Harris1984]: a greater algal density in general could ameliorate the negative effect of herbivory on any one species.
+This sort of indirect, associative facilitation may sometimes outweigh the strength of direct algal competition.
 Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c].
-These types of indirect facilitation in kelp forests have received relatively little attention[@Arkema2009a; @Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006].
+These types of indirect facilitation in kelp forests have received comparatively little attention[@Arkema2009a; @Benes2015] relative to the strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other ecosystems[@Baraza2006].
 Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an additional important structuring force.
 
-```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction (excluding intraspecies effects)\\label{fig:int_mean_range}"}
+```{r fig.cap="Mean absolute interaction strength and standard deviation for each estimated unidirectional species interaction, excluding intraspecies effects\\label{fig:int_mean_range}"}
 int_mean_range_plot
@@
 
-There is an inverse relationship between mean and range in interaction strengths (Figure \ref{fig:int_mean_range}).
+The empirical distributions in Figure \ref{fig:int_dist} are grouped by interaction type.
+We can also investigate the distribution of each individual species interaction (e.g., the effect of purple urchins on giant kelp).
+Between the five unique species, there were 14 significant unidirectional causal links (Figure \ref{fig:ccm}).
+Because S-maps estimate interactions for each historical ecosystem state, those 14 interactions each have their own distribution of interactions strengths (Figure \ref{fig:int_mean_range}).
+In general, interactions with larger mean strength (in absolute terms) also had a smaller range.
 Importantly, these reported ranges in interaction strength are distinct from estimation or experimental errors.
@@
 While inverse mean-range relationship was true overall, it varied by interaction type.
-For the three significant interactions between algae species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series.
+For the three significant interactions between algal species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series.
 The effects of algal on urchin species were the most variable in general, and the herbivory effects the least.
@@
 Solid lines: mean interactions under each regime.\\label{fig:mac_pter}"}
-mac_effect_on_pter_plot
+# mac_effect_on_pter_plot
+include_graphics("plots/mac_effect_on_pter_phys_w_icons.pdf")
 ```
